### 2018-06-12 complex.tex

We now add some complex theories and morphisms to \mmt.
\begin{grammar}
T      & \{\} \alt \bigcup\{\rep{T}\} \alt M(T)   & \text{complex theories}\\
M      & \{\} \alt \bigcup\{\rep{M}\} \alt M^T \alt M^T(M)   & \text{complex morphisms}
\end{grammar}

Their meanings are explained in the sequel.

\paragraph{Empty Theory}
The empty theory $\{\}\in\Thy$ and the empty morphism $\{\}\in\Mo{\{\}}{T}$ for $T\in\Thy$ are always well-formed.
Their semantics is given by $\flt{\{\}}=\es$ (in both cases).
$\{\}$ is an initial object in the category of theories.

The empty morphism is implicit.

\paragraph{Union of Theories}
For any finite set $\Theta\sq\Thy$ of theories, the union theory $\bigcup \Theta\in\Thy$ is always well-formed.
Its semantics is given by $\flt{(\bigcup\Theta)}=\bigcup_{T\in\Theta}\flt{T}$.
As special cases, we can recover $T\equiv\bigcup\{T\}$ and $\{\}\equiv\bigcup\es$.

We write $S\cup T$ for the binary union $\bigcup\{S,T\}$.
Binary union is a semi-lattice (with respect to $\harr$ and $\equiv$) with least element $\{\}$.

Clearly, we have $T\harr \bigcup\Theta$ if $T\in\Theta$.
Moreover, we define the include closure of a set of theories by $\cls{\Theta}=\{C|C\harr T, T\in\Theta\}$ (which is a topological closure operator).
Then we have $\bigcup Z\harr \bigcup \Theta$ iff $\cls{Z}\sq\cls{\Theta}$ and thus also $\bigcup Z\equiv \bigcup \Theta$ iff $\cls{Z}=\cls{\Theta}$.

Now consider morphisms $M_i\in\Mo{S_i}{T_i}$ for $i=1,\ldots,n$.
Let $\mu=\{M_1,\ldots,M_n\}$, $Z=\{S_1,\ldots,S_n\}$ and $\Theta=\{T_1,\ldots,T_n\}$.
The union morphism $\bigcup \mu: \bigcup Z \to \bigcup \Theta$ is well-formed if $M_i|_C\equiv M_j|_C$ whenever $C\harr S_i$ and $C\harr S_j$.
Its semantics is given by $\flt{(\bigcup \mu)}=\bigcup_{M\in\mu} \flt{M}$.

$\bigcup \mu$ is implicit if all $M_i$ are.

\paragraph{Pushout along Inclusion}
For morphisms $M\in\Mo{S}{T}$ and $S\harr X$, the pushout $T\harr M(X)$ with the morphism $M^X\in\Mo{X}{M(X)}$ are well-formed if for all $C\harr X$ and $C\harr T$ also $C\harr S$.

In that case, for $u\in\Mo{T}{U}$ and $x\in\Mo{X}{U}$, the universal morphism $M^u(x)\in\Mo{M(X)}{U}$ of the pushout is well-formed if $x|_S\equiv M;u$.

Their semantics is defined by
\[\flt{M(X)}=\flt{T}\cup \{c[:M(t)][=M(d)]\;|\;c[:t][=d]\in\flt{X}\sm\flt{S}\}\]
\[\flt{(M^X)}=\flt{M}\cup\{c:=c\;|\;c\in\dom{X}\sm\dom{S}\}\]
\[\flt{M^u(x)}=\flt{u}\cup\{c:=x(c)\;|\;c\in\dom{X}\sm\dom{S}\}\]

Note that our definition of $\flt{m(X)}$ crucially exploits the condition that $\dom{X}\sm\dom{S}$ and $\dom{T}$ must be disjoint.
Thus, we obtain $\dom{X}\sm\dom{S}\sq\dom{m(X)}$ and $m^X$ maps these identifiers to themselves.

%(Note that if $M$ is implicit, then necessarily $m|_C\equiv\id{C}$.)

$m^X$ is implicit if $m$ is.\ednote{Can this always be implicit? Even if it can, it is probably not desirable. Compare left-/right-neutral element.}
\ednote{Should there be implicit extensions of morphism? E.g., $m^X$ could be implicit relative to $m$, i.e., we write $m$ and it is inferred as $m^X$. Does that yield canonical structures?}
$m^u(x)$ is implicit if $m$ and $u$ are.

%Pushouts are coherent:
%  $(m;n)(X)\equiv n(m(X))$ if the right hand side is defined.
%  If $X\harr Y$, then $m(X)\harr m(Y)$ if $m(Y)$ is defined.
%  $\id{S}(X)\equiv X$ (always defined).
%  $m(S)\equiv S$ (always defined).
%  
%Pushout over the empty morphism is defined iff $X$ and $T$ are disjoint.
%In that case, it is the union. (But not all unions arise in this way.)
%
%Pushout distributes over union:
%  $m(X\cup Y)\equiv m(X)\cup m(Y)$ (equi-defined)
%  $m^{X\cup Y}\equiv m^X\cup m^Y$ (?)
%  
%  does not make sense:
%  $(\bigcup m_i)(X)\equiv \bigcup \{m_i(X)\}$ 
%  $(\bigcup m_i)^X\equiv \bigcup\{m_i^X\}$
%

% Theorem: For every theory $T$, we have $\dom{T}=\dom{t_1}\cup\ldots\cup\dom{t_n}$ for atomic theories $t_i$.

### 2018-06-12 implicit.tex

\subsection{Motivation}

The motivation behind implicit morphisms is to generalize the benefits of include declarations even further.
Sometimes an atomic morphism is conceptually similar to an inclusion: For example, we might want to declare $m:\cn{Division}\to\cn{Group}=\{\cn{divide} := \lambda x,y.x\circ y^{-1}, \;\ldots\}$ to show how division is ``included'' into the theory of groups.
Other times we have to use structures because we need to duplicate declarations but we would still like to use them like includes.
For example, we have to declare the theory of rings using two structures (for the commutative group and the monoid).
But we still want a ring to be implicitly converted into, e.g., a commutative group.

Our key idea is to use a commutative{\footnotemark} subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
It contains all theories but only some of the morphisms -- the ones designated as \emph{implicit}.
Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ -- if this morphism is $M$, we write $\ipc{M}{S}{T}$.
\footnotetext{A diagram commutes if for any two morphisms $M,M'\in\Mo{S}{T}$, we have $M\equiv M'$.}
%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
%(This is well-defined because composition is associative and identity is neutral.)
%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.

The implicit-diagram generalizes the inclusion relation $\harr$.
In particular, all identity and inclusion morphisms are implicit, and we recover $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
Moreover, the relation ``exists $M$ such that $\ipc{M}{S}{T}$'' is also an order.

Consequently, many of the advantages of inclusions carry over to implicit morphisms.
\begin{compactitem}
\item It is very easy to maintain the implicit-diagram, e.g., as a partial map that assigns to a pair of theories the implicit morphism between them (if any).
\item We can generalize the visibility of identifiers: If $\ipc{M}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
\item We can still use canonical identifiers $s?n$. Because there can be at most one implicit morphism $\ipc{M}{s}{T}$, using $s?n$ as an identifier in $T$ is unambiguous.
Crucially, we can use $s?n$ without bothering what $M$ is.
\end{compactitem}

%Thus the main question arises what other morphisms we should make implicit.

\subsection{Definition}

\paragraph{Syntax}
We have already indicated that we will consider all identities and inclusions as implicit.
It remains to give users the option introduce additional implicit morphisms.
For that purpose, we allow two \mmt declarations to carry an attribute: 
\begin{grammar}
MDec   & Att\; m:T\to T=\{Ass,\ldots,Ass\}   & \text{attributed atomic morphism} \\
Dec    & Att\; \icl{S}                       & \text{attributed inclusion}\\
Att    & \keyword{implicit} \alt \keyword{invertible} & \text{attributes}
\end{grammar}

Every morphism with the attribute \keyword{implicit} occurs in the implicit-diagram.
(Of course, this is redundant for includes.)
Every morphism with the attribute \keyword{invertible} is also considered implicit, and additionally so is its inverse morphism.

We do not even have to provide a name for the inverse morphism: It is uniquely determined (up to $\equiv$) anyway, and it will be used only implicitly.
We only have to make sure that the inverse actually exists and that it can be found automatically.
This is not decidable in general, but it is decidable in many important special cases, in particular, renamings and definitional extensions.

\paragraph{Semantics}
Well-formedness and semantics are defined in parallel.

\begin{definition}[Implicit-Diagram]
Given a diagram $D$, its implicit-diagram $D^i$ contains all theories and the following morphisms:
\begin{compactitem}
\item all identities and inclusions,
\item all compositions of implicit morphisms,
\item every morphism with the \keyword{implicit} attribute,
\item for each morphism with the \keyword{invertible} attribute, that morphism and its inverse. 
\end{compactitem}
\end{definition}

The only requirement for well-formedness is that the implicit-diagram always commutes.
However, in general, commutativity may be undecidable.
Similarly, the existence of an inverse morphism may be undecidable.
Therefore, we use sound criteria for commutativity and invertibility as defined in Sect.~\ref{sec:commute} and \ref{sec:inverse}.

\begin{definition}[Well-Formedness]
Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.

Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
\end{definition}

Finally, we modify the definition of $\Exp{T}$ and $M(-)$ to add all implicitly imported identifiers:

\begin{definition}[Visible Declarations]
The set $\flti{T}$ contains the following declarations:
for every $\ipc{M}{S}{T}$ and every $c[:t][=d]$ in $\flt{S}$, the declaration $c[:M(t)]=M(c)$.

The set $\Expi{T}$ contains all objects formed from the symbols in $\flti{T}$.
The map $M(-)$ maps every $c$ for which $\flt{M}$ does not provide an assignment to $M(d)$ where $d$ is the definiens of $c$.
\end{definition}

This modification is conservative in the sense that $\flt{T}\sq\flti{T}$ and $\flti{T}\sm\flt{T}$ contains only declarations with definiens.
Moreover, $\Exp{T}\sq\Expi{T}$ and $M(-)$ remains unchanged on all $o\in\Exp{T}$.
It is easy to see that $M(-)$ still preserve all judgments.

\subsection{Commutativity}\label{sec:commute}

To prove commutativity, we have to find all paths in the implicit-diagram and check $M\equiv M'$ for all pairs of morphisms between the same nodes.

\mmt reduces $M\equiv M'$ for $M,M'\in\Mo{S}{T}$ to a first-order problem in the theory of categories, which uses the following axioms:
\begin{compactitem}
\item associativity of composition,
\item neutrality of identity,
\item $M;M'\equiv \id{S}$ as well as $M';M\equiv\id{T}$ if $M\in\Mo{S}{T}$ and $M'$ is the known *inverse of $M$,
\item if an atomic morphism $m$ includes $M\in\Mo{R}{T}$, then $M|_R\equiv M$.
\end{compactitem}

This is a sound but not complete axiomatization of $M\equiv M'$.
We can obtain a complete axiomatization in a totally different way: Check $\der_T M(c)\equiv M'(c)$ for all $c\in\dom{S}$.
However, this can be extremely expensive: It requires computing $\dom{S}$ and proving an object level equality for each identifier.
Even if the equality of objects is decidable, this is often too expensive.

In fact, because implicit morphisms are meant to be used in very particular cases, we expect the above incomplete calculus to perform reasonably well in practice.
If showing the equality of certain morphisms is hard, one should probably not make them implicit in the first place.

\subsection{Inverse Morphisms}\label{sec:inverse}

We only two use sufficient criteria to determine an *inverse of a morphism $M$.

\paragraph{Conservative Extensions}
Intuitively, $t$ extends $S$ definitionally, if it only adds defined symbols, e.g., abbreviations or theorems.
That is the most important case of conservative extensions.

\begin{definition}[Definitional Extension]
An include declaration \[t=\{\keyword{invertible}\;\icl{S},\Sigma\}\] is \textbf{definitional} if
\begin{compactitem}
 \item all symbol declarations in $\Sigma$ have a definiens,
 \item all include declarations $\icl{R}$ in $\Sigma$ are such that $R$ also definitionally includes $S$.
\end{compactitem}
\end{definition}

There are multiple, subtly different definitions of conservative extension.
However, definitional extensions as defined above are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable formal language).
In fact, they are always invertible: The inverse morphism $i:t\to S$ maps
\begin{compactitem}
 \item an identifier $c\in\dom{S}$ to $c$ and everythemselves and the others to their definiens,
 \item any other identifier, which must have a definiens $d$, to $i(d)$.
\end{compactitem}

%We use two auxiliary definitions.
%Given $u:T\to U$, we write $u|_S$ for the restriction of $u$ to $S$, i.e., the composition of the inclusion $S\harr T$ and $u$.
%An extension $S\to T$ is retractable if there is a morphism $r:T\to S$ such that $r|_S=\id{S}$.
%
%Whether a given extension $S\harr T$ is conservative depends on the specific definition of conservativity and the used logic.
%However, the most important kind of conservative extension is very easy to handle: the extension of a theory with a defined constant (e.g., an abbreviation or a theorem).
%These extensions are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable logic).
%In fact, they are always isomorphic to the original theory.
%
%Extensions with definition are almost exclusively used to add something to $S$.
%Therefore, it is reasonable to automatically designate the retraction as implicit.
%In particular, this provides an elegant way to add theorems to $S$ in such a way that they are available to any theory importing $S$, but without actually changing $S$.
%
%More generally, any retractable extension $S\harr T$ is always conservative.
%But if the retraction usually does not exist uniquely (Often an arbitrary non-canonical choice is needed, e.g., any type.), this is not enough to make the retraction implicit.

\paragraph{Renamings}
Intuitively, an atomic morphism $m:S\to T$ is a renaming if it maps all $S$-identifiers to $T$-identifiers (rather than $T$-objects).

\begin{definition}[Renaming]
Consider a morphism $M\in\Mo{S}{T}$.
Let $A\sq\dom{S}$ and $B\sq\dom{T}$ contain those identifiers without definiens.
$M$ is an (injective, surjective) \textbf{renaming} if the restriction of $M(-)$ is an (injective, surjective) mapping $A\to B$.
\end{definition}

Clearly, if $m$ is an injective and surjective renaming, it is straightforward to obtain its inverse by inverting $M(-)$.

\subsection{Applications}

Canonical isomorphisms.

User-declared atomic morphisms.

\paragraph{Extending Theories from the Outside}

\paragraph{Transparent Refactoring}
A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.

With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.

### 2018-06-12 include.tex

\subsection{Motivation}

\mmt as introduced in \cite{RK:mmt:10} allowed only two declarations in atomic theories: symbol declarations $c$ and structure declarations, which instantiate and import a theory.
(We omitted the latter in Sect.~\ref{sec:mmt}.)
We will now introduce include declarations to \mmt.

We can see include and structure declarations as duals.
Both import a previously declared theory, and both are transitive in the sense that if $T$ imports $S$, it also imports anything $S$ imports.
They differ in the sharing semantics when the same theory is imported multiple times: Multiple includes of $S$ are redundant and introduce a single copy of $S$; multiple structures of $S$ introduce multiple fresh copies of $S$.

Consequently, include declarations are \emph{unnamed} and the importing theory refers to the included symbols using the identifiers from $\dom{S}$.
Structure declarations on the other hand introduce a name in order to form qualified names for the fresh copy of the imported symbols.

\subsection{Definition}

We extend the \mmt grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into an atomic theory} \\
  Ass & \icl{M} & \text{include a morphism into an atomic morphism}
\end{grammar}

We extend the definitions of $\flt{\Sigma}$ with the case
  \[\flt{(\Sigma,\;\icl{S})} = \flt{\Sigma}\cup\flt{S}\]
and the definition of $\flt{\sigma}$ with the case
  \[\flt{(\sigma,\;\icl{M})}=\flt{\sigma}\cup\flt{M}\]
%  \[\flt{(\Sigma,\;\strc{s}{S})} = \flt{\Sigma}\cup\{t?s/n[:t'][=d'] | n[:t][=d]\in \flt{S}\}\]
%where $t'$ and $d'$ arise from $t$ and $d$ by renaming all symbol identifiers accordingly.
%  \[\flt{(\sigma,\;\strca{M})}=\flt{\sigma}\cup\{t?s/n:=M(\flt{M}\]
In particular, this yields $S\harr t$ whenever $t$ declares an include of $S$.

Includes between theories are always well-formed (assuming the included theory is).
But includes between morphisms are more difficult -- we have to impose two restrictions.
Firstly, an assignment $\icl{M}$ is well-formed in an atomic morphism $m:S\to T$ if $R\harr S$ and $M\in\Mo{R}{T}$.
Its intended effect is that $m$ maps all $c\in\dom{R}$ to $M(c)$.
That yields the second restriction: For any two morphisms $M_i\in\Mo{R_i}{T}$ included into $m$ and any theory $C$ satisfying $C\harr R_1$ and $C\harr R_2$, we require that $M_1|_C\equiv M_2|_C$.
Thus, any two included morphisms must agree on their common domain $C$; that is necessary so that $\flt{m}$ does not contain multiple assignments for the same identifier.

\subsection{Discussion}

The original structure declarations of \mmt are significantly more expressive than include declarations: They allow fine-tuning the sharing by instantiating some symbols of the imported theory.
In particular, they can emulate include declarations as a special case.

But the inclusion relation $S\harr T$ induced by include declarations has a number of practically valuable properties.
Therefore, the author chose to add include declarations even though they do not adding expressive strength.
Indeed, the practical advantages are huge.
For example, after implementing support for inclusions, the author refactored the LATIN atlas \cite{CHKMR:latinabs:11} -- a large case study on formalizing logics using an \mmt style module system.
At this point it contains approximately $2000$ imports, of which $1600$ were changed to inclusions.
\medskip

There are several reasons for these advantages.
Firstly, inclusion is an order relation between theories: Each theory can be included at most once into another one.
That makes the inclusion relation very easy and efficient to maintain.

Secondly, we can form canonical qualified identifiers $s?n$ for identifiers of $s$ included into a theory $t$.
These are canonical identifiers that uniquely identify the included symbols and that are obvious to users.
Users can easily see where a symbol was imported from, but the import chain along which it was imported remains transparent.
%A qualified identifier $q.c$ where $q$ is provided by the import.
%This quickly leads to long names like $p.q.r.c$. Moreover, because the import path is part of the name, refactoring becomes more costly and multiple qualified identifiers may actually be the same.

Thirdly, $\Exp{s}\sq\Exp{t}$ so that the included declarations do not have to be translated from $s$ to $t$.
Consequently, implementations do not have to duplicate the translations: They can store all local declarations with the declaring theory, and use the relation $s\harr t$ to determine which theory may access which declarations.
Tools that duplicate imported declarations on the other hand quickly suffer from exponential blowups when handling large libraries.

Another advantage is that inclusion allows a form of implicit conversion.
As explained in \cite{RK:mmt:10}, we can consider morphisms $M\in\Mo{T}{F}$ as models of $T$ defined in terms of $F$.
For example, consider $T=\cn{Group}$ and $F=\cn{ZFC}$; then concrete group corresponds to morphisms $G\in\Mo{T}{F}$.
If we further assume $\cn{Monoid}\harr \cn{Group}$, then $G|_{\cn{Monoid}}$ is $G$ seen as a monoid.
Moreover, we can easily make the conversion $G\mapsto G|_{\cn{Monoid}}$ implicit without confusing humans or tools.
\medskip

Of course the above advantages are not a new discovery.
Many languages have long used a combination of unnamed and named imports (e.g., inheritance and aggregation in object-oriented programming correspond to includes and structures).
However, the above appears to be the first time that the semantics of include declarations is defined formally in the context of OpenMath-style qualified identifiers.

%\begin{compactitem}
%\item The original identifier $c$. This quickly leads to name clashes.
%\end{compactitem}
%There is one special case, where choosing identifies is easy: If $S\harr T$, e.g., $S$ is included into $T$.
%Then we can use $S.c$ as a qualified identifier.
%
%The reason why $S.c$ works as a qualified identifier is that $S$ can only be included into $T$ once and in a uniquely determined way.
%Implicit morphisms generalize this property: $S$ can only be implicitly mapped into $T$ once and in a that is uniquely determined by the implicit-diagram.
%Consequently, we can use any declaration $c$ of $S$ in $T$ using the identifier $S.c$.

### 2018-06-12 intro.tex

The contribution of this paper is threefold.

Firstly, we describe the syntax and semantics of include declarations in \mmt.
Our main interest here is that inclusion are the most important form of implicit morphisms.
However, this is also the first item that the semantics of includes in MMT is specified rigorously.

Secondly, we generalize inclusions to implicit morphisms and show how we can use implicit morphisms to structure large diagrams of theories.
Most notably, we show that many of the goals of realms \cite{realms} (which won the best paper award at MKM 2014) are already realized by \mmt in the presence of implicit morphisms.

Thirdly, we give an extension of \mmt with complex theory and morphism expressions.
Our focus here is on defining which of these complex morphisms can be designated as implicit.
However, the language is novel and a valuable contribution in itself.

### 2018-06-12 mmt.tex

The \mmt language \cite{RK:mmt:10} provides an extremely general setting in which we can define theories.
The precise definitions of the well-formed theories $T$ and theory morphisms $M$ on the one hand and their semantics $\flt{T}$ and $\flt{M}$ on the other hand are \textbf{mutually recursive}.
Therefore, we first give an overview and then recap the definitions below.

\textbf{Flat theories} are lists of symbol declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
\textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of symbol assignments $c:=e$ where $c$ is declared in $S$ and $e$ is an expression over $T$.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Every morphism $M$ induces a \textbf{homomorphic extension} $M(-):\Exp{S}\to\Exp{T}$, which replaces every $c\in\dom{S}$ in an $S$-expressions with the $T$-expressions $o$ such that $(c:=o) \in \flt{M}$.
Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.

Flat theories and morphisms are rarely constructed in practice --- instead, users often construct larger theories from smaller ones.
We distinguish two kinds of constructions:
\begin{compactitem}
 \item \textbf{Structured theories} contain additional declarations such imports and instantiations or high-level declarations.
 \item \textbf{Theory expressions} are operators that take (normal or theory) expressions as input and return theory expressions.
\end{compactitem}
This distinction is orthogonal to the semantic variety, and both kinds are equally expressive, e.g., we can construct the union of two theories using either construction.
For example, we can construct the theory of commutative groups as the structured theory containing the declarations $\icl{Group},\icl{CommutativeMagma}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
All of the above applies analogously to structured \emph{morphisms} and \emph{morphism} expressions.

Despite the syntactic difference, the semantics of both kinds of constructions can be defined by uniformly by mapping structured theories/theory expressions $T$ to flat theories $\flt{T}$.
We call this mapping \textbf{flattening}.

An MMT \textbf{diagram} consists of a set of structured theory/morphism declarations and induces sets of theory and morphism expressions.
The names of the former are the \textbf{atomic theories/morphisms}, and they provide the base case of the inductive construction of the latter.
Vice versa, the latter may already occur in the bodies of the structured declarations.
Thus, the two concepts must be defined by mutual recursion.

More formally, every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of atomic theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$ (seen as a set of declarations),
 \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$ (seen as a set of assignments).
\end{compactitem}

Within this framework, we can define new kinds of structuring declarations and new kinds of expression-forming operators along with their flattening.
In both cases, the semantics should be \textbf{compositional} in the following sense:
The flattening of a structured theory arises as the union of the flattenings of each declaration in it.
More formally, for a list $D$ of declarations and a declaration $d$, we have $\flt{D,d}=\flt{D}\cup\fltd{\flt{D}}{d}$ where $\fltd{T}{d}$ defines the semantics of declaration $d$ in the context of theory $T$.
Similarly, the flattening of a theory expression such as $f(S,T)$ depends only on the flattenings of its arguments
More formally, e.g., for theory expressions $S$ and $T$ and a binary theory-forming operator $f$, we have $(\flt{f(S,T)}=\flt{f}(\flt{S},\flt{T})$ where $\flt{f}$ is a function defining the semantics of $f$.

\subsection{Syntax}

We use the non-modular variant of \mmt, which does not use imports and instantiations between theories.
The full version can be found in \cite{RK:mmt:10}.

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{atomic theory declaration} \\
Dec      & n\opt{:o}\opt{=o}     & \text{symbol declaration}\\
c        & t?n                   & \text{qualified symbol identifiers} \\
o        & c \alt \ldots         & \text{OpenMath objects built from symbols $c$} \\
T        & t                     & \text{theories} 
\end{grammar}

In an atomic theory, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed OpenMath objects using only previously introduced symbols.
We omit the remaining productions for objects here.
\end{definition}

These theories are similar to OpenMath content dictionaries \cite{openmath}.
In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
But the interface always knows the qualified identifier and shows it, e.g., on hover.}
Thus, symbols of the same name declared in different atomic theories are always distinguished.

Symbol declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
Similarly, objects subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: They single out the well-formed objects by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

\begin{definition}[Morphism]\label{def:morphism}
For two theories $S$ and $T$, a \textbf{morphism} $M$ is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{atomic morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m \alt \id{T} \alt M;M          & \text{morphisms} 
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

Crucially, morphisms \textbf{preserve judgments}, in particular they map theorems to theorems.
Formally, an atomic morphism $m:S\to T$ is \emph{well-formed} if
\begin{compactitem}
\item whenever $c$ has type $t$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o:m(t)$,
\item whenever $c$ has definiens $d$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o\equiv m(d)$.
\end{compactitem}
Then we can show that for every $M\in\Mo{S}{T}$ if $\der_S o:o'$ (or $\der_S o\equiv o'$), then $\der_T M(o):M(o')$ (or $\der_T M(o)\equiv M(o')$).
The details can be found in \cite{rabe:howto:14}.
%In particular, this means that $M$ maps $S$-theorems to $T$-theorems.

\begin{definition}[Diagram]
A \textbf{diagram} is a list of atomic theory and morphism declarations:
\begin{grammar}
Dia    & \rep{(TDec \alt MDec)}          & \text{diagrams}\\
\end{grammar}
A diagram is well-formed if each atomic theory/morphism declaration is well-formed relative to the diagram preceding it.
\end{definition}

\subsection{Semantics}

\paragraph{Induced Declarations and Assignments}
For every atomic theory $t=\{\Sigma\}$, we have $t\in\Thy$.
We define $\flt{t}$ by induction on the declarations in $\Sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\Sigma,\;n[:t][=d])}=\flt{\Sigma}\cup\{t?n[:t][=d]\}\]
where $\cdot$ is the empty sequence.
Correspondingly, for every atomic morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;c:=o)}=\flt{\sigma}\cup\{c:=o\}\]

For the remaining morphisms, the semantics is not as trivial but still straightforward.
For the identity, we have $\id{T}\in\Mo{T}{T}$ if $S\in\Thy$, and $\flt{\id{T}}=\{c:=c|c\in\dom{T}\}$.
For the composition, we have $M;N\in\Mo{R}{T}$ if $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, and $\flt{(M;N)}=\{c:=N(M(c))|c\in\dom{R}\}$.

\paragraph{Equality and Inclusion}
We say that $S$ is \textbf{included} into $T$, written $S\harr T$, if $\flt{S}\sq\flt{T}$.
%In that case, there is an inclusion morphism $S\to T$ (which we also denote $S\harr T$) defined by $\flt{m}=\flt{\id{S}}$.
%Clearly, $m(o)=o$ for all $S$-objects $o$.
Thus, every declaration of $S$ is also available in $T$ and $\Exp{S}\sq\Exp{T}$.
This induces an inclusion morphism, which maps every $c\in\dom{S}$ to itself.
We do not introduce a name for that morphism and simply define $\id{S}\in\Mo{S}{T}$ whenever $S\harr T$ (which includes the special case $S=T$).
Moreover, if $S\harr T$ and $M\in\Mo{T}{U}$, we define the restriction $M|_S\in\Mo{S}{U}$ of $M$ to $S$ by $\id{S};M$.

We say that two theories $S,T\in\Thy$ are \textbf{equal}, written $S\equiv T$, if $\flt{S}=\flt{T}$.
This is equivalent to $S\harr T$ and $T\harr S$.

Correspondingly, we say that two morphisms $M,N\in\Mo{S}{T}$ are \textbf{equal}, written $M\equiv N$, if $\flt{M}=\flt{N}$.
In that case, $\der_T M(o)\equiv N(o)$ for all well-formed $S$-objects $o\in\Exp{S}$.

Clearly, $\equiv$ is an equivalence relation and $\harr$ is an order relation (if anti-symmetry is taken with respect to $\equiv$).
Identity and composition are congruent with respect to $\equiv$.
Moreover, composition is associative and identity neutral with respect to $\equiv$.
Thus, each diagram yields the structure of a category.

\paragraph{Preview}
The above recap of (the non-modular fragment of) \mmt deviates from \cite{RK:mmt:10}, very carefully wording the definitions in a way that makes stating the results of this paper most elegant.
Thus, we intentionally introduced some peculiarities:
\begin{compactitem}
\item We have $\Thy=\thy$ because there are no non-atomic theories.
We will introduce some in Sect.~\ref{sec:complex}.
\item The definitions of $\flt{t}$ and $\flt{m}$ are overly complicated.
That will pay off when we introduce additional declarations in Sect.~\ref{sec:include}: We can define their semantics simply by adding a case to the inductions in $\flt{\Sigma}$ and $\flt{\sigma}$.
\item $s\harr t$ never holds because all $c\in\dom{s}$ are of the form $s?n$ and all $c\in\dom{t}$ are of the form $t?n$.
In fact, $\flt{S}$ and $\flt{T}$ are disjoint whenever $S\neq T$.
That will change when we introduce include declarations in Sect.~\ref{sec:include}.
\end{compactitem}

### 2018-06-12 paper.tex

\documentclass[orivec]{llncs}

\usepackage{graphicx,url,wrapfig,amsmath,amssymb}

\usepackage[show]{ed} %for ednotes

\usepackage{sembracket}
\usepackage{mytikz}
\usepackage{basics}
%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
\usepackage{local}

\setcounter{tocdepth}{3}
\usepackage[bookmarks,linkcolor=red,citecolor=blue,urlcolor=gray,colorlinks,breaklinks,bookmarksopen,bookmarksnumbered]{hyperref}

\setlength{\hfuzz}{3pt}
\hbadness=10001 %make box warning less strict

\begin{document}
\title{Structuring Theories with Implicit Morphisms}
\author{Florian Rabe$^{1,2}$ \and Dennis M\"uller$^2$}
\institute{LRI Paris \and FAU Erlangen-Nuremberg}
\maketitle

\begin{abstract}
We introduce \emph{implicit} morphisms as a concept in formal systems based on theories and theory morphisms.
The idea is that there may be at most one implicit morphism from a theory $S$ to a theory $T$, and if $S$-expressions are used in $T$ their semantics is obtained by automatically inserting the implicit morphism.
The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
\end{abstract}

\input{wadtabstract}
\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
\end{document}

\section{Introduction}
\input{intro}

\section{Theories and Theory Morphism}
\input{mmt}

\section{Inclusion Morphisms}\label{sec:include}
\input{include}

\section{Implicit Morphisms}
\input{implicit}

\section{Complex Theories and Morphisms}\label{sec:complex}
\input{complex}

\section{Related Work}
\input{related}

\section{Conclusion}
\input{conc}

\bibliographystyle{alpha}
\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}

\end{document}

### 2018-06-12 related.tex

\subsection{Realms}

\cite{realms} defines a realm (essentially) as a certain cluster of theories:
\begin{compactitem}
\item a set of isomorphic theories $B_1,\ldots,$
\item for each $B_i$, a set of retractable extensions $E^i_1,\ldots,$
\item a theory $F$ including all of the above.
\end{compactitem}
The intuition is that the $B_i$ are alternative definitions of a theory; the $E^i_j$ add definitions and theorems; and $F$ aggregates everything into the outwardly visible interface (the face) of the realm.

A good example is the realm of topological spaces with the $B_i$ corresponding to the various possible definitions (via neighborhoods, open sets, closure operator, etc.).

Users of the realm include only the face and thus (i) gain access to all derived knowledge in the theory of topological spaces, and (ii) do not have to commit to one particular definition.
Indeed, this corresponds more closely to the way how conventional mathematics sees the theory of topological spaces.

\cite{realms} calls for an implementation of realms as a new primitive concept in addition to theories and morphisms.
However, in the presence of implicit morphisms, we do not need it: All we have to do is to designate the isomorphisms and the retraction is implicit and include any one of the $B_i$ into $F$.

\subsection{Canonical Structures in Coq}

\cite{gonthier_packaging}

\subsection{Implicit Conversions in Scala}

The Scala programming language \cite{scala} has a clean implementation of implicit conversions between any two types.
If we think of functions between classes as theory morphisms, this corresponds to implicit theory morphisms.

Contrary to \mmt, a conversion is only available if explicitly imported into the current namespace.
Moreover, Scala does not chain conversions, i.e, the composition of implicit morphisms is not implicit.
Scala's limitations are intentional: Their goal is to prevent programmers from overusing implicit conversions because that can be very confusing to readers.

These limitations are lifted in \mmt's conversions can only act between theories anyway.
Thus, their potential to confuse is smaller.
Moreover, the use of qualified identifiers in \mmt means that readers can always tell where an identifier is implicitly-included from, and the \mmt IDE can always show the implicit morphism along which it is imported.

\ednote{drawback: global data structure}

### 2018-06-12 wadtabstract.tex

Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps}, and knowledge representation \cite{RK:mmt:10}.

%The latter of these tools is our \mmt system.
%We use \mmt as both a theoretical framework and a concrete implementation of our results.
%But our ideas are independent of \mmt and can be transferred very easily to any other language using theory morphisms.

These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
This diagram is generated by all theories/morphisms defined, induced, or referenced in a user's development.

\begin{wrapfigure}{r}{5.5cm}
\vspace{-3em}
\begin{tikzpicture}
\node (M) at (0,0) {$\cn{Monoid}$};
\node (G) at (-2,-2) {$\cn{Group}$};
\node (D) at (2,-2) {$\cn{DivGroup}$};
\draw[mono](M) -- (G);
\draw[arrow](G) to[out=30,in=150] node[above] {$\cn{G2DG}$} (D);
\draw[arrow](D) to[out=-150,in=-30] node[above] {$\cn{DG2G}$} (G);
\end{tikzpicture}
\vspace{-3.5em}
\end{wrapfigure}

For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
This would result in the diagram on the right.
Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
Crucially, $m(-)$ preserves typing and provability.

The key idea behind implicit morphisms is very simple:
We maintain an additional diagram $I$, which is commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipce{S}{T}$.
Commutativity makes the following language extension well-defined: if $\ipce{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions; and if $c$ is used in a $T$-expression, the semantics of $c$ is $i(c)$ where $i$ is the uniquely determined implicit morphism $i:S\to T$.

Despite their simplicity, the practical implications of implicit morphism are huge.
For example, in the diagram above, we may choose to label $\cn{G2DG}$ implicit.
Immediately, every abbreviation or theorem that we have formulated in the theory $\cn{Group}$ becomes available for use in $\cn{DivGroup}$ without any syntactic overhead.
We can even label $\cn{DG2G}$ implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that $\cn{Group}$ and $\cn{DivGroup}$ are just different formalizations of the same concept.
While these morphisms must be labeled manually, any inclusion morphism like the one from $\cn{Monoid}$ to $\cn{Group}$ is implicit automatically.

In fact, this principle works so well that we have refactored our \mmt system (our long-standing implementation) in such a way that implicit morphisms are now more primitive than inclusion morphisms.
The semantics of inclusion morphisms is obtained by saying that inclusions are implicit morphisms that map all identifiers to themselves.
Even the fundamental property that a theory may reference its own identifiers is now just a consequence of the fact that all identity morphisms are implicit.
Therefore, surprisingly, adding implicit morphisms deep in the \mmt kernel has made its design more elegant.

### 2018-09-26 intro.tex
rewritten Sect. 1+2

@@
-The contribution of this paper is threefold.
+\paragraph{Motivation}
+Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
+They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
+Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps}, and knowledge representation \cite{RK:mmt:10}.
 
-Firstly, we describe the syntax and semantics of include declarations in \mmt.
-Our main interest here is that inclusion are the most important form of implicit morphisms.
-However, this is also the first item that the semantics of includes in MMT is specified rigorously.
+%The latter of these tools is our \mmt system.
+%We use \mmt as both a theoretical framework and a concrete implementation of our results.
+%But our ideas are independent of \mmt and can be transferred very easily to any other language using theory morphisms.
 
-Secondly, we generalize inclusions to implicit morphisms and show how we can use implicit morphisms to structure large diagrams of theories.
-Most notably, we show that many of the goals of realms \cite{realms} (which won the best paper award at MKM 2014) are already realized by \mmt in the presence of implicit morphisms.
+These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
+This diagram is generated by all theories/morphisms defined, induced, or referenced in a user's development.
 
-Thirdly, we give an extension of \mmt with complex theory and morphism expressions.
-Our focus here is on defining which of these complex morphisms can be designated as implicit.
-However, the language is novel and a valuable contribution in itself.
+\begin{wrapfigure}{r}{5.5cm}
+\vspace{-3em}
+\begin{tikzpicture}
+\node (M) at (0,0) {$\cn{Monoid}$};
+\node (G) at (-2,-2) {$\cn{Group}$};
+\node (D) at (2,-2) {$\cn{DivGroup}$};
+\draw[mono](M) -- (G);
+\draw[arrow](G) to[out=30,in=150] node[above] {$\cn{G2DG}$} (D);
+\draw[arrow](D) to[out=-150,in=-30] node[above] {$\cn{DG2G}$} (G);
+\end{tikzpicture}
+\vspace{-3.5em}
+\end{wrapfigure}
+
+For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
+This would result in the diagram on the right.
+Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
+Crucially, $m(-)$ preserves typing and provability.
+
+The key idea behind implicit morphisms is very simple:
+We maintain an additional diagram $I$, which is commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
+The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
+Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions; and if $c$ is used in a $T$-expression, the semantics of $c$ is $i(c)$ where $i$ is the uniquely determined implicit morphism $i:S\to T$.
+
+Despite their simplicity, the practical implications of implicit morphism are huge.
+For example, in the diagram above, we may choose to label $\cn{G2DG}$ implicit.
+Immediately, every abbreviation or theorem that we have formulated in the theory $\cn{Group}$ becomes available for use in $\cn{DivGroup}$ without any syntactic overhead.
+We can even label $\cn{DG2G}$ implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that $\cn{Group}$ and $\cn{DivGroup}$ are just different formalizations of the same concept.
+While these morphisms must be labeled manually, any inclusion morphism like the one from $\cn{Monoid}$ to $\cn{Group}$ is implicit automatically.
+
+\paragraph{Contribution}
+We present a formal system for developing structured theories with implicit morphisms.
+Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
+\mmt is foundation-independent in the sense that it allows embedding a large variety of declarative languages (logics, type-theories, etc.).
+Therefore, all our results are foundation-independent --- users of a particular language $L$ can use \mmt directly as a structuring formalism for $L$ or can easily transfer our results to a dedicated implementation of $L$.
+
+More precisely, out contribution is twofold.
+
+Firstly, we give a novel presentation of the syntax and semantics \mmt.
+Contrary to the original presentation in \cite{RK:mmt:10}, which hard-coded a single structuring mechanism --- there called \textit{structures} --- our definitions allows for adding arbitrary independent structuring mechanisms.
+This required a from-scratch redesign of the original presentation in \cite{RK:mmt:10} (which is why this paper has no preliminaries section).
+
+Based on this extensible definition, we can introduce several additional structuring mechanisms.
+Many of these have already been implemented in the \mmt tool but were not formally defined  a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
+With our reformulation, this is now not only possible but very easy and elegant.
+As an example, we give includes and unions of theories, which induce the most important special cases of implicit morphisms.
+
+Secondly, we describe implicit morphisms as a novel feature of \mmt and give a number of examples.
+In particular, we recover the concept of realms \cite{realms} as a special case of implicit morphisms.
+
+In fact, implicit morphisms work so well that we have refactored the \mmt tool in such a way that implicit morphisms are now more primitive than inclusion morphisms.
+The semantics of inclusion morphisms is obtained by saying that inclusions are implicit morphisms that map all identifiers to themselves.
+Even the fundamental property that a theory may reference its own identifiers is then just a consequence of the fact that all identity morphisms are implicit.
+Therefore, surprisingly, adding the new feature of implicit morphisms to the \mmt kernel has made its design much simpler.

### 2018-09-26 mmt.tex
rewritten Sect. 1+2

@@
-The \mmt language \cite{RK:mmt:10} provides an extremely general setting in which we can define theories.
-The precise definitions of the well-formed theories $T$ and theory morphisms $M$ on the one hand and their semantics $\flt{T}$ and $\flt{M}$ on the other hand are \textbf{mutually recursive}.
-Therefore, we first give an overview and then recap the definitions below.
+\subsection{Overview}
 
-\textbf{Flat theories} are lists of symbol declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
-\textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of symbol assignments $c:=e$ where $c$ is declared in $S$ and $e$ is an expression over $T$.
+\paragraph{Modularity in MMT}
+As much as possible, theories and morphisms are treated uniformly, and we use the word \textbf{module} to refer to them collectively.
+\mmt allows constructing large modules from small ones in two ways:
+\begin{compactitem}
+  \item \textbf{structured modules} contain declarations that import other, previously defined ones,
+  \item \textbf{module expressions} are anonymous expressions (akin to formulas, terms, etc.) that denote modules.
+\end{compactitem}
+
+These two kinds of constructions have very different advantages but the same expressivity.
+For example, we can construct the theory of commutative groups as the structured theory $\cn{CommutativeGroup}=\{\icl{Group},\icl{CommutativeMagma}\}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
+The former is more practical when building named toplevel theories in software systems because each include is a separate declaration that can be parsed, checked, and flattened individually.
+While negligible in this example, this becomes relevant quickly in large theories with complex structure.
+The latter is more practical when building anonymous theory that are only needed temporarily.
+In software systems, they also allow reusing operations like equality and $\lambda$-abstraction that already be implemented at the level of the base logic.
+Building large morphism is as important as building large theories: for example, we want to be able to build morphisms out of $\cn{CommutativeGroup}$ by combining compatible morphisms out of $\cn{Group}$ and $\cn{CommutativeMagma}$.
+
+\paragraph{Flat Modules}
+\mmt provides formal syntax for both structured modules and module expressions and defines their semantics via \textbf{flattening}, which defines for every module $M$ a flat module $\flt{M}$.
+
+\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
 We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
-Every morphism $M$ induces a \textbf{homomorphic extension} $M(-):\Exp{S}\to\Exp{T}$, which replaces every $c\in\dom{S}$ in an $S$-expressions with the $T$-expressions $o$ such that $(c:=o) \in \flt{M}$.
+Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
+In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
+Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.
+
+Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
+%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
+The details can be found in \cite{rabe:howto:14}.
+
+\textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of assignments $c:=e$ where $c\in\dom{S}$ and $e\in\Exp{T}$.
+Every morphism $M$ induces a \textbf{homomorphic extension} $M(-):\Exp{S}\to\Exp{T}$, which replaces every $c\in\dom{S}$ in an $S$-expression with the $T$-expression $e$ such that $c:=e$ in $M$.
+
 Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.
+We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
+This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.
 
-Flat theories and morphisms are rarely constructed in practice --- instead, users often construct larger theories from smaller ones.
-We distinguish two kinds of constructions:
+An \mmt \textbf{diagram} consists of a set of named structured module declarations and induces sets of anonymous module expressions.
+These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
+Every diagram induces sets
 \begin{compactitem}
- \item \textbf{Structured theories} contain additional declarations such imports and instantiations or high-level declarations.
- \item \textbf{Theory expressions} are operators that take (normal or theory) expressions as input and return theory expressions.
-\end{compactitem}
-This distinction is orthogonal to the semantic variety, and both kinds are equally expressive, e.g., we can construct the union of two theories using either construction.
-For example, we can construct the theory of commutative groups as the structured theory containing the declarations $\icl{Group},\icl{CommutativeMagma}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
-All of the above applies analogously to structured \emph{morphisms} and \emph{morphism} expressions.
-
-Despite the syntactic difference, the semantics of both kinds of constructions can be defined by uniformly by mapping structured theories/theory expressions $T$ to flat theories $\flt{T}$.
-We call this mapping \textbf{flattening}.
-
-An MMT \textbf{diagram} consists of a set of structured theory/morphism declarations and induces sets of theory and morphism expressions.
-The names of the former are the \textbf{atomic theories/morphisms}, and they provide the base case of the inductive construction of the latter.
-Vice versa, the latter may already occur in the bodies of the structured declarations.
-Thus, the two concepts must be defined by mutual recursion.
-
-More formally, every diagram induces sets
-\begin{compactitem}
- \item $\thy$ of identifiers of atomic theories $t$,
+ \item $\thy$ of identifiers of theories $t$,
  \item $\Thy\supseteq \thy$ of theory expressions $T$,
@@
 \begin{compactitem}
- \item to each $T\in\Thy$ the flat theory $\flt{T}$ (seen as a set of declarations),
- \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$ (seen as a set of assignments).
+ \item to each $T\in\Thy$ the flat theory $\flt{T}$,
+ \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
 \end{compactitem}
-
-Within this framework, we can define new kinds of structuring declarations and new kinds of expression-forming operators along with their flattening.
-In both cases, the semantics should be \textbf{compositional} in the following sense:
-The flattening of a structured theory arises as the union of the flattenings of each declaration in it.
-More formally, for a list $D$ of declarations and a declaration $d$, we have $\flt{D,d}=\flt{D}\cup\fltd{\flt{D}}{d}$ where $\fltd{T}{d}$ defines the semantics of declaration $d$ in the context of theory $T$.
-Similarly, the flattening of a theory expression such as $f(S,T)$ depends only on the flattenings of its arguments
-More formally, e.g., for theory expressions $S$ and $T$ and a binary theory-forming operator $f$, we have $(\flt{f(S,T)}=\flt{f}(\flt{S},\flt{T})$ where $\flt{f}$ is a function defining the semantics of $f$.
 
@@
 
-We use the non-modular variant of \mmt, which does not use imports and instantiations between theories.
-The full version can be found in \cite{RK:mmt:10}.
+We start with the syntax for theories (which arises as a special case of the one given in \cite{RK:mmt:10}):
 
@@
 \begin{grammar}
-TDec     & t=\{Dec,\ldots,Dec\}  & \text{atomic theory declaration} \\
-Dec      & n\opt{:o}\opt{=o}     & \text{symbol declaration}\\
-c        & t?n                   & \text{qualified symbol identifiers} \\
-o        & c \alt \ldots         & \text{OpenMath objects built from symbols $c$} \\
-T        & t                     & \text{theories} 
+TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
+Dec      & n\opt{:o}\opt{=o}     & \text{constant declaration}\\
+c        & t?n                   & \text{qualified constant identifiers} \\
+o        & c \alt \ldots         & \text{expressions built from constants} \\
+T        & t                     & \text{theory expressions}
 \end{grammar}
 
-In an atomic theory, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed OpenMath objects using only previously introduced symbols.
-We omit the remaining productions for objects here.
+In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions only previously introduced constants.
+We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
 \end{definition}
 
-These theories are similar to OpenMath content dictionaries \cite{openmath}.
-In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
-\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
-But the interface always knows the qualified identifier and shows it, e.g., on hover.}
-Thus, symbols of the same name declared in different atomic theories are always distinguished.
+At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
+It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
+But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding production for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
+This makes it easy to incrementally add new declaration kinds and operators to the language.
 
-Symbol declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
-In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
-Similarly, objects subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.
+%These theories are similar to OpenMath content dictionaries \cite{openmath}.
+%In particular, \mmt uses OpenMath-style qualified symbol identifiers $t?n$.%
+%\footnote{Of course, the \mmt IDE and web interface parse and display these identifiers only as $n$ (if that is unambiguous), especially if they are used in formulas.
+%But the interface always knows the qualified identifier and shows it, e.g., on hover.}
+%Thus, symbols of the same name declared in different atomic theories are always distinguished.
 
-Individual formal languages arise as fragments of \mmt: They single out the well-formed objects by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
-%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
-The details can be found in \cite{rabe:howto:14}.
+We proceed accordingly for flat morphisms:
 
 \begin{definition}[Morphism]\label{def:morphism}
-For two theories $S$ and $T$, a \textbf{morphism} $M$ is of the form
+A \textbf{morphism} is of the form
 \begin{grammar}
-MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{atomic morphism declaration}\\
+MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
 Ass    & c:=o                            & \text{assignment to symbol}\\
-M      & m \alt \id{T} \alt M;M          & \text{morphisms} 
+M      & m                               & \text{morphism expressions}
 \end{grammar}
@@
 
-Crucially, morphisms \textbf{preserve judgments}, in particular they map theorems to theorems.
-Formally, an atomic morphism $m:S\to T$ is \emph{well-formed} if
-\begin{compactitem}
-\item whenever $c$ has type $t$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o:m(t)$,
-\item whenever $c$ has definiens $d$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o\equiv m(d)$.
-\end{compactitem}
-Then we can show that for every $M\in\Mo{S}{T}$ if $\der_S o:o'$ (or $\der_S o\equiv o'$), then $\der_T M(o):M(o')$ (or $\der_T M(o)\equiv M(o')$).
-The details can be found in \cite{rabe:howto:14}.
-%In particular, this means that $M$ maps $S$-theorems to $T$-theorems.
+%Crucially, morphisms \textbf{preserve judgments}, in particular they map theorems to theorems.
+%Formally, an atomic morphism $m:S\to T$ is \emph{well-formed} if
+%\begin{compactitem}
+%\item whenever $c$ has type $t$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o:m(t)$,
+%\item whenever $c$ has definiens $d$ in $\flt{S}$ and $c:= o$ in $\flt{m}$, then $\der_T o\equiv m(d)$.
+%\end{compactitem}
+%Then we can show that for every $M\in\Mo{S}{T}$ if $\der_S o:o'$ (or $\der_S o\equiv o'$), then $\der_T M(o):M(o')$ (or $\der_T M(o)\equiv M(o')$).
+%The details can be found in \cite{rabe:howto:14}.
+%%In particular, this means that $M$ maps $S$-theorems to $T$-theorems.
+
+Finally, we define diagrams as collections of modules:
 
 \begin{definition}[Diagram]
-A \textbf{diagram} is a list of atomic theory and morphism declarations:
+A \textbf{diagram} is a list of theory and morphism declarations:
 \begin{grammar}
@@
 \end{grammar}
-A diagram is well-formed if each atomic theory/morphism declaration is well-formed relative to the diagram preceding it.
+A diagram is well-formed if each atomic theory/morphism declaration has a unique name is well-formed relative to the diagram preceding it.
 \end{definition}
+
+We give some examples how to add modularity principles to the grammar:
+
+\begin{example}[Includes]\label{syn:incl}
+We extend the grammar with
+\begin{grammar}
+  Dec & \icl{S} & \text{include a theory into a theory} \\
+  Ass & \icl{M} & \text{include a morphism into a morphism}
+\end{grammar}
+\end{example}
+
+\begin{example}[Union]\label{syn:union}
+We extend the grammar with
+\begin{grammar}
+  T   & T\cup T & \text{union of theories} \\
+  M   & M\cup M & \text{union of morphisms}
+\end{grammar}
+\end{example}
+
+\begin{example}[Category of Theories]\label{syn:cat}
+To obtain the category structure of theories and morphisms, we need identity and composition.
+We add them to the grammar as follows:
+\begin{grammar}
+  M   & \id{T} & \text{identity morphism} \\
+  M   & M;M    & \text{morphism composition}
+\end{grammar}
+\end{example}
+
+For each example, we will define the semantics in the next section.
 
@@
 
-\paragraph{Induced Declarations and Assignments}
-For every atomic theory $t=\{\Sigma\}$, we have $t\in\Thy$.
-We define $\flt{t}$ by induction on the declarations in $\Sigma$:
+\paragraph{Flattening}
+Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
+In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
+This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.
+
+For theories $T$, at this point, this induction only has the base case of a reference to a declared theory $t$.
+Further cases arise when we add structuring declarations.
+
+For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
+\[\flt{t}=\flt{\Sigma}\]
+where
  \[\flt{\cdot}=\es\]
- \[\flt{(\Sigma,\;n[:t][=d])}=\flt{\Sigma}\cup\{t?n[:t][=d]\}\]
-where $\cdot$ is the empty sequence.
-Correspondingly, for every atomic morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
+ \[\flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
+Here $\fltd{T}{D}$ is the flattening of declaration $d$ in the context of theory $T$.
+
+At this point, we only have one case for declarations $D$, namely constant declarations.
+Their flattening is trivial:
+ \[\fltd{\Sigma}{n[:E][=e]}=\{t?n[:E][=e]\}\]
+where $t$ is the name of the containing theory.
+
+Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
  \[\flt{\cdot}=\es\]
- \[\flt{(\sigma,\;c:=o)}=\flt{\sigma}\cup\{c:=o\}\]
+ \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
+with the trivial base case
+ \[\fltd{\sigma}{c:=e}=\{c:=o\}\]
+\bigskip
 
-For the remaining morphisms, the semantics is not as trivial but still straightforward.
-For the identity, we have $\id{T}\in\Mo{T}{T}$ if $S\in\Thy$, and $\flt{\id{T}}=\{c:=c|c\in\dom{T}\}$.
-For the composition, we have $M;N\in\Mo{R}{T}$ if $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, and $\flt{(M;N)}=\{c:=N(M(c))|c\in\dom{R}\}$.
+The semantics serves two purposes:
+firstly, it flattens every well-formed module into a flat one;
+secondly, it determines which modules are well-formed at all.
+It is desirable to define well-formedness independently of flattening.
+However, in practice, it shows that implementations typically have to do almost the entire flattening to check well-formedness anyway.
 
-\paragraph{Equality and Inclusion}
-We say that $S$ is \textbf{included} into $T$, written $S\harr T$, if $\flt{S}\sq\flt{T}$.
-%In that case, there is an inclusion morphism $S\to T$ (which we also denote $S\harr T$) defined by $\flt{m}=\flt{\id{S}}$.
+Therefore, it is convenient to merge the two into one operation by simply making flattening a partial function: a module $X$ is \textbf{well-formed} iff $\flt{X}$ is defined.
+\bigskip
+
+\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
+Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
+We do this as follows
+  \[\fltd{\Sigma}{\icl{S}} = \flt{S}\]
+This has the effect of copying over all declarations from the included into the including theory.
+
+Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
+Therefore, includes can never lead to name clashes.
+
+Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
+The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
+Therefore, flattening and well-formedness must be distinguished:
+  \[\fltd{\sigma}{\icl{M}} = \flt{M}\]
+under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
+\end{example}
+
+\begin{example}[Union (continued from Ex.~\ref{syn:union})]\label{sem:union}
+We define the semantics of unions of theories as
+\[\flt{(S\cup T)}=\flt{S}\cup\flt{T}\]
+The handling of name clashes is the same as for includes:
+Because the theories declared in a diagram have unique names and declarations in the flattening are qualified by theory names, union theories cannot have name clashes.
+But we have to watch out when taking unions of morphisms:
+We define $M_1\cup M_2\in \Mo{S_1\cup S_2}{T}$ if $M_i\in\Mo{S_i}{T}$ and $M_1$ and $M_2$ agree for every constant in $\dom{\flt{S_1}}\cap\dom{\flt{S_2}}$.
+In that case, we put
+\[\flt{(M\cup N)}=\flt{M}\cup\flt{N}\]
+\end{example}
+
+Finally, we define the semantics of the category structure:
+\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
+For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
+\[\flt{\id{T}}=\{c:=c \;|\; c\in\dom{\flt{T}}\}\]
+
+For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
+\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c\in\dom{\flt{R}}\}\]
+\end{example}
+
+%Includes between theories are always well-formed (assuming the included theory is).
+%But includes between morphisms are more difficult -- we have to impose two restrictions.
+%Firstly, an assignment $\icl{M}$ is well-formed in an atomic morphism $m:S\to T$ if $R\harr S$ and $M\in\Mo{R}{T}$.
+%Its intended effect is that $m$ maps all $c\in\dom{R}$ to $M(c)$.
+%That yields the second restriction: For any two morphisms $M_i\in\Mo{R_i}{T}$ included into $m$ and any theory $C$ satisfying $C\harr R_1$ and $C\harr R_2$, we require that $M_1|_C\equiv M_2|_C$.
+%Thus, any two included morphisms must agree on their common domain $C$; that is necessary so that $\flt{m}$ does not contain multiple assignments for the same identifier.
+
+\subsection{Semantic Relations}
+
+For two theories $S,T\in\Thy$, we say that $S$ is \textbf{included} into $T$, written $S\harr T$, if $\flt{S}\sq\flt{T}$.
 %Clearly, $m(o)=o$ for all $S$-objects $o$.
@@
 Correspondingly, we say that two morphisms $M,N\in\Mo{S}{T}$ are \textbf{equal}, written $M\equiv N$, if $\flt{M}=\flt{N}$.
-In that case, $\der_T M(o)\equiv N(o)$ for all well-formed $S$-objects $o\in\Exp{S}$.
+In that case, $M(e)=N(e)$ for all expressions $e\in\Exp{S}$.
 
 Clearly, $\equiv$ is an equivalence relation and $\harr$ is an order relation (if anti-symmetry is taken with respect to $\equiv$).
+
+\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{rel:incl}
+We have the following soundness theorem for includes: if a theory is declared as $t=\{\Sigma\}$ and $\Sigma$ contains $\icl{S}$, then $S\harr t$.
+
+Note that in the absence of include declarations or unions, $s\harr t$ never holds because all $c\in\dom{s}$ are of the form $s?n$ and all $c\in\dom{t}$ are of the form $t?n$.
+In fact, $\flt{s}$ and $\flt{t}$ unless $s=t$.
+\end{example}
+
+\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
+We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
+Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
+\end{example}
+
+\begin{example}[Category of Theories (continued from Ex.~\ref{sem:cat})]\label{rel:cat}
 Identity and composition are congruent with respect to $\equiv$.
 Moreover, composition is associative and identity neutral with respect to $\equiv$.
-Thus, each diagram yields the structure of a category.
-
-\paragraph{Preview}
-The above recap of (the non-modular fragment of) \mmt deviates from \cite{RK:mmt:10}, very carefully wording the definitions in a way that makes stating the results of this paper most elegant.
-Thus, we intentionally introduced some peculiarities:
-\begin{compactitem}
-\item We have $\Thy=\thy$ because there are no non-atomic theories.
-We will introduce some in Sect.~\ref{sec:complex}.
-\item The definitions of $\flt{t}$ and $\flt{m}$ are overly complicated.
-That will pay off when we introduce additional declarations in Sect.~\ref{sec:include}: We can define their semantics simply by adding a case to the inductions in $\flt{\Sigma}$ and $\flt{\sigma}$.
-\item $s\harr t$ never holds because all $c\in\dom{s}$ are of the form $s?n$ and all $c\in\dom{t}$ are of the form $t?n$.
-In fact, $\flt{S}$ and $\flt{T}$ are disjoint whenever $S\neq T$.
-That will change when we introduce include declarations in Sect.~\ref{sec:include}.
-\end{compactitem}
+Thus, each diagram indeed yields the structure of a category with objects $\Thy$ and arrows $\Mo{S}{T}$.
+\end{example}

### 2018-09-26 wadtabstract.tex
rewritten Sect. 1+2

@@
+\documentclass[orivec]{llncs}
+
+\usepackage{graphicx,url,wrapfig,amsmath,amssymb}
+
+\usepackage[show]{ed} %for ednotes
+
+\usepackage{sembracket}
+\usepackage{mytikz}
+\usepackage{basics}
+%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
+\usepackage{local}
+
+\setcounter{tocdepth}{3}
+\usepackage[bookmarks,linkcolor=red,citecolor=blue,urlcolor=gray,colorlinks,breaklinks,bookmarksopen,bookmarksnumbered]{hyperref}
+
+\setlength{\hfuzz}{3pt}
+\hbadness=10001 %make box warning less strict
+
+\begin{document}
+\title{Structuring Theories with Implicit Morphisms}
+\author{Florian Rabe$^{1,2}$ \and Dennis M\"uller$^2$}
+\institute{LRI Paris \and FAU Erlangen-Nuremberg}
+\maketitle
+
+\begin{abstract}
+We introduce \emph{implicit} morphisms as a concept in formal systems based on theories and theory morphisms.
+The idea is that there may be at most one implicit morphism from a theory $S$ to a theory $T$, and if $S$-expressions are used in $T$ their semantics is obtained by automatically inserting the implicit morphism.
+The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
+\end{abstract}
+
 Theory morphisms have proved an essential tool for managing collections of theories in logics and related formal systems.
@@
 Therefore, surprisingly, adding implicit morphisms deep in the \mmt kernel has made its design more elegant.
+
+\bibliographystyle{alpha}
+\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
+\end{document}

### 2018-09-26 mmt.tex

@@
+\ednote{@Dennis: can you formally work through the example from the intro in here?
+That would be several example environments for (i) after introducing the syntax of theories, the theory Group and DivGroup (one in detail, one sketched), (ii) after introducing morphisms the two isomorphisms (one in detail, one sketched), (iii) after introducing includes, the refactoring of Group using an include, (iv) examples for how the refactored Group is flattened}
+
 \subsection{Overview}

### 2018-09-26 appl.tex

--- 2018-06-12 implicit.tex
@@
-\subsection{Motivation}
-
-The motivation behind implicit morphisms is to generalize the benefits of include declarations even further.
-Sometimes an atomic morphism is conceptually similar to an inclusion: For example, we might want to declare $m:\cn{Division}\to\cn{Group}=\{\cn{divide} := \lambda x,y.x\circ y^{-1}, \;\ldots\}$ to show how division is ``included'' into the theory of groups.
-Other times we have to use structures because we need to duplicate declarations but we would still like to use them like includes.
-For example, we have to declare the theory of rings using two structures (for the commutative group and the monoid).
-But we still want a ring to be implicitly converted into, e.g., a commutative group.
-
-Our key idea is to use a commutative{\footnotemark} subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
-It contains all theories but only some of the morphisms -- the ones designated as \emph{implicit}.
-Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ -- if this morphism is $M$, we write $\ipc{M}{S}{T}$.
-\footnotetext{A diagram commutes if for any two morphisms $M,M'\in\Mo{S}{T}$, we have $M\equiv M'$.}
-%Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
-%We can extend the correspondence by identifying all paths with the composition of all edges along the path.
-%Then concatenation of paths is just composition of morphisms, and the empty path from $T$ to itself is the identity morphism $\id{T}$.
-%(This is well-defined because composition is associative and identity is neutral.)
-%A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.
-
-The implicit-diagram generalizes the inclusion relation $\harr$.
-In particular, all identity and inclusion morphisms are implicit, and we recover $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
-Moreover, the relation ``exists $M$ such that $\ipc{M}{S}{T}$'' is also an order.
-
-Consequently, many of the advantages of inclusions carry over to implicit morphisms.
-\begin{compactitem}
-\item It is very easy to maintain the implicit-diagram, e.g., as a partial map that assigns to a pair of theories the implicit morphism between them (if any).
-\item We can generalize the visibility of identifiers: If $\ipc{M}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
-Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
-\item We can still use canonical identifiers $s?n$. Because there can be at most one implicit morphism $\ipc{M}{s}{T}$, using $s?n$ as an identifier in $T$ is unambiguous.
-Crucially, we can use $s?n$ without bothering what $M$ is.
-\end{compactitem}
-
-%Thus the main question arises what other morphisms we should make implicit.
-
-\subsection{Definition}
-
-\paragraph{Syntax}
-We have already indicated that we will consider all identities and inclusions as implicit.
-It remains to give users the option introduce additional implicit morphisms.
-For that purpose, we allow two \mmt declarations to carry an attribute: 
-\begin{grammar}
-MDec   & Att\; m:T\to T=\{Ass,\ldots,Ass\}   & \text{attributed atomic morphism} \\
-Dec    & Att\; \icl{S}                       & \text{attributed inclusion}\\
-Att    & \keyword{implicit} \alt \keyword{invertible} & \text{attributes}
-\end{grammar}
-
-Every morphism with the attribute \keyword{implicit} occurs in the implicit-diagram.
-(Of course, this is redundant for includes.)
-Every morphism with the attribute \keyword{invertible} is also considered implicit, and additionally so is its inverse morphism.
-
-We do not even have to provide a name for the inverse morphism: It is uniquely determined (up to $\equiv$) anyway, and it will be used only implicitly.
-We only have to make sure that the inverse actually exists and that it can be found automatically.
-This is not decidable in general, but it is decidable in many important special cases, in particular, renamings and definitional extensions.
-
-\paragraph{Semantics}
-Well-formedness and semantics are defined in parallel.
-
-\begin{definition}[Implicit-Diagram]
-Given a diagram $D$, its implicit-diagram $D^i$ contains all theories and the following morphisms:
-\begin{compactitem}
-\item all identities and inclusions,
-\item all compositions of implicit morphisms,
-\item every morphism with the \keyword{implicit} attribute,
-\item for each morphism with the \keyword{invertible} attribute, that morphism and its inverse. 
-\end{compactitem}
-\end{definition}
-
-The only requirement for well-formedness is that the implicit-diagram always commutes.
-However, in general, commutativity may be undecidable.
-Similarly, the existence of an inverse morphism may be undecidable.
-Therefore, we use sound criteria for commutativity and invertibility as defined in Sect.~\ref{sec:commute} and \ref{sec:inverse}.
-
-\begin{definition}[Well-Formedness]
-Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
-Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
-$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
-$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.
-
-Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
-Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
-$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
-\end{definition}
-
-Finally, we modify the definition of $\Exp{T}$ and $M(-)$ to add all implicitly imported identifiers:
-
-\begin{definition}[Visible Declarations]
-The set $\flti{T}$ contains the following declarations:
-for every $\ipc{M}{S}{T}$ and every $c[:t][=d]$ in $\flt{S}$, the declaration $c[:M(t)]=M(c)$.
-
-The set $\Expi{T}$ contains all objects formed from the symbols in $\flti{T}$.
-The map $M(-)$ maps every $c$ for which $\flt{M}$ does not provide an assignment to $M(d)$ where $d$ is the definiens of $c$.
-\end{definition}
-
-This modification is conservative in the sense that $\flt{T}\sq\flti{T}$ and $\flti{T}\sm\flt{T}$ contains only declarations with definiens.
-Moreover, $\Exp{T}\sq\Expi{T}$ and $M(-)$ remains unchanged on all $o\in\Exp{T}$.
-It is easy to see that $M(-)$ still preserve all judgments.
-
-\subsection{Commutativity}\label{sec:commute}
-
-To prove commutativity, we have to find all paths in the implicit-diagram and check $M\equiv M'$ for all pairs of morphisms between the same nodes.
-
-\mmt reduces $M\equiv M'$ for $M,M'\in\Mo{S}{T}$ to a first-order problem in the theory of categories, which uses the following axioms:
-\begin{compactitem}
-\item associativity of composition,
-\item neutrality of identity,
-\item $M;M'\equiv \id{S}$ as well as $M';M\equiv\id{T}$ if $M\in\Mo{S}{T}$ and $M'$ is the known *inverse of $M$,
-\item if an atomic morphism $m$ includes $M\in\Mo{R}{T}$, then $M|_R\equiv M$.
-\end{compactitem}
-
-This is a sound but not complete axiomatization of $M\equiv M'$.
-We can obtain a complete axiomatization in a totally different way: Check $\der_T M(c)\equiv M'(c)$ for all $c\in\dom{S}$.
-However, this can be extremely expensive: It requires computing $\dom{S}$ and proving an object level equality for each identifier.
-Even if the equality of objects is decidable, this is often too expensive.
-
-In fact, because implicit morphisms are meant to be used in very particular cases, we expect the above incomplete calculus to perform reasonably well in practice.
-If showing the equality of certain morphisms is hard, one should probably not make them implicit in the first place.
-
-\subsection{Inverse Morphisms}\label{sec:inverse}
-
-We only two use sufficient criteria to determine an *inverse of a morphism $M$.
-
-\paragraph{Conservative Extensions}
-Intuitively, $t$ extends $S$ definitionally, if it only adds defined symbols, e.g., abbreviations or theorems.
-That is the most important case of conservative extensions.
-
-\begin{definition}[Definitional Extension]
-An include declaration \[t=\{\keyword{invertible}\;\icl{S},\Sigma\}\] is \textbf{definitional} if
-\begin{compactitem}
- \item all symbol declarations in $\Sigma$ have a definiens,
- \item all include declarations $\icl{R}$ in $\Sigma$ are such that $R$ also definitionally includes $S$.
-\end{compactitem}
-\end{definition}
-
-There are multiple, subtly different definitions of conservative extension.
-However, definitional extensions as defined above are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable formal language).
-In fact, they are always invertible: The inverse morphism $i:t\to S$ maps
-\begin{compactitem}
- \item an identifier $c\in\dom{S}$ to $c$ and everythemselves and the others to their definiens,
- \item any other identifier, which must have a definiens $d$, to $i(d)$.
-\end{compactitem}
-
-%We use two auxiliary definitions.
-%Given $u:T\to U$, we write $u|_S$ for the restriction of $u$ to $S$, i.e., the composition of the inclusion $S\harr T$ and $u$.
-%An extension $S\to T$ is retractable if there is a morphism $r:T\to S$ such that $r|_S=\id{S}$.
-%
-%Whether a given extension $S\harr T$ is conservative depends on the specific definition of conservativity and the used logic.
-%However, the most important kind of conservative extension is very easy to handle: the extension of a theory with a defined constant (e.g., an abbreviation or a theorem).
-%These extensions are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable logic).
-%In fact, they are always isomorphic to the original theory.
-%
-%Extensions with definition are almost exclusively used to add something to $S$.
-%Therefore, it is reasonable to automatically designate the retraction as implicit.
-%In particular, this provides an elegant way to add theorems to $S$ in such a way that they are available to any theory importing $S$, but without actually changing $S$.
-%
-%More generally, any retractable extension $S\harr T$ is always conservative.
-%But if the retraction usually does not exist uniquely (Often an arbitrary non-canonical choice is needed, e.g., any type.), this is not enough to make the retraction implicit.
-
-\paragraph{Renamings}
-Intuitively, an atomic morphism $m:S\to T$ is a renaming if it maps all $S$-identifiers to $T$-identifiers (rather than $T$-objects).
-
-\begin{definition}[Renaming]
-Consider a morphism $M\in\Mo{S}{T}$.
-Let $A\sq\dom{S}$ and $B\sq\dom{T}$ contain those identifiers without definiens.
-$M$ is an (injective, surjective) \textbf{renaming} if the restriction of $M(-)$ is an (injective, surjective) mapping $A\to B$.
-\end{definition}
-
-Clearly, if $m$ is an injective and surjective renaming, it is straightforward to obtain its inverse by inverting $M(-)$.
-
-\subsection{Applications}
-
 Canonical isomorphisms.

### 2018-09-26 implicit.tex

@@
-\subsection{Motivation}
+\subsection{Overview}
 
@@
 
-\subsection{Definition}
+\subsection{Syntax}
 
-\paragraph{Syntax}
-We have already indicated that we will consider all identities and inclusions as implicit.
-It remains to give users the option introduce additional implicit morphisms.
-For that purpose, we allow two \mmt declarations to carry an attribute: 
+We introduce the family of sets $\IMo{S}{T}$ as a subset of $\Mo{S}{T}$, holding the \textit{implicit} morphisms.
+The intuition is that $\Thy$ and $\IMo{S}{T}$ (up to equality of morphisms) form a thin broad subcategory of $\Thy$ and $\Mo{S}{T}$.
+
+It remains to define which morphisms are implicit.
+For that purpose, we allow \mmt declarations to carry \textit{attributes}.
+Precisely, we add the following productions
 \begin{grammar}
-MDec   & Att\; m:T\to T=\{Ass,\ldots,Ass\}   & \text{attributed atomic morphism} \\
-Dec    & Att\; \icl{S}                       & \text{attributed inclusion}\\
-Att    & \keyword{implicit} \alt \keyword{invertible} & \text{attributes}
+MDec   & Att\; MDec  & \text{attributed morphism} \\
+Dec    & Att\; Dec   & \text{attributed declaration}\\
+Att    & \keyword{implicit} \alt \ldots & \text{attributes}
 \end{grammar}
 
-Every morphism with the attribute \keyword{implicit} occurs in the implicit-diagram.
-(Of course, this is redundant for includes.)
-Every morphism with the attribute \keyword{invertible} is also considered implicit, and additionally so is its inverse morphism.
+The set of attributes is itself extensible, and the above grammar only list the one that we use to get started.
+Additional attributes may be added with the condition that any additional attribute increases the set of implicit morphisms.
 
-We do not even have to provide a name for the inverse morphism: It is uniquely determined (up to $\equiv$) anyway, and it will be used only implicitly.
-We only have to make sure that the inverse actually exists and that it can be found automatically.
-This is not decidable in general, but it is decidable in many important special cases, in particular, renamings and definitional extensions.
+Specifically, the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms contains the following elements:
+\begin{itemize}
+ \item all identity morphisms $\id{T}$,
+ \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
+ \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$.
+\end{itemize}
 
-\paragraph{Semantics}
-Well-formedness and semantics are defined in parallel.
+\subsection{Semantics}
 
-\begin{definition}[Implicit-Diagram]
-Given a diagram $D$, its implicit-diagram $D^i$ contains all theories and the following morphisms:
-\begin{compactitem}
-\item all identities and inclusions,
-\item all compositions of implicit morphisms,
-\item every morphism with the \keyword{implicit} attribute,
-\item for each morphism with the \keyword{invertible} attribute, that morphism and its inverse. 
-\end{compactitem}
-\end{definition}
+We only have to make two minor changes to the semantics to accommodate implicit morphisms.
+The first change govern how we obtain implicit morphisms, the second one how we use them.
 
-The only requirement for well-formedness is that the implicit-diagram always commutes.
-However, in general, commutativity may be undecidable.
-Similarly, the existence of an inverse morphism may be undecidable.
-Therefore, we use sound criteria for commutativity and invertibility as defined in Sect.~\ref{sec:commute} and \ref{sec:inverse}.
+\paragraph{Obtaining}
+We define well-formedness for our extended syntax.
+We define that an attribute to a declaration is only well-formed if the category of implicit morphisms remains thin, i.e., for any two theories, there may be only one (up to equality) implicit morphism between them.
+In other the diagram containing only the implicit morphisms must commute.
 
-\begin{definition}[Well-Formedness]
-Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
-Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
-$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
-$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.
+Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
+Therefore, in the general case, implementations may have to employ sound but incomplete criteria for commutativity (see Sect.~\ref{sec:commute}).
 
-Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
-Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
-$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
-\end{definition}
+%\begin{definition}[Well-Formedness]
+%Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
+%Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
+%$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
+%$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.
+%
+%Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
+%Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
+%$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
+%\end{definition}
 
-Finally, we modify the definition of $\Exp{T}$ and $M(-)$ to add all implicitly imported identifiers:
+\paragraph{Using}
+We make a small modification to the rules in
 
@@
 Clearly, if $m$ is an injective and surjective renaming, it is straightforward to obtain its inverse by inverting $M(-)$.
-
-\subsection{Applications}
-
-Canonical isomorphisms.
-
-User-declared atomic morphisms.
-
-\paragraph{Extending Theories from the Outside}
-
-\paragraph{Transparent Refactoring}
-A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
-This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
-That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.
-
-With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
-Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
-All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.

### 2018-09-26 mmt.tex

@@
 Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
-In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
+In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
 Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.
 
-Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
+Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T e:e'$ (typing) and $\der_T e\equiv e'$ (equality) for every theory $T$ and $e,e'\in\Exp{T}$.
 %This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
@@
 
-Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.
+The declarations in theories and morphisms are subject to typing conditions that are not essential for our purposes here.
 We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
@@
 \end{compactitem}
+
+\paragraph{Well-Formedness}
+The inference system to define well-formed modules, and expressions is not a primary interest of this paper.
+Therefore, we only mention those aspects that are relevant later on.
+
+However, one detail will be critical later on:
+the inference rules for the well-formedness of expressions include the following base case for constants:
+\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c:E} \tb \rul{c[:E]=e \minn \flt{T}}{\vdash_T c=e}\]
+Correspondingly, the definition of the homomorphic extension of a morphism includes the following base case for constants:
+\[M(c) = e \mwhere c:= e\in\flt{M}\]
+
+These base cases introduce a mutual recursion between well-formedness and flattening:
+the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
+and the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
+Vice versa, well-formedness is a precondition for defining the flattening.
+It is desirable to define well-formedness independently of flattening.
+But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems will first parse, check, and flatten the first declaration, then continue with the next declaration.
+In fact, often the checking of well-formedness is as expensive as flattening anyway.
+
+Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is defined iff the module $X$ is well-formed.
+
+%Consider any declaration $n[:E][=e]$ in the body of a theory $t$.
+%It is well-formed if $n$ is a fresh name $E$ and $e$ are well-formed expressions over $t$.
+%(Additionally, $E$ and $e$ must satisfy typing conditions that are not essential for our purposes.)
+%Importantly, $E$ and $e$ must be from $\Exp{\flt{t}}$ using only constants that were introduced in preceding declarations.
+%
+%Similarly, consider an assignment $c:=e$ in a morphism $m$ from $S$ to $T$, where $c:E$ is a declaration is $S$.
+%It is well-formed if $\vdash_T e:\flt{m}(E)$.
 
@@
  \[\fltd{\sigma}{c:=e}=\{c:=o\}\]
-\bigskip
-
-The semantics serves two purposes:
-firstly, it flattens every well-formed module into a flat one;
-secondly, it determines which modules are well-formed at all.
-It is desirable to define well-formedness independently of flattening.
-However, in practice, it shows that implementations typically have to do almost the entire flattening to check well-formedness anyway.
-
-Therefore, it is convenient to merge the two into one operation by simply making flattening a partial function: a module $X$ is \textbf{well-formed} iff $\flt{X}$ is defined.
-\bigskip
 
@@
 
-\subsection{Semantic Relations}
+\subsection{Semantic Properties}
 

### 2018-09-27 macros/listings.tex
update

\usepackage{listings}\lstset{basicstyle=\sf\footnotesize,columns=fullflexible}

\lstnewenvironment{code}[1][mathescape=false]{\lstset{language=XML, #1}}{}

\newcommand{\delima}{\mid\negmedspace \mid}
\newcommand{\delimb}{\delima\negmedspace \delima}
\newcommand{\GS}{\ensuremath{\textcolor{RedViolet}{\delimb\negmedspace\delimb}}}
\newcommand{\RS}{\ensuremath{\textcolor{Green}{\delimb}}}
\newcommand{\US}{\ensuremath{\textcolor{Turquoise}{\delima}}}

\lstdefinestyle{mmt}{	
	keywordstyle=\color{Emerald},
	columns=flexible
}
\lstdefinelanguage{mmt}{
	morekeywords={theory,view,include,namespace,rule,role,implicit,import},
	morecomment=[l][\color{Gray}]{//},
	% morecomment={[s]{//}{❚},[s]{//}{❙},[s]{//}{❘}}
}
\lstnewenvironment{mmtcode}[1][language=mmt]{\lstset{language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
\lstnewenvironment{nmmtcode}[1][language=mmt]{\lstset{xleftmargin=1.2em,numbers=left,language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
\lstdefinestyle{scala}{	
	keywordstyle=\color{BurntOrange}
}
\lstdefinelanguage{scala}{
	emphstyle=\underbar,
	classoffset=0,keywordstyle=\bfseries\color{BurntOrange},
	morekeywords={case,class,object,trait,extends,match,if,else,def,implicit,import,val,var,this,override,with,return,abstract},
	classoffset=1,keywordstyle=\color{blue},
	morekeywords={OMA,OMS,OMBIND,OMV,OMID,Context,VarDecl,Term,LocalName,GlobalName,MPath,Solver,Stack,Rule,
	History,Inhabitable,Subtyping,Equality,Typing,Common,Judgment,FormationRule,CheckingRule,IntroductionRule,
	EliminationRule,InferenceRule,TypingRule,ComputationRule,TypeBasedEqualityRule,TermBasedEqualityRule,
	TermHeadBasedEqualityRule,Continue,CheckingCallback,Simplifiability,Simplify,RecurseOnly,SubtypingRule,
	VarianceRule,TypeBasedSolutionRule,DerivedDeclaration,StructuralFeature,Elaboration,FormatBasedExtension,
	Marker,ExtendedCheckingEnvironment,DeclaredModule,Universe,InhabitableRule,UniverseRule,ElementContainer,Declaration},
	classoffset=2,keywordstyle=\color{PineGreen},
	morekeywords={Option,Some,None,Boolean,String,List,Int,true,false,Nil},
	classoffset=0,
	morecomment=[l][\color{Gray}]{//},
	morecomment=[s][\color{Gray}]{/*}{*/},
	morecomment=[s][\color{OliveGreen}]{/**}{**/},
	morestring=[b][\color{Purple}]",
}
\lstnewenvironment{scalacode}[1][language=scala]{\lstset{language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}
\lstnewenvironment{nscalacode}[1][language=scala]{\lstset{xleftmargin=1.2em,numbers=left,language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}

%\newunicodechar{→}{\ensuremath{\rightarrow}}
\lstset{
	literate={→}{\ensuremath{\rightarrow\ }}1
			{⟶}{\ensuremath{\rightarrow\ }}1
			{Π}{\ensuremath{\Pi}}1
			{⟹}{\ensuremath{\Longrightarrow\ }}1
			{λ}{\ensuremath{\lambda}}1
			{⟨}{\ensuremath{\langle}}1
			{⟩}{\ensuremath{\rangle}}1
			{❚}{\GS}1
			{❙}{\RS}1
			{❘}{\US}1
			{Σ}{\ensuremath{\sum}}1
			{…}{\ldots}1
			{×}{\ensuremath{\times}}1
			{π}{\ensuremath{\pi}}1
			{⨁}{\ensuremath{\bigoplus}}1
			{↪}{\ensuremath{\hookrightarrow}}1
			{↩}{\ensuremath{\hookleftarrow}}1
			{⊕}{\ensuremath{\oplus}}1
			{∅}{\ensuremath{\emptyset}}1
			{★}{\ensuremath{\star}}1
			{ℕ}{\ensuremath{\mathbb N}}1
			{∘}{\ensuremath{\circ}}1
			{∀}{\ensuremath{\forall}}1
			{≐}{\ensuremath{\doteq}}1
			{⊦}{\ensuremath{\vdash}}1
			{⁻}{\ensuremath{^-}}1
			{¹}{\ensuremath{^1}}1
}

### 2018-09-27 mmt.tex
update

@@
 
+\begin{example}\label{ex:thgroup}
+The (flat) theory for $\cn{Group}$ and $\cn{DivGroup}$ as presented in the introduction are:
+\begin{mmtcode}
+theory Group : ?Meta =
+  universe	: type		❘ # U ❙
+  op		: U ⟶ U ⟶ U	❘ # 1 ∘ 2 ❙
+  unit		: U		❘ # e❙
+  inverse : U ⟶ U ❘  # 1 ⁻¹ ❙
+
+  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z ❙
+  axiom_right_unit	: ⊦ ∀[x] x ∘ e ≐ x ❙
+  axiom_left_unit	: ⊦ ∀[x] e ∘ x ≐ x ❙
+  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ e ❙
+❚
+\end{mmtcode}
+\begin{mmtcode}
+theory DivisionGroup : ?Meta =
+  universe : type ❘ # U prec 1 ❙
+  div : U ⟶ U ⟶ U ❘ # 1 / 2 ❙
+  nonEmpty : U ❘ # e prec 1 ❙
+  // axioms ... ❙
+❚
+\end{mmtcode}
+where $\GS,\US,\RS$ are delimiters and the (optional) \lstinline|# <not>| is an optional notation. \lstinline|?Meta| is an intended meta-theory containing e.g. the symbols $\forall,\doteq$ etc.
+\end{example}
+
 At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
@@
 \end{definition}
+
+\begin{example}
+	We can now specify the morphisms between our two definitions of groups.
+The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
+\begin{mmtcode}
+view DG2G : ?DivisionGroup -> ?Group =
+  universe = universe ❙
+  div = [a,b] a ∘ (b⁻¹) ❙
+  nonEmpty = e ❙
+  // axioms ...❙
+❚
+\end{mmtcode}
+The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group.
+Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.
+
+The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
+\end{example}
 
@@
 \end{grammar}
+This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
+\begin{mmtcode}
+theory Monoid : ?Meta =
+  universe	: type		❘ # U ❙
+  op		: U ⟶ U ⟶ U	❘ # 1 ∘ 2 ❙
+  unit		: U		❘ # e❙
+
+  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z ❙
+  axiom_right_unit	: ⊦ ∀[x] x ∘ e ≐ x ❙
+  axiom_left_unit	: ⊦ ∀[x] e ∘ x ≐ x ❙
+❚
+
+theory Group : ?Meta =
+  include ?Monoid ❙
+
+  inverse : U ⟶ U ❘  # 1 ⁻¹ ❙
+  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ e ❙
+❚
+\end{mmtcode}
+The flat variant $\cn{Group}^\flat$ we get by making all included symbols (from $\cn{Monoid}$) available to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup}.
 \end{example}

### 2018-09-27 mmt.tex
update

@@
 \end{mmtcode}
-where $\GS,\US,\RS$ are delimiters and the (optional) \lstinline|# <not>| is an optional notation. \lstinline|?Meta| is an intended meta-theory containing e.g. the symbols $\forall,\doteq$ etc.
+where $\GS,\US,\RS$ are delimiters and the \lstinline|# <not>| components are (optional) notations. \lstinline|?Meta| is an intended meta-theory containing e.g. the symbols $\forall,\doteq$ etc.
 \end{example}
@@
 \end{mmtcode}
-The flat variant $\cn{Group}^\flat$ we get by making all included symbols (from $\cn{Monoid}$) available to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup}.
 \end{example}
@@
 under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
+
+The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers.
 \end{example}

### 2018-09-27 example.tex
Bands example

A good example for the usefulness of implicit morphisms is the lattice of \emph{bands}. A band is an idempotent semigroup. There are several varieties of bands which can all be defined by a single governing equation, resulting in the lattice in \autoref{fig:bands}.

\begin{figure}\begin{center}
\includegraphics[width=0.6\textwidth]{bands}\end{center}
\caption{The Lattice of Varieties of Bands}\label{fig:bands}
\end{figure}

As the diagram suggests, even though each node in the lattice is definable via a single equation, each of them implies the connected equations above it. Hence formalizing the varieties of bands via includes would add the implied equations as \emph{additional axioms}, even though they are provable.

Using implict morphisms between the varieties however can turn the axiom in a variety of bands into a theorem in another variety. \autoref{fig:bandsmmt} shows a subset of this lattice implemented with implict morphisms.

\begin{figure}
\begin{tabular}{c|c}
\begin{mmtcode}
theory Band : ?Meta =
  include ?SemiGroup ❙	
  axiom_idemp : ⊦ ∀[x] x ∘ x ≐ x ❙
❚

theory Regular : ?Meta =
  include ?Band ❙
  axiom_regular : ⊦ ∀[x]∀[y]∀[z] 
    z ∘ x ∘ z ∘ y ∘ z ≐ z ∘ x ∘ y ∘ z ❙
❚

theory LeftNormal : ?Meta =
  include ?Band ❙
  axiom_leftnormal : ⊦ ∀[x]∀[y]∀[z] 
    z ∘ x ∘ z ∘ y ≐ z ∘ x ∘ y ❙ 
❚

theory RightNormal : ?Meta =
  include ?Band ❙
  axiom_rightnormal : ⊦ ∀[x]∀[y]∀[z] 
    y ∘ z ∘ x ∘ z ≐ y ∘ x ∘ z ❙ 
❚

theory Normal : ?Meta =
  include ?Band ❙
  axiom_normal :  ⊦ ∀[x]∀[y]∀[z] 
    z ∘ x ∘ y ∘ z ≐ z ∘ y ∘ x ∘ z ❙
❚
\end{mmtcode} &
\begin{mmtcode}
implicit view Reg2LeftNormal : 
    ?Regular -> ?LeftNormal =
  include ?Band = ?Band ❙
  axiom_regular = sketch "trivial" ❙
❚

implicit view Reg2RightNormal : 
    ?Regular -> ?RightNormal =
  include ?Band = ?Band ❙
  axiom_regular = sketch "trivial" ❙
❚

implicit view LeftNormal2Normal : 
    ?LeftNormal -> ?Normal =
  include ?Band = ?Band ❙
  axiom_leftnormal = sketch "trivial" ❙
❚

implicit view RightNormal2Normal : 
    ?RightNormal -> ?Normal =
  include ?Band = ?Band ❙
  axiom_rightnormal = sketch "trivial" ❙
❚
\end{mmtcode}
\end{tabular}

\caption{Exemplary Theories for Some Varieties of Bands}\label{fig:bandsmmt}
\end{figure}

### 2018-09-27 implicit.tex
Bands example

@@
 
+\begin{example}\label{ex:dg2gimplicit}
+Going back to the morphism presented in \autoref{ex:dg2g}, we simply add the keyword \lstinline|implicit| in front, to declare it to be implicit:
+\begin{mmtcode}
+implicit view DG2G : ?DivisionGroup -> ?Group =
+  universe = universe ❙
+  div = [a,b] a ∘ (b⁻¹) ❙
+  nonEmpty = e ❙
+  // axioms ...❙
+❚
+\end{mmtcode}
+\end{example}
+
 \subsection{Semantics}
@@
 It is easy to see that $M(-)$ still preserve all judgments.
+
+\begin{example}
+	For our morphism $\cn{DG2G}$ from \autoref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see \autoref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
+\end{example}
 

### 2018-10-13 implicit.tex

@@
 The set of attributes is itself extensible, and the above grammar only list the one that we use to get started.
-Additional attributes may be added with the condition that any additional attribute increases the set of implicit morphisms.
-
-Specifically, the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms contains the following elements:
-\begin{itemize}
- \item all identity morphisms $\id{T}$,
- \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
- \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$.
-\end{itemize}
+Additional attributes can be added when adding modularity principles.
 
@@
 
-\paragraph{Obtaining}
-We define well-formedness for our extended syntax.
-We define that an attribute to a declaration is only well-formed if the category of implicit morphisms remains thin, i.e., for any two theories, there may be only one (up to equality) implicit morphism between them.
-In other the diagram containing only the implicit morphisms must commute.
+\paragraph{Obtaining Implicit Morphisms}
+Based on the above attributes, we define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
+\begin{itemize}
+ \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
+ \item all morphisms that additional modularity principles designate as implicit based on the use of additional attributes.
+\end{itemize}
 
+To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
 Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
-Therefore, in the general case, implementations may have to employ sound but incomplete criteria for commutativity (see Sect.~\ref{sec:commute}).
+Therefore, in the general case, implementations may have to employ sound but incomplete criteria for equality (see Sect.~\ref{sec:commute}).
 
-%\begin{definition}[Well-Formedness]
-%Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
-%Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
-%$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
-%$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.
-%
-%Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
-%Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
-%$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
-%\end{definition}
+The following example phrases the restriction on implicit morphisms in categorical terms:
 
-\paragraph{Using}
-We make a small modification to the rules in
+\begin{example}[Category of Theories and Implicit Morphisms (continued from Ex.~\ref{sem:cat})]\label{impl:cat}
+We define the following morphisms are implicit:
+\begin{compactitem}
+ \item all identity morphisms $\id{T}$,
+ \item all compositions $M;N$ of implicit morphisms $M$ and $N$.
+\end{compactitem}
 
-\begin{definition}[Visible Declarations]
-The set $\flti{T}$ contains the following declarations:
-for every $\ipc{M}{S}{T}$ and every $c[:t][=d]$ in $\flt{S}$, the declaration $c[:M(t)]=M(c)$.
+Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
+In other words, the implicit morphisms form commutative diagram.
+\end{example}
 
-The set $\Expi{T}$ contains all objects formed from the symbols in $\flti{T}$.
-The map $M(-)$ maps every $c$ for which $\flt{M}$ does not provide an assignment to $M(d)$ where $d$ is the definiens of $c$.
-\end{definition}
+\begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
+For include morphism, we add the following definition: if theory $t$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $t$ is implicit.
 
-This modification is conservative in the sense that $\flt{T}\sq\flti{T}$ and $\flti{T}\sm\flt{T}$ contains only declarations with definiens.
-Moreover, $\Exp{T}\sq\Expi{T}$ and $M(-)$ remains unchanged on all $o\in\Exp{T}$.
-It is easy to see that $M(-)$ still preserve all judgments.
+Combined with composition morphisms, we see that all transitive includes between theories are implicit.
+That corresponds to the intuition that anything that is include should be available directly.
+\end{example}
 
@@
 
-\subsection{Commutativity}\label{sec:commute}
+It is plausible to say that there are implicit morphisms $S_i\harr S_1\cup S_2$ for $i=1,2$ that map all constants to themselves.
+However, union expressions do not interact very well with implicit morphisms.
+First of all, we have to be careful to maintain the uniqueness condition.
+For example, assume there is a implicit morphism $M:S\to T$ that is not simply an inclusion.
+Then the union $S\cup T$ cannot be well-formed.
+If it were, we would have two different implicit morphisms $S\to S\cup T$, namely $M$ and the inclusion into the union.
 
-To prove commutativity, we have to find all paths in the implicit-diagram and check $M\equiv M'$ for all pairs of morphisms between the same nodes.
+Secondly, it becomes difficult to lookup whether an implicit morphism exists between union theories.
+We need to consider the equational theory of union (including commutativity, idempotence, and associativity), which is doable for unions but may become undecidable if more module expression formation operators are added.
+Already the interaction between union and composition morphisms can be tricky.
 
-\mmt reduces $M\equiv M'$ for $M,M'\in\Mo{S}{T}$ to a first-order problem in the theory of categories, which uses the following axioms:
+Therefore, we use the following definition in our system:
+
+\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{impl:union}
+We assume the definitions from Ex.~\ref{impl:cat} and~\ref{impl:incl} are in place.
+We want to add union theories.
+First of all, we note that every theory can be normalized into a union of named theories.
+
+Now we make the following restriction on implicit morphisms: we maintain a diagram of implicit morphisms in which all nodes are \emph{named} theories, i.e., we do not allow implicit morphisms between union theories.
+Because there are only finitely many named theories, this is a finite data structure.
+
+Consequently, we have to forbid morphism declarations $\keyword{implicit}\,m:S\to T_1\cup T_2$ is a union --- there would be no way to record them in our diagram.
+We can allow declarations declarations $\keyword{implicit}\,m:s_1\cup\ldots\cup s_n\to t$ where the domain is a union of named theories --- in that case we simply record $m$ as an implicit morphism from $s_i$ to $t$ for all $i$.
+Similarly, any declaration $\icl{s_1\cup\ldots\cup s_n}$ in the body of theory $t$ poses no problems --- we record inclusion morphism $s_1\harr t$.
+
+Now to look up the implicit morphism (if any) from $s_1\cup\ldots\cup s_k\to t_1\cup\ldots\cup t_l$, we look up for every $i$ an implicit morphism $m_i:s_i\to t_j$ for some $j$.
+If all these $m_i$ exist and $m=m_1\cup \ldots\cup m_k$ is well-formed, then $m$ is the needed implicit morphism.
+\end{example}
+
+\paragraph{Using Implicit Morphisms}
+The intuition behind implicit morphisms is that all $S$-constants $c:E$ that can be mapped into the current theory via an implicit morphism $M:S\to T$ are directly available in $T$.
+We can practically realize this by adding new defined constants $c:M(E)=M(c)$ to $T$.
+However, physically adding definitions can be inefficient.
+It is more elegant to modify the typing rules such that $\vdash_T c:M(c)=M(E)$.
+
+To do that, we only have to make a small modification to the original rules of \mmt as presented in Sect.~\ref{sec:mmt}.
+To illustrate how simple the modification is, we repeat the original rule first for comparison:
+\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
+Now our modified rule is
+\[\rul{c:E[=e] \minn \flt{S}\tb \ipc{M}{S}{T}}{\vdash_T c=M(c):M(E)}\]
+
+Note that the modified rule gives every constant a definiens.
+This is a technical trick to subsume the original rule: if $c$ is already declared in $T$, we use $M=\id{T}$ and obtain $\vdash_T c=c:E$.
+We write $\ov{E}$ for the expression that arises from $E$ by recursively replacing every $c$ with its definiens.
+
+The following theorem is our central theoretical result.
+It shows that the above modification has the intended properties:
+\begin{theorem}[Conservativity of Implicit Morphisms]
+Consider any combination of modular features from the examples above that includes at least implicit identity morphisms.
+
+Then for well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$
+\begin{compactenum}
+ \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
+ \item Whenever the modified system proves $\vdash_T e=e':E$, then the original inference system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
+\end{compactenum}
+\end{theorem}
+
+\begin{proof}
+For the first claim, we proceed by induction on derivations.
+We only need to consider case where the original rule was applied.
+So assume it yields $\vdash_T c[=e]:E$, i.e., $(c:E[=e]) \minn \flt{T}$.
+We apply the modified rule for the special case $\ipc{\id{T}}{T}{T}$.
+The conclusion reduces to
 \begin{compactitem}
-\item associativity of composition,
-\item neutrality of identity,
-\item $M;M'\equiv \id{S}$ as well as $M';M\equiv\id{T}$ if $M\in\Mo{S}{T}$ and $M'$ is the known *inverse of $M$,
-\item if an atomic morphism $m$ includes $M\in\Mo{R}{T}$, then $M|_R\equiv M$.
+ \item if $e$ is absent: $\vdash_T c=c:E$, which is equivalent to $\vdash_T c:E$,
+ \item if $e$ is present: $\vdash_T c=e:E$ because $M(c)=M(e)$ according to the definition of $M(-)$.
 \end{compactitem}
 
-This is a sound but not complete axiomatization of $M\equiv M'$.
-We can obtain a complete axiomatization in a totally different way: Check $\der_T M(c)\equiv M'(c)$ for all $c\in\dom{S}$.
-However, this can be extremely expensive: It requires computing $\dom{S}$ and proving an object level equality for each identifier.
-Even if the equality of objects is decidable, this is often too expensive.
+For the second claim, we proceed by induction on derivations.
+We only need to consider the cases where the modified rule was applied.
+So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
+We distinguish two cases:
+\begin{compactitem}
+ \item $M(c)=c$, i.e., $c$ does not have a definiens:
+   According to the definition of $M(-)$ this is only possible if $e$ is absent.
+   According to the definition of $\flt{T}$, this is only possible if $c=S?n$ for $S\harr T$ (including the special case $S=T$).
+   In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
+   Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
+ \item $M(c)\neq c$, i.e., $c$ has a non-trivial definiens:
+  Definition expansion eliminates $c$ in favor of $e$.
+  Clearly $\ov{c}=\ov{M(c)}$, and we only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
+\end{compactitem}
+\end{proof}
 
-In fact, because implicit morphisms are meant to be used in very particular cases, we expect the above incomplete calculus to perform reasonably well in practice.
-If showing the equality of certain morphisms is hard, one should probably not make them implicit in the first place.
+\section{Identifying Theories via Implicit Isomorphisms}\label{sec:inverse}
 
-\subsection{Inverse Morphisms}\label{sec:inverse}
+In this section, we introduce several language extensions that introduce implicit isomorphisms.
 
-We only two use sufficient criteria to determine an *inverse of a morphism $M$.
+Note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
+Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.
 
-\paragraph{Conservative Extensions}
-Intuitively, $t$ extends $S$ definitionally, if it only adds defined symbols, e.g., abbreviations or theorems.
-That is the most important case of conservative extensions.
+\paragraph{Renamings}
+We say that a named morphism $r:S\to T=\{\ldots\}$ is a \textbf{renaming} if
+\begin{compactitem}
+ \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
+ \item every such $T$-constants $c'$ occurs in exactly one assignment.
+\end{compactitem}
+Clearly, every renaming is an isomorphism.
+The inverse morphisms contains the flipped assignments $c':=c$.
 
-\begin{definition}[Definitional Extension]
-An include declaration \[t=\{\keyword{invertible}\;\icl{S},\Sigma\}\] is \textbf{definitional} if
+We make the following extension to syntax and semantics:
 \begin{compactitem}
- \item all symbol declarations in $\Sigma$ have a definiens,
- \item all include declarations $\icl{R}$ in $\Sigma$ are such that $R$ also definitionally includes $S$.
-\end{compactitem}
-\end{definition}
-
-There are multiple, subtly different definitions of conservative extension.
-However, definitional extensions as defined above are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable formal language).
-In fact, they are always invertible: The inverse morphism $i:t\to S$ maps
-\begin{compactitem}
- \item an identifier $c\in\dom{S}$ to $c$ and everythemselves and the others to their definiens,
- \item any other identifier, which must have a definiens $d$, to $i(d)$.
+  \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
+  \item This is well-formed if there are no implicit morphism between $s$ and $t$ yet.
+  \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
 \end{compactitem}
 
-%We use two auxiliary definitions.
-%Given $u:T\to U$, we write $u|_S$ for the restriction of $u$ to $S$, i.e., the composition of the inclusion $S\harr T$ and $u$.
-%An extension $S\to T$ is retractable if there is a morphism $r:T\to S$ such that $r|_S=\id{S}$.
+\ednote{@DM: add example, e.g., a definition of Monoid that uses different names}
+
+\paragraph{Definitional Extensions}
+We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
+\begin{compactitem}
+ \item only constant declarations with definiens,
+ \item only include declarations of theories that are definitional extensions of $S$.
+\end{compactitem}
+\ednote{@DM: give example that extends Group with the theorem $(x^{-1})^{-1}\doteq x$ (with omitted proof)}
+
+If $t$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
+In particular, the isomorphism $S\to t$ maps $S$-constants to themselves and expands the definiens of all other constants.
+
+We make the following extension to syntax and semantics:
+\begin{compactitem}
+  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
+  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
+\end{compactitem}
+
+\ednote{@DM: repeat previous example now with \keyword{definitional} and give another definitional extensions of Group; this new extension may already use the implicitly available definition of the first extension}
+
+\begin{remark}[Conservative Extensions]
+A definitional extension is a special case of a conservative extension.
+More generally, all retractable extensions are conservative, i.e., all extensions $S\harr T$ such that there is a morphism $r:T\to S$ such that $r$ is the identity on $S$.
+
+But we cannot make the retractions implicit morphisms in general because they are not necessarily isomorphisms.
+\end{remark}
+
+\paragraph{Canonical Isomorphisms}
+If we have isomorphisms $m:s\to t$ and $n:t\to s$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
+This requires no language extensions.
+
+\ednote{@DM: add the isomorphism of DG2G}
+\begin{example}\label{group:iso}
+We give the morphism $\cn{G2DG}$.
+\end{example}
+
+While the first one of these declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphism.
+Otherwise, the uniqueness condition would be violated.
+Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
+In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
+Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.
+
+However, this does not work in practice.
+Already elementary examples require stronger, undecidable equality relations are used:
+
+\begin{example}
+Consider the isomorphism from Ex.~\ref{group:iso}.
+The result of mapping $x\circ y$ from $\cn{Group}$ to $\cn{DivGroup}$ and back is $x\circ(unit\circ y^{-1})^{-1}$.
+Clearly, the group axioms imply that this is equal to $x\circ y$.
+But formally that requires working with the undecidable equality of first-order logic.
+\end{example}
+
+Therefore, in our running example, we only make one of the two isomorphisms implicit.
+
+In the sequel, we design a general solution to this problem.
+It allows systematically proving the equality of two morphisms and using that to make both isomorphisms implicit.
+This is novel work, but it requires significant prerequisites and is only peripherally related to implicit morphisms.
+Therefore, we only sketch the idea and leave the details to future work.
+
+We add a language feature to \mmt to prove equalities between morphisms:
+We add the productions
+\begin{grammar}
+Dia   & \rep{(TDec \alt MDec\alt MEq)}  &\\
+MEq   & \keyword{equal} M=M:T\to T by \{\rep{Ass}\} &\\
+\end{grammar}
+
+%Firstly, the intuition of $\keyword{equality} T = \{c:=E,\ldots\}$ is that it provides to every base type $c:\type$ of $T$ a judgment $E:c\to c\to \type$ that defines the $T$-specific equality relation for objects of type $c$.
+%Technically, this must be a binary logical relation in the sense of \cite{RS:logrels:12} on $\id{T}$.
+%As described in \cite{RS:logrels:12}, this induces an equality predicate $\cn{Equal}_E:E\to E\to \type$ on every type $E$.
+
+We define the declaration $M=N:S\to T by \{\sigma\}$ to be well-formed iff
+\begin{compactitem} 
+  \item $M:S\to T$ and $N:S\to T$ are well-formed morphisms
+  \item $\sigma$ contains exactly one assignment $c:=p$ for every $(c:E)\in\flt{S}$
+  \item for each of these assignments $c:=p$, the term $p$ is a proof of $\vdash_T M(c)=N(c)$.
+\end{compactitem}
+
+To make $m$ and $n$ from above implicit isomorphisms, we have to do three things: define $m$ and $n$, prove the equalities of $m;n=\id{s}$ and $n;m=\id{t}$, and make $m$ and $n$ implicit.
+Note that we cannot make both $m$ and $n$ implicit right away because that is only well-formed after proving the equalities.)
+Thus, we define a new attribute \keyword{implicit-later}, which states that a morphism should be considered implicit as soon as subsequent equality proves make it well-formed to do so.
+
+\begin{example}[Isomorphisms]
+We can now add declarations
+ \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
+ \[\keyword{implicit-later}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
+ \[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
+ \[\keyword{equal}\cn{DG2D};\cn{G2DG}=\id{\cn{DivGroup}}:\cn{DivGroup}\to\cn{DivGroup}=\text{(omitted)}\]
+where the isomorphisms are as above and we omit all the equality proofs.
+\end{example}
+
+%\subsection{Commutativity}\label{sec:commute}
 %
-%Whether a given extension $S\harr T$ is conservative depends on the specific definition of conservativity and the used logic.
-%However, the most important kind of conservative extension is very easy to handle: the extension of a theory with a defined constant (e.g., an abbreviation or a theorem).
-%These extensions are always conservative (i.e., for any reasonable definition of conservativity and in any reasonable logic).
-%In fact, they are always isomorphic to the original theory.
+%To prove commutativity, we have to find all paths in the implicit-diagram and check $M\equiv M'$ for all pairs of morphisms between the same nodes.
 %
-%Extensions with definition are almost exclusively used to add something to $S$.
-%Therefore, it is reasonable to automatically designate the retraction as implicit.
-%In particular, this provides an elegant way to add theorems to $S$ in such a way that they are available to any theory importing $S$, but without actually changing $S$.
+%\mmt reduces $M\equiv M'$ for $M,M'\in\Mo{S}{T}$ to a first-order problem in the theory of categories, which uses the following axioms:
+%\begin{compactitem}
+%\item associativity of composition,
+%\item neutrality of identity,
+%\item $M;M'\equiv \id{S}$ as well as $M';M\equiv\id{T}$ if $M\in\Mo{S}{T}$ and $M'$ is the known *inverse of $M$,
+%\item if an atomic morphism $m$ includes $M\in\Mo{R}{T}$, then $M|_R\equiv M$.
+%\end{compactitem}
 %
-%More generally, any retractable extension $S\harr T$ is always conservative.
-%But if the retraction usually does not exist uniquely (Often an arbitrary non-canonical choice is needed, e.g., any type.), this is not enough to make the retraction implicit.
-
-\paragraph{Renamings}
-Intuitively, an atomic morphism $m:S\to T$ is a renaming if it maps all $S$-identifiers to $T$-identifiers (rather than $T$-objects).
-
-\begin{definition}[Renaming]
-Consider a morphism $M\in\Mo{S}{T}$.
-Let $A\sq\dom{S}$ and $B\sq\dom{T}$ contain those identifiers without definiens.
-$M$ is an (injective, surjective) \textbf{renaming} if the restriction of $M(-)$ is an (injective, surjective) mapping $A\to B$.
-\end{definition}
-
-Clearly, if $m$ is an injective and surjective renaming, it is straightforward to obtain its inverse by inverting $M(-)$.
+%This is a sound but not complete axiomatization of $M\equiv M'$.
+%We can obtain a complete axiomatization in a totally different way: Check $\der_T M(c)\equiv M'(c)$ for all $c\in\dom{S}$.
+%However, this can be extremely expensive: It requires computing $\dom{S}$ and proving an object level equality for each identifier.
+%Even if the equality of objects is decidable, this is often too expensive.
+%
+%In fact, because implicit morphisms are meant to be used in very particular cases, we expect the above incomplete calculus to perform reasonably well in practice.
+%If showing the equality of certain morphisms is hard, one should probably not make them implicit in the first place.

### 2018-10-13 intro.tex

@@
 More precisely, out contribution is twofold.
+\medskip
 
@@
 As an example, we give includes and unions of theories, which induce the most important special cases of implicit morphisms.
+\medskip
 
-Secondly, we describe implicit morphisms as a novel feature of \mmt and give a number of examples.
+Secondly, we describe implicit morphisms as a novel feature of \mmt.
+Due to our novel presentation of the syntax and semantics, only minor modifications are needed to define the syntax and semantics of implicit morphisms.
+
+A very common situation in developing large libraries (both in formal and informal developments) is to identity two theories $S$ and $T$ via a canonical choice of isomorphisms.
+In these cases, it is often desirable to identify $S$ and $T$, e.g., when working with $S$, we want to be able to use $T$-syntax directly without having to apply an isomorphism.
+One of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
+We show that implicit morphisms elegantly allow exactly that kind of intuitive identification.
+As concrete examples, we consider manual isomorphisms, definitional extensions, and renamings.
 In particular, we recover the concept of realms \cite{realms} as a special case of implicit morphisms.

### 2018-10-13 mmt.tex
Bands example

--- 2018-09-26 mmt.tex
@@
 
-\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
-We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
+\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expression, and the latter is optional.
+We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$ (see below for the definition of well-formed expressions).
 Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
@@
 
-\paragraph{Well-Formedness}
-The inference system to define well-formed modules, and expressions is not a primary interest of this paper.
-Therefore, we only mention those aspects that are relevant later on.
+\paragraph{Well-Formed Expressions}
+The inference system to define well-formed expressions is not a primary interest of this paper.
+Intuitively, the set $\Exp{T}$ of expressions over $T$ includes all syntax trees that can be formed using the constants from $\flt{T}$.
+Moreover, well-formed modules may only use expressions that satisfy certain typing and equality judgments.
+We refer to \cite{rabe:howto:14} for details.
 
-However, one detail will be critical later on:
-the inference rules for the well-formedness of expressions include the following base case for constants:
-\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c:E} \tb \rul{c[:E]=e \minn \flt{T}}{\vdash_T c=e}\]
-Correspondingly, the definition of the homomorphic extension of a morphism includes the following base case for constants:
-\[M(c) = e \mwhere c:= e\in\flt{M}\]
+The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
+\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
+Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
+Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
+\[M(c) = \cas{e\mifc (c:E) \in \flt{S}, (c:=e)\in\flt{M} \\ M(e) \mifc (c:E=e)\in \flt{S}}\]
+Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
+\footnote{The \mmt tool also $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition.
+In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$.
+This is important for efficiency but not essential for our purposes here.}
 
-These base cases introduce a mutual recursion between well-formedness and flattening:
+Note that these base cases introduce a mutual recursion between well-formedness and flattening:
 the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
-and the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
-Vice versa, well-formedness is a precondition for defining the flattening.
+correspondingly, the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
+Vice versa, well-formedness is a precondition for defining the flattening --- the definition of flattening may become nonsensical if applied to ill-formed modules.
+
 It is desirable to define well-formedness independently of flattening.
-But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems will first parse, check, and flatten the first declaration, then continue with the next declaration.
-In fact, often the checking of well-formedness is as expensive as flattening anyway.
+But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems first parse, check, and flatten a declaration entirely before moving on to the next declaration.
+Moreover, often the checking of well-formedness is as difficult or expensive as flattening anyway.
+In fact, inspecting practical systems with modular features (which often do not have a full rigorous formal specification) shows that our definition elegantly captures the essential commonalities between them.
 
-Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is defined iff the module $X$ is well-formed.
+Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is undefined if the module $X$ is not well-formed.
 
@@
 
+\paragraph{Logics in \mmt}
+\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}
+
 \subsection{Syntax}
@@
 TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
-Dec      & n\opt{:o}\opt{=o}     & \text{constant declaration}\\
+Dec      & n:o\opt{=o}           & \text{constant declaration}\\
 c        & t?n                   & \text{qualified constant identifiers} \\
@@
 \end{definition}
+
+\begin{example}\label{ex:thgroup}
+The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
+\begin{mmtcode}
+Group =
+  universe	: type	 # U
+  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
+  unit		: U
+  inverse : U ⟶ U   # 1 ⁻¹
+
+  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
+  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
+  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x 
+  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
+\end{mmtcode}
+\begin{mmtcode}
+DivisionGroup =
+  universe : type  # U prec 1 
+  div : U ⟶ U ⟶ U  # 1 / 2 
+  unit : U  # e prec 1 
+  // axioms ... 
+\end{mmtcode}
+Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.
+
+Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
+\end{example}
 
@@
 \end{definition}
+
+\begin{example}\label{ex:dg2g}
+	We can now specify the morphisms between our two definitions of groups.
+The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
+\begin{mmtcode}
+DG2G : DivisionGroup -> Group =
+  universe = universe 
+  div = [a,b] a ∘ (b⁻¹) 
+  unit = unit
+  // axioms ...
+\end{mmtcode}
+The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group.
+Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.
+
+The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
+\end{example}
 
@@
 \end{grammar}
+This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
+\begin{mmtcode}
+Monoid =
+  universe	: type		 # U 
+  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
+  unit		: U
+  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
+  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
+  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x
+
+Group =
+  include Monoid
+  inverse : U ⟶ U   # 1 ⁻¹ 
+  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
+\end{mmtcode}
 \end{example}
@@
 Their flattening is trivial:
- \[\fltd{\Sigma}{n[:E][=e]}=\{t?n[:E][=e]\}\]
+ \[\fltd{\Sigma}{n:E[=e]}=\{t?n:E[=e]\}\]
 where $t$ is the name of the containing theory.
@@
 with the trivial base case
- \[\fltd{\sigma}{c:=e}=\{c:=o\}\]
+ \[\fltd{\sigma}{c:=e}=\{c:=e\}\]
 
@@
 For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
-\[\flt{\id{T}}=\{c:=c \;|\; c\in\dom{\flt{T}}\}\]
+\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\dom{\flt{T}}\}\]
 
 For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
-\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c\in\dom{\flt{R}}\}\]
+\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\dom{\flt{R}}\}\]
+
+Note that in both cases, we only have to consider constants $c:E$ without definiens.
 \end{example}

### 2018-10-13 paper.tex
rewritten Sect. 1+2
update
Bands example

@@
 
-\usepackage{sembracket}
-\usepackage{mytikz}
-\usepackage{basics}
+\usepackage{macros/sembracket}
+\usepackage{macros/mytikz}
+\usepackage{macros/basics}
 %\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
 \usepackage{local}
+
+\input{macros/listings}
 
@@
 \hbadness=10001 %make box warning less strict
+
+\pagestyle{plain}
 
@@
 
-\input{wadtabstract}
-\bibliographystyle{alpha}
-\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
-\end{document}
-
 \section{Introduction}
@@
 
-\section{Theories and Theory Morphism}
+\section{Theories and Theory Morphism}\label{sec:mmt}
 \input{mmt}
 
-\section{Inclusion Morphisms}\label{sec:include}
-\input{include}
+%\section{Inclusion Morphisms}\label{sec:include}
+%\input{include}
 
@@
 
-\section{Complex Theories and Morphisms}\label{sec:complex}
-\input{complex}
+\section{Applications}
+\input{appl}
+
+\section{Extended Example}
+\input{example}
+
+%\section{Complex Theories and Morphisms}\label{sec:complex}
+%\input{complex}
 
@@
 \bibliographystyle{alpha}
-\bibliography{../../../Program_Data/Latex/bib/rabe,../../../Program_Data/Latex/bib/systems,../../../Program_Data/Latex/bib/institutions,../../../Program_Data/Latex/bib/pub_rabe,../../../Program_Data/Latex/bib/historical}
+\bibliography{../../../../frabe/Program_Data/Latex/bib/rabe,../../../../frabe/Program_Data/Latex/bib/systems,../../../../frabe/Program_Data/Latex/bib/institutions,../../../../frabe/Program_Data/Latex/bib/pub_rabe,../../../../frabe/Program_Data/Latex/bib/historical}
 

### 2018-10-13 implicit.tex
remove module expressions and unions

@@
 \begin{example}\label{ex:dg2gimplicit}
-Going back to the morphism presented in \autoref{ex:dg2g}, we simply add the keyword \lstinline|implicit| in front, to declare it to be implicit:
-\begin{mmtcode}
-implicit view DG2G : ?DivisionGroup -> ?Group =
-  universe = universe ❙
-  div = [a,b] a ∘ (b⁻¹) ❙
-  nonEmpty = e ❙
-  // axioms ...❙
-❚
-\end{mmtcode}
+We can now change the declaration of the morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2g} by adding the attribute \keyword{implicit}.
 \end{example}
@@
  \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
- \item all morphisms that additional modularity principles designate as implicit based on the use of additional attributes.
+\begin{nomodexp}
+ \item all identity morphisms $\id{T}$,
+ \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
+\end{nomodexp}
+ \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
 \end{itemize}
@@
 To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
+Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
+In other words, the implicit morphisms form commutative diagram.
+
+\begin{modexp}
 Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
 Therefore, in the general case, implementations may have to employ sound but incomplete criteria for equality (see Sect.~\ref{sec:commute}).
+\end{modexp}
 
+\begin{modexp}
 The following example phrases the restriction on implicit morphisms in categorical terms:
-
 \begin{example}[Category of Theories and Implicit Morphisms (continued from Ex.~\ref{sem:cat})]\label{impl:cat}
@@
 \end{compactitem}
-
 Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
@@
 \end{example}
+\end{modexp}
 
@@
 \begin{example}
-	For our morphism $\cn{DG2G}$ from \autoref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see \autoref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
+	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
 \end{example}
 
+\begin{union}
 It is plausible to say that there are implicit morphisms $S_i\harr S_1\cup S_2$ for $i=1,2$ that map all constants to themselves.
@@
 \end{example}
+\end{union}
 
@@
 We give the morphism $\cn{G2DG}$.
+
 \end{example}

### 2018-10-13 mmt.tex
remove module expressions and unions

@@
 
+\begin{modexp}
 \paragraph{Modularity in MMT}
@@
 Building large morphism is as important as building large theories: for example, we want to be able to build morphisms out of $\cn{CommutativeGroup}$ by combining compatible morphisms out of $\cn{Group}$ and $\cn{CommutativeMagma}$.
+\end{modexp}
 
@@
 
-An \mmt \textbf{diagram} consists of a set of named structured module declarations and induces sets of anonymous module expressions.
+An \mmt \textbf{diagram} consists of a set of named structured module declarations.
+\begin{nomodexp}
+For a given diagram, we write $\Thy$ for the set of theory names.
+And we write $\Mo{S}{T}$ for the set of morphisms defined by
+\begin{compactitem}
+ \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
+ \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
+ \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
+\end{compactitem}
+$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
+\end{nomodexp}
+\begin{modexp}
+Every diagram induces sets of anonymous module expressions.
 These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
@@
 \end{compactitem}
+\end{modexp}
 Then flattening assigns
@@
  \item to each $T\in\Thy$ the flat theory $\flt{T}$,
- \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
+ \item to each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
 \end{compactitem}
@@
 \begin{definition}[Theory]\label{def:theory}
-A \textbf{theory} $T$ is of the form
+The grammar for theories and expressions is of the form
 \begin{grammar}
-TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
-Dec      & n:o\opt{=o}           & \text{constant declaration}\\
+TDec     & T=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
+Dec      & n:E\opt{=E}           & \text{constant declaration}\\
 c        & t?n                   & \text{qualified constant identifiers} \\
-o        & c \alt \ldots         & \text{expressions built from constants} \\
-T        & t                     & \text{theory expressions}
+E        & c \alt \ldots         & \text{expressions built from constants} \\
 \end{grammar}
@@
 Group =
-  universe	: type	 # U
+  U	: type
   op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
@@
   inverse : U ⟶ U   # 1 ⁻¹ 
-
-  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
-  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
-  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x 
-  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
+  // axioms omitted
 \end{mmtcode}
@@
 DivisionGroup =
-  universe : type  # U prec 1 
+  U : type
   div : U ⟶ U ⟶ U  # 1 / 2 
-  unit : U  # e prec 1 
-  // axioms ... 
+  unit : U
+  // axioms omitted
 \end{mmtcode}
@@
 Ass    & c:=o                            & \text{assignment to symbol}\\
-M      & m                               & \text{morphism expressions}
+M      & m \alt \id{T} \alt M;M          & \text{morphism expressions}
 \end{grammar}
@@
 
-\begin{example}\label{ex:dg2g}
-	We can now specify the morphisms between our two definitions of groups.
-The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
+\begin{example}[Morphisms]\label{ex:dg2g}
+We give the morphism \cn{DG2G} between the theories from Ex.~\ref{syn:incl}:
 \begin{mmtcode}
 DG2G : DivisionGroup -> Group =
-  universe = universe 
-  div = [a,b] a ∘ (b⁻¹) 
-  unit = unit
-  // axioms ...
+  U := U
+  div := [a,b] a ∘ (b⁻¹) 
+  unit := unit
+  // assignments to axioms omitted
 \end{mmtcode}
-The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group.
-Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.
-
-The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
+It maps the universe and unit of a division group to the corresponding notions of a group.
+And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
+Additionally, the morphism must maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
 \end{example}
@@
 \end{grammar}
-This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
+This allows writing the theory $\cn{Group}$ from Ex.~\ref{ex:thgroup} by extending a theory of monoids.
+Again omitting all axioms, this looks as follows:
 \begin{mmtcode}
 Monoid =
-  universe	: type		 # U 
+  U: type
   op : U ⟶ U ⟶ U	 # 1 ∘ 2 
   unit	: U
-  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
-  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
-  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x
-
 Group =
@@
   inverse : U ⟶ U   # 1 ⁻¹ 
-  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
 \end{mmtcode}
@@
 
+\begin{union}
 \begin{example}[Union]\label{syn:union}
@@
 \end{example}
+\end{union}
 
+\begin{modexp}
 \begin{example}[Category of Theories]\label{syn:cat}
@@
 In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
-This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.
+This follows the principle of compositional semantics, e.g., $\flt{f(X,Y)}$ depends only on $\flt{X}$ and $\flt{Y}$.
 
@@
 Their flattening is trivial:
- \[\fltd{\Sigma}{n:E[=e]}=\{t?n:E[=e]\}\]
+ \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
 where $t$ is the name of the containing theory.
@@
 with the trivial base case
- \[\fltd{\sigma}{c:=e}=\{c:=e\}\]
+ \[\fltd{\sigma}{(c:=e)}=\{c:=e\}\]
+\begin{nomodexp}
+Moreover, for $T\in\Thy$ we define
+\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\flt{T}\}\]
+And for $M;N\in\Mo{R}{T}$, we define
+\[\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\flt{R}\}\]
+Note that in both cases, we only have to consider $R$-constants without definiens.
+\end{nomodexp}
 
@@
 We do this as follows
-  \[\fltd{\Sigma}{\icl{S}} = \flt{S}\]
+  \[\fltd{\Sigma}{(\icl{S})} = \flt{S}\]
 This has the effect of copying over all declarations from the included into the including theory.
@@
 Therefore, flattening and well-formedness must be distinguished:
-  \[\fltd{\sigma}{\icl{M}} = \flt{M}\]
+  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
 under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
@@
 
+\begin{union}
 \begin{example}[Union (continued from Ex.~\ref{syn:union})]\label{sem:union}
@@
 \end{example}
-
-Finally, we define the semantics of the category structure:
-\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
-For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
-\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\dom{\flt{T}}\}\]
-
-For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
-\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\dom{\flt{R}}\}\]
-
-Note that in both cases, we only have to consider constants $c:E$ without definiens.
-\end{example}
+\end{union}
 
@@
 
+\begin{modexp}
 \subsection{Semantic Properties}
@@
 
+\begin{union}
 \begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
@@
 \end{example}
+\end{modexp}

### 2018-10-13 mmt.tex
typos

--- 2018-10-13 mmt.tex
@@
 Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
-\footnote{The \mmt tool also $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition.
+\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition.
 In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$.
@@
 
-In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions only previously introduced constants.
+In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
 We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
@@
 It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
-But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding production for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
+But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
 This makes it easy to incrementally add new declaration kinds and operators to the language.
@@
 \end{grammar}
-A diagram is well-formed if each atomic theory/morphism declaration has a unique name is well-formed relative to the diagram preceding it.
+A diagram is well-formed if each atomic theory/morphism declaration has a unique name and is well-formed relative to the diagram preceding it.
 \end{definition}
@@
 In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
-This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.
+This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$.
 
@@
 
-The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers.
+The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e.
+the prefixes $t$ in $t?n$.
 \end{example}
@@
 We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
-Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
+Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and given $M_1\in\Mo(S_1,T)$ and $M2\in\Mo{S_2,T}$ we have the universal morphism $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ out of it.
 \end{example}

### 2018-10-13 mmt.tex

--- 2018-10-13 mmt.tex
@@
 In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
-This follows the principle of compositional semantics, e.g., $\flt{f(X,Y)}$ depends only on $\flt{X}$ and $\flt{Y}$.
+This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$.
 

### 2018-10-13 intro.tex
typos

@@
 
-More precisely, out contribution is twofold.
+More precisely, our contribution is twofold.
 \medskip
 
-Firstly, we give a novel presentation of the syntax and semantics \mmt.
+Firstly, we give a novel presentation of the syntax and semantics of \mmt.
 Contrary to the original presentation in \cite{RK:mmt:10}, which hard-coded a single structuring mechanism --- there called \textit{structures} --- our definitions allows for adding arbitrary independent structuring mechanisms.
@@
 Based on this extensible definition, we can introduce several additional structuring mechanisms.
-Many of these have already been implemented in the \mmt tool but were not formally defined  a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
-With our reformulation, this is now not only possible but very easy and elegant.
+Many of these have already been implemented in the \mmt tool but were not formally defined  -- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
+With our reformulation, this is now not only possible but very easy.
 As an example, we give includes and unions of theories, which induce the most important special cases of implicit morphisms.
@@
 
-A very common situation in developing large libraries (both in formal and informal developments) is to identity two theories $S$ and $T$ via a canonical choice of isomorphisms.
+A very common situation in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
 In these cases, it is often desirable to identify $S$ and $T$, e.g., when working with $S$, we want to be able to use $T$-syntax directly without having to apply an isomorphism.

### 2018-10-13 appl.tex
further cleanup, conclusion still missing

@@
-Canonical isomorphisms.
+\subsection{Identifying Theories via Implicit Isomorphisms}\label{sec:inverse}
 
-User-declared atomic morphisms.
+In this section, we introduce several language extensions that introduce implicit isomorphisms.
 
-\paragraph{Extending Theories from the Outside}
+Note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
+Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.
 
-\paragraph{Transparent Refactoring}
-A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
-This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
-That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.
+\paragraph{Renamings}
+We say that a named morphism $r:S\to T=\{\ldots\}$ is a \textbf{renaming} if
+\begin{compactitem}
+ \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
+ \item every such $T$-constants $c'$ occurs in exactly one assignment.
+\end{compactitem}
+Clearly, every renaming is an isomorphism.
+The inverse morphisms contains the flipped assignments $c':=c$.
 
-With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
-Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
-All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.
+We make the following extension to syntax and semantics:
+\begin{compactitem}
+  \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
+  \item This is well-formed if there are no implicit morphism between $s$ and $t$ yet.
+  \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
+\end{compactitem}
+
+\ednote{@DM: add example, e.g., a definition of Monoid that uses different names}
+
+\paragraph{Definitional Extensions}
+We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
+\begin{compactitem}
+ \item only constant declarations with definiens,
+ \item only include declarations of theories that are definitional extensions of $S$.
+\end{compactitem}
+\ednote{@DM: give example that extends Group with the theorem $(x^{-1})^{-1}\doteq x$ (with omitted proof)}
+
+If $t$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
+In particular, the isomorphism $S\to t$ maps $S$-constants to themselves and expands the definiens of all other constants.
+
+We make the following extension to syntax and semantics:
+\begin{compactitem}
+  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
+  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
+\end{compactitem}
+
+\ednote{@DM: repeat previous example now with \keyword{definitional} and give another definitional extensions of Group; this new extension may already use the implicitly available definition of the first extension}
+
+\begin{remark}[Conservative Extensions]
+A definitional extension is a special case of a conservative extension.
+More generally, all retractable extensions are conservative, i.e., all extensions $S\harr T$ such that there is a morphism $r:T\to S$ such that $r$ is the identity on $S$.
+
+But we cannot make the retractions implicit morphisms in general because they are not necessarily isomorphisms.
+\end{remark}
+
+\paragraph{Canonical Isomorphisms}
+If we have isomorphisms $m:s\to t$ and $n:t\to s$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
+This requires no language extensions.
+
+\ednote{@DM: add the isomorphism of DG2G}
+\begin{example}\label{group:iso}
+We give the morphism $\cn{G2DG}$.
+
+\end{example}
+
+While the first one of these declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphism.
+Otherwise, the uniqueness condition would be violated.
+Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
+In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
+Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.
+
+However, this does not work in practice.
+Already elementary examples require stronger, undecidable equality relations are used:
+
+\begin{example}
+Consider the isomorphism from Ex.~\ref{group:iso}.
+The result of mapping $x\circ y$ from $\cn{Group}$ to $\cn{DivGroup}$ and back is $x\circ(unit\circ y^{-1})^{-1}$.
+Clearly, the group axioms imply that this is equal to $x\circ y$.
+But formally that requires working with the undecidable equality of first-order logic.
+\end{example}
+
+Therefore, in our running example, we only make one of the two isomorphisms implicit.
+
+In the sequel, we design a general solution to this problem.
+It allows systematically proving the equality of two morphisms and using that to make both isomorphisms implicit.
+This is novel work, but it requires significant prerequisites and is only peripherally related to implicit morphisms.
+Therefore, we only sketch the idea and leave the details to future work.
+
+We add a language feature to \mmt to prove equalities between morphisms:
+We add the productions
+\begin{grammar}
+Dia   & \rep{(TDec \alt MDec\alt MEq)}  &\\
+MEq   & \keyword{equal} M=M:T\to T by \{\rep{Ass}\} &\\
+\end{grammar}
+
+%Firstly, the intuition of $\keyword{equality} T = \{c:=E,\ldots\}$ is that it provides to every base type $c:\type$ of $T$ a judgment $E:c\to c\to \type$ that defines the $T$-specific equality relation for objects of type $c$.
+%Technically, this must be a binary logical relation in the sense of \cite{RS:logrels:12} on $\id{T}$.
+%As described in \cite{RS:logrels:12}, this induces an equality predicate $\cn{Equal}_E:E\to E\to \type$ on every type $E$.
+
+We define the declaration $M=N:S\to T by \{\sigma\}$ to be well-formed iff
+\begin{compactitem} 
+  \item $M:S\to T$ and $N:S\to T$ are well-formed morphisms
+  \item $\sigma$ contains exactly one assignment $c:=p$ for every $(c:E)\in\flt{S}$
+  \item for each of these assignments $c:=p$, the term $p$ is a proof of $\vdash_T M(c)=N(c)$.
+\end{compactitem}
+
+To make $m$ and $n$ from above implicit isomorphisms, we have to do three things: define $m$ and $n$, prove the equalities of $m;n=\id{s}$ and $n;m=\id{t}$, and make $m$ and $n$ implicit.
+Note that we cannot make both $m$ and $n$ implicit right away because that is only well-formed after proving the equalities.)
+Thus, we define a new attribute \keyword{implicit-later}, which states that a morphism should be considered implicit as soon as subsequent equality proves make it well-formed to do so.
+
+\begin{example}[Isomorphisms]
+We can now add declarations
+ \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
+ \[\keyword{implicit-later}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
+ \[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
+ \[\keyword{equal}\cn{DG2D};\cn{G2DG}=\id{\cn{DivGroup}}:\cn{DivGroup}\to\cn{DivGroup}=\text{(omitted)}\]
+where the isomorphisms are as above and we omit all the equality proofs.
+\end{example}
+
+\subsection{Fine-Granular and Flexible Theory Hierarchies}
+\input{example}
+
+\subsection{Transparent Refactoring}
+
+A major drawback of using modular theories is that it can preclude transparent refactoring.
+For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $r$ to $s$.
+Thus, the change to should be straightforward as it does not change the semantics of $t$.
+
+However, this is not a local change.
+It also requires updating every qualified reference from $r?n$ to $s?n$.
+Such references can occur anywhere where $t$ is used.
+That may include theories that the person who does the refactoring does not know or does not have access to.
+Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
+
+With implicit morphisms, we can solve this problem by making only the following local changes:
+\begin{compactenum}
+  \item We rename $s$ to $s'$.
+  \item We delete the declaration $n:E$ from $s'$.
+  \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.
+  \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
+  \item We add an implicit morphism $s\to t$ that maps $s?n$ to $r'?n$.
+\end{compactenum}
+Now $t$ has the desired new structure.
+But, all old references to $s?n$ stay well-formed so that no global changes are needed.
+
+\ednote{@DM: maybe make a tikz for the before/after situations}

### 2018-10-13 conc.tex
further cleanup, conclusion still missing

--- 2018-09-26 appl.tex
@@
-Canonical isomorphisms.
+\subsection{Related Work}
 
-User-declared atomic morphisms.
+\ednote{finish}
 
-\paragraph{Extending Theories from the Outside}
+\cite{realms} defines a realm (essentially) as a certain cluster of theories:
+\begin{compactitem}
+\item a set of isomorphic theories $B_1,\ldots,$
+\item for each $B_i$, a set of retractable extensions $E^i_1,\ldots,$
+\item a theory $F$ including all of the above.
+\end{compactitem}
+The intuition is that the $B_i$ are alternative definitions of a theory; the $E^i_j$ add definitions and theorems; and $F$ aggregates everything into the outwardly visible interface (the face) of the realm.
 
-\paragraph{Transparent Refactoring}
-A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
-This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
-That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.
+A good example is the realm of topological spaces with the $B_i$ corresponding to the various possible definitions (via neighborhoods, open sets, closure operator, etc.).
 
-With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
-Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
-All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.
+Users of the realm include only the face and thus (i) gain access to all derived knowledge in the theory of topological spaces, and (ii) do not have to commit to one particular definition.
+Indeed, this corresponds more closely to the way how conventional mathematics sees the theory of topological spaces.
+
+\cite{realms} calls for an implementation of realms as a new primitive concept in addition to theories and morphisms.
+However, in the presence of implicit morphisms, we do not need it: All we have to do is to designate the isomorphisms and the retraction is implicit and include any one of the $B_i$ into $F$.
+
+\paragraph{Canonical Structures Coq}
+\cite{gonthier_packaging}
+
+\paragraph{Type Classes}
+unnamed implict conversions, uniqueness problem does not arise
+
+\paragraph{Implicit Conversions in Scala}
+The Scala programming language \cite{scala} has a clean implementation of implicit conversions between any two types.
+If we think of functions between classes as theory morphisms, this corresponds to implicit theory morphisms.
+
+Contrary to \mmt, a conversion is only available if explicitly imported into the current namespace.
+Moreover, Scala does not chain conversions, i.e, the composition of implicit morphisms is not implicit.
+Scala's limitations are intentional: Their goal is to prevent programmers from overusing implicit conversions because that can be very confusing to readers.
+
+These limitations are lifted in \mmt's conversions can only act between theories anyway.
+Thus, their potential to confuse is smaller.
+Moreover, the use of qualified identifiers in \mmt means that readers can always tell where an identifier is implicitly-included from, and the \mmt IDE can always show the implicit morphism along which it is imported.
+
+\subsection{Future Work}
+
+\ednote{finish}
+
+build big library, necessary to investigate scoping problem
+
+\paragraph{Scoping}
+drawback: global data structure
+
+problem in big libraries?
+
+We anticipate that this will result in an evolution of our solution that allows more localized control over which morphisms are implicit.
+But we defer this until the current implementation has been used to conduct very large case studies.

### 2018-10-13 example.tex
further cleanup, conclusion still missing

@@
-A good example for the usefulness of implicit morphisms is the lattice of \emph{bands}.
-A band is an idempotent semigroup.
-There are several varieties of bands which can all be defined by a single governing equation, resulting in the lattice in \autoref{fig:bands}.
+A common problem when definition modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
+For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
+Instead, we could include \cn{Monoid} into the the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
+This change is not possible in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents an insurmountable refactoring problem.
+So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
+But this yields a very deep and complex hierarchy that is hard to navigate for casual users.
+Moreover, it does not protect us from later on discovering yet another intermediate theory that should have been added.
+
+Implicit morphisms provide a simple solution to this problem because they behave effectively like inclusions but can be added later on.
+In the above example, we would
+\begin{compactitem}
+ \item define \cn{CancellationMonoid} with an include from \cn{Monoid},
+ \item keep \cn{Group} as it is, i.e., also with an include \cn{Monoid},
+ \item add an implicit morphism $\cn{CancellationMonoid}\to\cn{Group}$.
+\end{compactitem}
+
+In the sequel, we develop a similar, more complex example: we build a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
+\ednote{@DM: this is not quite the lattice from the Wikipedia page: it consists of the 3x3 square plus Bands at the top}
+All of these theories are of the form $t=\{\icl{\cn{Band}},\,a:\vdash F\}$, where $F$ is a universally quantified equational axiom.\footnote{All varieties of bands can be axiomatized in this way.}
+In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a:\vdash \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
+Fig.~\ref{fig:bands} shows these theories.
+
+There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
+All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\ednote{@DM: no need to give the morphisms, but link to the online resource if you have time to work them all out}
+We make all of these morphisms implicit.
+It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom.
+Here, \mmt can discharge the proof obligation easily using proof irrelevance.
 
@@
 
-As the diagram suggests, even though each node in the lattice is definable via a single equation, each of them implies the connected equations above it.
-Hence formalizing the varieties of bands via includes would add the implied equations as \emph{additional axioms}, even though they are provable.
-
-Using implict morphisms between the varieties however can turn the axiom in a variety of bands into a theorem in another variety.
-\autoref{fig:bandsmmt} shows a subset of this lattice implemented with implict morphisms.
-
-\begin{figure}
-\begin{tabular}{c|c}
-\begin{mmtcode}
-theory Band : ?Meta =
-  include ?SemiGroup ❙	
-  axiom_idemp : ⊦ ∀[x] x ∘ x ≐ x ❙
-❚
-
-theory Regular : ?Meta =
-  include ?Band ❙
-  axiom_regular : ⊦ ∀[x]∀[y]∀[z] 
-    z ∘ x ∘ z ∘ y ∘ z ≐ z ∘ x ∘ y ∘ z ❙
-❚
-
-theory LeftNormal : ?Meta =
-  include ?Band ❙
-  axiom_leftnormal : ⊦ ∀[x]∀[y]∀[z] 
-    z ∘ x ∘ z ∘ y ≐ z ∘ x ∘ y ❙ 
-❚
-
-theory RightNormal : ?Meta =
-  include ?Band ❙
-  axiom_rightnormal : ⊦ ∀[x]∀[y]∀[z] 
-    y ∘ z ∘ x ∘ z ≐ y ∘ x ∘ z ❙ 
-❚
-
-theory Normal : ?Meta =
-  include ?Band ❙
-  axiom_normal :  ⊦ ∀[x]∀[y]∀[z] 
-    z ∘ x ∘ y ∘ z ≐ z ∘ y ∘ x ∘ z ❙
-❚
-\end{mmtcode} &
-\begin{mmtcode}
-implicit view Reg2LeftNormal : 
-    ?Regular -> ?LeftNormal =
-  include ?Band = ?Band ❙
-  axiom_regular = sketch "trivial" ❙
-❚
-
-implicit view Reg2RightNormal : 
-    ?Regular -> ?RightNormal =
-  include ?Band = ?Band ❙
-  axiom_regular = sketch "trivial" ❙
-❚
-
-implicit view LeftNormal2Normal : 
-    ?LeftNormal -> ?Normal =
-  include ?Band = ?Band ❙
-  axiom_leftnormal = sketch "trivial" ❙
-❚
-
-implicit view RightNormal2Normal : 
-    ?RightNormal -> ?Normal =
-  include ?Band = ?Band ❙
-  axiom_rightnormal = sketch "trivial" ❙
-❚
-\end{mmtcode}
-\end{tabular}
-
-\caption{Exemplary Theories for Some Varieties of Bands}\label{fig:bandsmmt}
-\end{figure}
+%
+%\begin{figure}
+%\begin{tabular}{c|c}
+%\begin{mmtcode}
+%Band =
+%  include ?SemiGroup
+%  axiom_idemp : ⊦ ∀[x] x ∘ x ≐ x
+%❚
+%
+%Regular =
+%  include ?Band
+%  axiom_regular : ⊦ ∀[x]∀[y]∀[z] 
+%    z ∘ x ∘ z ∘ y ∘ z ≐ z ∘ x ∘ y ∘ z
+%❚
+%
+%LeftNormal =
+%  include ?Band
+%  axiom_leftnormal : ⊦ ∀[x]∀[y]∀[z] 
+%    z ∘ x ∘ z ∘ y ≐ z ∘ x ∘ y
+%
+%theory RightNormal =
+%  include ?Band ❙
+%  axiom_rightnormal : ⊦ ∀[x]∀[y]∀[z] 
+%    y ∘ z ∘ x ∘ z ≐ y ∘ x ∘ z ❙ 
+%❚
+%
+%theory Normal : ?Meta =
+%  include ?Band ❙
+%  axiom_normal :  ⊦ ∀[x]∀[y]∀[z] 
+%    z ∘ x ∘ y ∘ z ≐ z ∘ y ∘ x ∘ z ❙
+%❚
+%\end{mmtcode} &
+%\begin{mmtcode}
+%implicit view Reg2LeftNormal : 
+%    ?Regular -> ?LeftNormal =
+%  include ?Band = ?Band ❙
+%  axiom_regular = sketch "trivial" ❙
+%❚
+%
+%implicit view Reg2RightNormal : 
+%    ?Regular -> ?RightNormal =
+%  include ?Band = ?Band ❙
+%  axiom_regular = sketch "trivial" ❙
+%❚
+%
+%implicit view LeftNormal2Normal : 
+%    ?LeftNormal -> ?Normal =
+%  include ?Band = ?Band ❙
+%  axiom_leftnormal = sketch "trivial" ❙
+%❚
+%
+%implicit view RightNormal2Normal : 
+%    ?RightNormal -> ?Normal =
+%  include ?Band = ?Band ❙
+%  axiom_rightnormal = sketch "trivial" ❙
+%❚
+%\end{mmtcode}
+%\end{tabular}
+%
+%\caption{Exemplary Theories for Some Varieties of Bands}\label{fig:bandsmmt}
+%\end{figure}

### 2018-10-13 implicit.tex
typos
more typos
further cleanup, conclusion still missing

@@
 To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
-Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
-In other words, the implicit morphisms form commutative diagram.
+Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms --
+in other words, a commutative diagram.
 
@@
 
-\section{Identifying Theories via Implicit Isomorphisms}\label{sec:inverse}
-
-In this section, we introduce several language extensions that introduce implicit isomorphisms.
-
-Note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
-Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.
-
-\paragraph{Renamings}
-We say that a named morphism $r:S\to T=\{\ldots\}$ is a \textbf{renaming} if
-\begin{compactitem}
- \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
- \item every such $T$-constants $c'$ occurs in exactly one assignment.
-\end{compactitem}
-Clearly, every renaming is an isomorphism.
-The inverse morphisms contains the flipped assignments $c':=c$.
-
-We make the following extension to syntax and semantics:
-\begin{compactitem}
-  \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
-  \item This is well-formed if there are no implicit morphism between $s$ and $t$ yet.
-  \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
-\end{compactitem}
-
-\ednote{@DM: add example, e.g., a definition of Monoid that uses different names}
-
-\paragraph{Definitional Extensions}
-We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
-\begin{compactitem}
- \item only constant declarations with definiens,
- \item only include declarations of theories that are definitional extensions of $S$.
-\end{compactitem}
-\ednote{@DM: give example that extends Group with the theorem $(x^{-1})^{-1}\doteq x$ (with omitted proof)}
-
-If $t$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
-In particular, the isomorphism $S\to t$ maps $S$-constants to themselves and expands the definiens of all other constants.
-
-We make the following extension to syntax and semantics:
-\begin{compactitem}
-  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
-  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
-\end{compactitem}
-
-\ednote{@DM: repeat previous example now with \keyword{definitional} and give another definitional extensions of Group; this new extension may already use the implicitly available definition of the first extension}
-
-\begin{remark}[Conservative Extensions]
-A definitional extension is a special case of a conservative extension.
-More generally, all retractable extensions are conservative, i.e., all extensions $S\harr T$ such that there is a morphism $r:T\to S$ such that $r$ is the identity on $S$.
-
-But we cannot make the retractions implicit morphisms in general because they are not necessarily isomorphisms.
-\end{remark}
-
-\paragraph{Canonical Isomorphisms}
-If we have isomorphisms $m:s\to t$ and $n:t\to s$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
-This requires no language extensions.
-
-\ednote{@DM: add the isomorphism of DG2G}
-\begin{example}\label{group:iso}
-We give the morphism $\cn{G2DG}$.
-
-\end{example}
-
-While the first one of these declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphism.
-Otherwise, the uniqueness condition would be violated.
-Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
-In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
-Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.
-
-However, this does not work in practice.
-Already elementary examples require stronger, undecidable equality relations are used:
-
-\begin{example}
-Consider the isomorphism from Ex.~\ref{group:iso}.
-The result of mapping $x\circ y$ from $\cn{Group}$ to $\cn{DivGroup}$ and back is $x\circ(unit\circ y^{-1})^{-1}$.
-Clearly, the group axioms imply that this is equal to $x\circ y$.
-But formally that requires working with the undecidable equality of first-order logic.
-\end{example}
-
-Therefore, in our running example, we only make one of the two isomorphisms implicit.
-
-In the sequel, we design a general solution to this problem.
-It allows systematically proving the equality of two morphisms and using that to make both isomorphisms implicit.
-This is novel work, but it requires significant prerequisites and is only peripherally related to implicit morphisms.
-Therefore, we only sketch the idea and leave the details to future work.
-
-We add a language feature to \mmt to prove equalities between morphisms:
-We add the productions
-\begin{grammar}
-Dia   & \rep{(TDec \alt MDec\alt MEq)}  &\\
-MEq   & \keyword{equal} M=M:T\to T by \{\rep{Ass}\} &\\
-\end{grammar}
-
-%Firstly, the intuition of $\keyword{equality} T = \{c:=E,\ldots\}$ is that it provides to every base type $c:\type$ of $T$ a judgment $E:c\to c\to \type$ that defines the $T$-specific equality relation for objects of type $c$.
-%Technically, this must be a binary logical relation in the sense of \cite{RS:logrels:12} on $\id{T}$.
-%As described in \cite{RS:logrels:12}, this induces an equality predicate $\cn{Equal}_E:E\to E\to \type$ on every type $E$.
-
-We define the declaration $M=N:S\to T by \{\sigma\}$ to be well-formed iff
-\begin{compactitem} 
-  \item $M:S\to T$ and $N:S\to T$ are well-formed morphisms
-  \item $\sigma$ contains exactly one assignment $c:=p$ for every $(c:E)\in\flt{S}$
-  \item for each of these assignments $c:=p$, the term $p$ is a proof of $\vdash_T M(c)=N(c)$.
-\end{compactitem}
-
-To make $m$ and $n$ from above implicit isomorphisms, we have to do three things: define $m$ and $n$, prove the equalities of $m;n=\id{s}$ and $n;m=\id{t}$, and make $m$ and $n$ implicit.
-Note that we cannot make both $m$ and $n$ implicit right away because that is only well-formed after proving the equalities.)
-Thus, we define a new attribute \keyword{implicit-later}, which states that a morphism should be considered implicit as soon as subsequent equality proves make it well-formed to do so.
-
-\begin{example}[Isomorphisms]
-We can now add declarations
- \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
- \[\keyword{implicit-later}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
- \[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
- \[\keyword{equal}\cn{DG2D};\cn{G2DG}=\id{\cn{DivGroup}}:\cn{DivGroup}\to\cn{DivGroup}=\text{(omitted)}\]
-where the isomorphisms are as above and we omit all the equality proofs.
-\end{example}
-
 %\subsection{Commutativity}\label{sec:commute}

### 2018-10-13 intro.tex
further cleanup, conclusion still missing

@@
 Based on this extensible definition, we can introduce several additional structuring mechanisms.
-Many of these have already been implemented in the \mmt tool but were not formally defined  -- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
+Many of these have already been implemented in the \mmt tool but were not formally defined --- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
 With our reformulation, this is now not only possible but very easy.
-As an example, we give includes and unions of theories, which induce the most important special cases of implicit morphisms.
+As an example, we define include declarations, the most important special case of implicit morphisms.
 \medskip

### 2018-10-13 mmt.tex
further cleanup, conclusion still missing

@@
  \[\flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
-Here $\fltd{T}{D}$ is the flattening of declaration $d$ in the context of theory $T$.
+Here $\fltd{T}{D}$ is the flattening of declaration $d$ relative to $\flt{\Sigma}$.
 
@@
 
-The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from \autoref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e.
+The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from Ex.~\ref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e.
 the prefixes $t$ in $t?n$.

### 2018-10-14 appl.tex
update

@@
 
-\ednote{@DM: add example, e.g., a definition of Monoid that uses different names}
+\begin{example}
+	Consider a variation of the theory $\cn{Monoid}$ from \autoref{syn:incl} in a different library:
+	\begin{mmtcode}
+Monoid2 =
+  M: type
+  connective : M ⟶ M ⟶ M	 # 1 ∘ 2 
+  neutral	: M
+\end{mmtcode}
+This theory is isomorphic to the previously introduced theory $\cn{Monoid}$ under the trivial renaming
+	\begin{mmtcode}
+Monoids : Monoid2 -> Monoid =
+  M 		:= U
+  connective 	:= op 
+  neutral	:= unit
+\end{mmtcode}
+\end{example}
 
@@
 \end{compactitem}
-\ednote{@DM: give example that extends Group with the theorem $(x^{-1})^{-1}\doteq x$ (with omitted proof)}
+\begin{example}
+	Imagine extending the theory $\cn{Group}$ from \autoref{syn:incl} by provable theorems:
+		\begin{mmtcode}
+Group_Theorems =
+  include Group
+  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
+\end{mmtcode}
+\end{example}
 
@@
 \end{compactitem}
+\begin{example}
+	We adapt our theory $\cn{Group\_Theorems}$ with the newly introduced keyword, which allows us to extend $\cn{Group}$ directly while having the new theorem $\cn{inverse\_idem}$ available:
+		\begin{mmtcode}
+Group_Theorems =
+  definitional include Group
+  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
 
-\ednote{@DM: repeat previous example now with \keyword{definitional} and give another definitional extensions of Group; this new extension may already use the implicitly available definition of the first extension}
+More_Theorems =
+  include ?Group
+  some_other_theorem : ⊦ ??	= (proof using inverse_idem)
+\end{mmtcode}
+\end{example}
 
@@
 
-\ednote{@DM: add the isomorphism of DG2G}
 \begin{example}\label{group:iso}
-We give the morphism $\cn{G2DG}$.
-
+Having declared the morphism $\cn{DG2G}$ (as in \autoref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
+\begin{mmtcode}
+implicit G2DG : Group -> DivisionGroup =
+  U := U
+  op := [a,b] a / ((unit/unit) / b) 
+  unit := unit/unit
+  inverse := [a] (unit/unit) / a
+\end{mmtcode}
 \end{example}
 
-While the first one of these declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphism.
+While the first one of these isomorphism declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphisms.
 Otherwise, the uniqueness condition would be violated.

### 2018-10-14 example.tex
update

@@
-A common problem when definition modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
+A common problem when defining modular theory hierarchies is that the most natural include-hierarchy for the most important theories is not necessarily the same as the most comprehensive hierarchy.
 For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
-Instead, we could include \cn{Monoid} into the the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
+Instead, we could include \cn{Monoid} into the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
 This change is not possible in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents an insurmountable refactoring problem.
@@
 \begin{figure}\begin{center}
-\includegraphics[width=0.6\textwidth]{bands}\end{center}
+\begin{tikzpicture}[scale=1.5]
+  \node (s1) at (0,0) {$zxyz=zxzyz$};
+
+  \node (s2a) at (-1,-1) {$zxy=zxzy$};
+  \node (s2b) at (1,-1) {$yxz=yzxz$};
+  \draw (s1) to (s2a) {};
+  \draw (s1) to (s2b) {};
+
+  \node (s3a) at (-2,-2) {$xy=xyx$};
+  \node (s3b) at (0,-2) {$zxyz=zyxz$};
+  \node (s3c) at (2,-2) {$xy=yxy$};
+  \draw (s2a) to (s3a) {};
+  \draw (s2a) to (s3b) {};
+  \draw (s2b) to (s3b) {};
+  \draw (s2b) to (s3c) {};
+
+  \node (s4a) at (-1.5,-3) {$zxy=zyx$};
+  \node (s4b) at (0,-3) {$xyx=x$};
+  \node (s4c) at (1.5,-3) {$xyz=yxz$};
+  \draw (s3a) to (s4a) {};
+  \draw (s3b) to (s4a) {};
+  \draw (s3b) to (s4b) {};
+  \draw (s3b) to (s4c) {};
+  \draw (s3c) to (s4c) {};
+
+  \node (s5a) at (-1.5,-4) {$xy=x$};
+  \node (s5b) at (0,-4) {$xy=yx$};
+  \node (s5c) at (1.5,-4) {$xy=y$};
+  \draw (s4a) to (s5a) {};
+  \draw (s4a) to (s5b) {};
+  \draw (s4b) to (s5a) {};
+  \draw (s4b) to (s5c) {};
+  \draw (s4c) to (s5b) {};
+  \draw (s4c) to (s5c) {};
+
+  \node (s6) at (0,-5) {$x=y$};
+  \draw (s5a) to (s6) {};
+  \draw (s5b) to (s6) {};
+  \draw (s5c) to (s6) {};
+
+  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
+  %\node at (t1.north west) {$S$};
+  %\node at (t1) {$D$};
+  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
+  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
+  %\node at (t2.north west) {$T$};
+  %\node at (t2) {$C$};
+  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
+\end{tikzpicture}
+
+%\includegraphics[width=0.6\textwidth]{bands}
+\end{center}
 \caption{The Lattice of Varieties of Bands}\label{fig:bands}

### 2018-10-14 mmt.tex
update

@@
 \ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}
+\mmt itself is foundation independent and parmetric in a set of typing rules.
+We omit the details on implementing type systems and logical frameworks in \mmt, and will instead assume theories for the logical framework LF and an implementation of first-order logic for the examples in the remainder of this paper.
+
+Specifically, we assume a theory $\cn{LF}$ that implements the dependently-typed lambda calculus LF.
+This theory provides symbols for dependent function types $\prod_{c:A}B(x)$ (in surface syntax: \lstinline|{x:A}B x|), simple function types $A\to B$ (as abbreviation for $\prod_{\_:A}B$, lambda abstractions $\lambda x:A.\; t(x)$ (in surface syntax: \lstinline|[x:A] t x|) and function application $f a$.
+
+The theory \cn{LF} is used as a meta-theory for a theory \cn{FOL} implementing first-order logic with equality.
+This theory provides symbols for the quantifiers $\forall$ and $\exists$ (implemented via higher-order abstract syntax, i.e.
+the proposition $\forall x.\; P(x)$ is written as $\forall$\lstinline|[x:U]P x| for a fixed type $U$), equality $\doteq$ and the usual logical connectives $\neg,\wedge,\vee$, etc.
 
@@
 
-Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
+Moreover, we use the constants from the theories $\cn{LF}$ and $\cn{FOL}$ described above.
 \end{example}

### 2018-10-14 appl.tex
update

@@
 Now $t$ has the desired new structure.
-But, all old references to $s?n$ stay well-formed so that no global changes are needed.
+But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in \autoref{fig:refac}.
 
-\ednote{@DM: maybe make a tikz for the before/after situations}
+\begin{figure}\begin{center}
+
+\begin{minipage}[t]{0.4\textwidth}
+\begin{center}
+Before:
+
+\begin{tikzpicture}[scale=1.5]
+  \node (t) at (0,0) {$t$};
+  \node (r) at (-1,-2) {$r$};
+  \node (s) at (1,-1) {$s$};
+  \draw[right hook-latex] (r) to (t) {};
+  \draw[right hook-latex] (s) to (t) {};
+
+  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
+  %\node at (t1.north west) {$S$};
+  %\node at (t1) {$D$};
+  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
+  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
+  %\node at (t2.north west) {$T$};
+  %\node at (t2) {$C$};
+  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
+\end{tikzpicture}
+\end{center}
+\end{minipage}
+\begin{minipage}[t]{0.4\textwidth}
+\begin{center}
+After:
+
+\begin{tikzpicture}[scale=1.5]
+  \node (t) at (0,0) {$t$};
+  \node (rp) at (-1,-1) {$r'$};
+  \node (s) at (1,-1) {$s$};
+
+  \node (r) at (-1,-2) {$r$};
+  \node (sp) at (1,-2) {$s'$};
+  \draw[right hook-latex] (r) to (rp) {};
+  \draw[right hook-latex] (sp) to (s) {};
+
+  \draw[right hook-latex] (rp) to (t) {};
+  \draw[right hook-latex] (sp) to (t) {};
+
+  \draw[arrow] (s) to (t);
+
+  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
+  %\node at (t1.north west) {$S$};
+  %\node at (t1) {$D$};
+  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
+  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
+  %\node at (t2.north west) {$T$};
+  %\node at (t2) {$C$};
+  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
+\end{tikzpicture}
+\end{center}
+\end{minipage}
+
+\end{center}
+\caption{An Example for Transparent Refactoring}\label{fig:refac}
+\end{figure}

### 2018-10-14 example.tex
added links for bands

@@
 There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
-All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\ednote{@DM: no need to give the morphisms, but link to the online resource if you have time to work them all out}
+All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\footnote{Our formalization of the lattice of bands using implict morphisms can be found at \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/bands.mmt} and \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/lattice.mmt}}
 We make all of these morphisms implicit.

### 2018-10-14 appl.tex
some clean up

@@
 
-\begin{example}
-	Consider a variation of the theory $\cn{Monoid}$ from \autoref{syn:incl} in a different library:
+\begin{example}[Renaming]
+Consider a variation of the theory $\cn{Monoid}$ from Ex.~\ref{syn:incl} in a different library:
 \begin{mmtcode}
@@
 \end{mmtcode}
-This theory is isomorphic to the previously introduced theory $\cn{Monoid}$ under the trivial renaming
+This theory is isomorphic to the previously introduced theory \cn{Monoid} under the trivial renaming
 	\begin{mmtcode}
-Monoids : Monoid2 -> Monoid =
+renaming MonoidRen : Monoid2 -> Monoid =
   M 		:= U
@@
 \end{compactitem}
-\begin{example}
-	Imagine extending the theory $\cn{Group}$ from \autoref{syn:incl} by provable theorems:
-		\begin{mmtcode}
-Group_Theorems =
-  include Group
-  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
-\end{mmtcode}
-\end{example}
 
@@
 \end{compactitem}
-\begin{example}
-	We adapt our theory $\cn{Group\_Theorems}$ with the newly introduced keyword, which allows us to extend $\cn{Group}$ directly while having the new theorem $\cn{inverse\_idem}$ available:
+
+\begin{example}[Extension with a Theorem]
+We extend the theory $\cn{Group}$ from Ex.~\ref{syn:incl} with a theorem
 \begin{mmtcode}
-Group_Theorems =
-  definitional include Group
-  inverse_idem : ⊦∀[x] (x⁻¹)⁻¹≐x 	= (proof)
-
-More_Theorems =
-  include ?Group
-  some_other_theorem : ⊦ ??	= (proof using inverse_idem)
+definitional InverseInvolution =
+  include Group
+  inverse_invol : ∀[x] (x⁻¹)⁻¹≐x 	= (proof omitted)
 \end{mmtcode}
@@
 \begin{example}\label{group:iso}
-Having declared the morphism $\cn{DG2G}$ (as in \autoref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
+Having declared the morphism $\cn{DG2G}$ (as in Ex.~\ref{ex:dg2g}) implicit, we do the same with the reverse morphism  $\cn{G2DG}$:
 \begin{mmtcode}
@@
   U := U
-  op := [a,b] a / ((unit/unit) / b) 
-  unit := unit/unit
-  inverse := [a] (unit/unit) / a
+  op := [a,b] a/(unit/b) 
+  unit := unit
+  inverse := [a] unit/a
 \end{mmtcode}
@@
 Now $t$ has the desired new structure.
-But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in \autoref{fig:refac}.
+But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in Fig.~\ref{fig:refac}.
 
-\begin{figure}\begin{center}
-
-\begin{minipage}[t]{0.4\textwidth}
+The situation before (left) and after (right) refactoring is given below:
 \begin{center}
-Before:
-
-\begin{tikzpicture}[scale=1.5]
+\begin{tikzpicture}[scale=1]
   \node (t) at (0,0) {$t$};
@@
 \end{tikzpicture}
-\end{center}
-\end{minipage}
-\begin{minipage}[t]{0.4\textwidth}
-\begin{center}
-After:
-
-\begin{tikzpicture}[scale=1.5]
+\tb
+\begin{tikzpicture}[scale=1]
   \node (t) at (0,0) {$t$};
@@
 \end{center}
-\end{minipage}
-
-\end{center}
-\caption{An Example for Transparent Refactoring}\label{fig:refac}
-\end{figure}

### 2018-10-14 example.tex
some clean up

--- 2018-10-14 example.tex
@@
 \end{compactitem}
+\medskip
 
-In the sequel, we develop a similar, more complex example: we build a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
+As a more complex example, we have built a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
 \ednote{@DM: this is not quite the lattice from the Wikipedia page: it consists of the 3x3 square plus Bands at the top}
@@
 There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
-All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\ednote{@DM: no need to give the morphisms, but link to the online resource if you have time to work them all out}
+All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.%
+\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/bands.mmt}.}
 We make all of these morphisms implicit.

### 2018-10-14 mmt.tex
some clean up

--- 2018-10-13 mmt.tex
@@
 
-The declarations in theories and morphisms are subject to typing conditions that are not essential for our purposes here.
-We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
-This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.
-
 An \mmt \textbf{diagram} consists of a set of named structured module declarations.
@@
 
-\paragraph{Well-Formed Expressions}
-The inference system to define well-formed expressions is not a primary interest of this paper.
-Intuitively, the set $\Exp{T}$ of expressions over $T$ includes all syntax trees that can be formed using the constants from $\flt{T}$.
-Moreover, well-formed modules may only use expressions that satisfy certain typing and equality judgments.
+\paragraph{Logics and Well-Formed Expressions}
+The logic and the definition of well-formed expressions are not a primary interest of this paper, and we only recap the essential structure needed in the sequel.
 We refer to \cite{rabe:howto:14} for details.
 
-The only aspect of the inference system for these judgments that is relevant to our purposes here is the rules for constants:
+\mmt itself is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
+Our results in this paper are similarly logic-independent.
+The syntax and semantics of that base logic is provided by a set of rules for the judgments $\vdash_T e:E$ and $\vdash_T e=e'$ for $e,e',E\in\Exp{T}$.
+The declarations in theories and morphisms are subject to typing conditions using these judgments.
+The main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, e.g,., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
+This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.
+
+The only aspect of the inference system for these judgments that is relevant to our purposes here is the typing rule for constants:
 \[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
@@
 
-\paragraph{Logics in \mmt}
-\ednote{explain that we use a fixed instance of MMT which adds LF and FOL; only relevant for examples}
-
 \subsection{Syntax}
@@
 \begin{example}\label{ex:thgroup}
-The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
+For the purposes of our running examples, we assume a fixed base logic that provides a first-order logic a simple type system.
+\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
+We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.
+
+Then the (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
 \begin{mmtcode}
@@
 \end{mmtcode}
-Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.
-
-Moreover, we use a few constants from a type systems\ednote{finish}$\forall,\doteq$ etc.
 \end{example}

### 2018-10-14 appl.tex
cleaned up intro and conc

@@
 
-In this section, we introduce several language extensions that introduce implicit isomorphisms.
+A common need in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
+In these cases, it is often desirable to identify $S$ and $T$, e.g., when working with $S$, we want to be able to use $T$-syntax directly without having to apply an isomorphism.
+One of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
+In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification.
 

### 2018-10-14 conc.tex
update
cleaned up intro and conc

--- 2018-09-26 appl.tex
@@
-Canonical isomorphisms.
+\paragraph{Implicit Conversions}
+The need for implicit conversions has been recognized in many formal systems.
+In all cases, similar uniqueness constraints are employed as in ours.
 
-User-declared atomic morphisms.
+\textbf{Type-level} conversions are functions between types such as the conversion from natural numbers to integers.
+\textbf{Theory-level} conversions morphisms between theories, like in this paper.
+The latter can be seen as a special case of the former: if every theory is seen as the type of its models (as in \cite{MRK:modelsof:18}), then reduction along an implicit morphism $S\to T$ induces a conversion from $T$-models to $S$-models.
+Type-level conversions are present in many systems, e.g., the Coq proof assistant \cite{coq} or the Scala programming language.
+%In both cases, this includes theory-level conversions between the respective analogues of theories
+The novelty in our approach is to restrict conversions two-fold: firstly to the theory level, secondly to those conversion functions that can be expressed as theory morphisms.
+This significantly reduces the complexity and permits an elegant logic-independent semantics, while still being practically useful.
 
-\paragraph{Extending Theories from the Outside}
+Some formal systems support theory-level conversions without explicitly using theory morphisms.
+This is common in systems that use type classes as an analogue to theories.
+For example, the \texttt{sublocale} declarations of the proof assistant Isabelle \cite{isabelle_locales} or the \texttt{deriving} declarations of the programming language Haskell can be seen as \textbf{anonymous} implicit morphisms.
+Making theory morphisms primitive objects, as in our case, yields a simpler and more expressive implicit-conversion-system at the price of having an additional primitive concept.
+% Löh paper on deriving via not published yet
+%\cite{gonthier_packaging}
 
-\paragraph{Transparent Refactoring}
-A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
-This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
-That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.
+\paragraph{Structuring Theories}
+In systems that maintain large diagrams of theories, the problems solved by our approach have been recognized for some time.
+For example, the IMPS system \cite{imps} allowed using theory morphisms to retroactively add defined constants to a previously declared theory.
+This corresponds to a definitional extension with an implicit retraction morphism as in Sect.~\ref{sec:inverse}.
 
-With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
-Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
-All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.
+In \cite{realms}, the idea of \emph{realms} was introduced as a way to bundle a set of isomorphic theories and their definitional extensions into a single interface.
+The paper called for an implementation of realms as a new primitive concept in addition to theories and morphisms.
+In contrast, the much simpler feature of implicit morphisms achieves very similar goals: realms can be recovered by marking all isomorphisms as \emph{implicit} and all extensions as \emph{definitional}.
+
+%\begin{compactitem}
+%\item a set of isomorphic theories $B_1,\ldots,$
+%\item for each $B_i$, a set of retractable extensions $E^i_1,\ldots,$
+%\item a theory $F$ including all of the above.
+%\end{compactitem}
+%The intuition is that the $B_i$ are alternative definitions of a theory; the $E^i_j$ add definitions and theorems; and $F$ aggregates everything into the outwardly visible interface (the face) of the realm.
+%A good example is the realm of topological spaces with the $B_i$ corresponding to the various possible definitions (via neighborhoods, open sets, closure operator, etc.).
+%
+%Users of the realm include only the face and thus (i) gain access to all derived knowledge in the theory of topological spaces, and (ii) do not have to commit to one particular definition.
+%Indeed, this corresponds more closely to the way how conventional mathematics sees the theory of topological spaces.
+
+\paragraph{Scalability and Scoping}
+Future work will focus on utilizing and evaluating implicit morphisms in large libraries, i.e., diagrams with thousands of theories and as many implicit morphisms as possible.
+
+In doing so, we will pay particular attention to some problems that implicit conversions can cause at large scales.
+Users can be confused when implicit conversions are applied that they are not aware of.
+Different users may also have different preferences for which conversions should be implicit.
+And most critically, different developments may be incompatible if they introduce different implicit morphisms between the same theories.
+For those reasons, Scala, for example, only applies implicit conversions that are imported into the current namespace.
+
+We anticipate that these problems will lead to an evolution of our solution that allows more localized control over which morphisms are implicit.
+Thus, instead of a single global diagram of implicit morphisms, every context may carry its own local one.
+But we defer this until the current implementation has been used to conduct very large case studies.
+
+%Moreover, the use of qualified identifiers in \mmt means that readers can always tell where an identifier is implicitly-included from, and the \mmt IDE can always show the implicit morphism along which it is imported.

### 2018-10-14 example.tex
cleaned up intro and conc

--- 2018-10-14 example.tex
@@
 \end{compactitem}
+\medskip
 
-In the sequel, we develop a similar, more complex example: we build a hierarchy of theories between \cn{Band} and \cn{Semilattice}.
-\ednote{@DM: this is not quite the lattice from the Wikipedia page: it consists of the 3x3 square plus Bands at the top}
-All of these theories are of the form $t=\{\icl{\cn{Band}},\,a:\vdash F\}$, where $F$ is a universally quantified equational axiom.\footnote{All varieties of bands can be axiomatized in this way.}
-In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a:\vdash \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
-Fig.~\ref{fig:bands} shows these theories.
+As a more complex example, we have formalized a hierarchy of theories between \cn{Band} and \cn{Semilattice}:
+\begin{center}
+\begin{tikzpicture}[xscale=2]
+  \node (s1) at (0,0) {$z\circ x\circ y\circ z\doteq  z\circ x\circ z\circ y\circ z$};
 
-There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
-All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.\ednote{@DM: no need to give the morphisms, but link to the online resource if you have time to work them all out}
-We make all of these morphisms implicit.
-It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom.
-Here, \mmt can discharge the proof obligation easily using proof irrelevance.
+  \node (s2a) at (-1,-1) {$z\circ x\circ y\doteq  z\circ x\circ z\circ y$};
+  \node (s2b) at (1,-1) {$y\circ x\circ z\doteq  y\circ z\circ x\circ z$};
+  \draw[arrow] (s1) to (s2a) {};
+  \draw[arrow] (s1) to (s2b) {};
 
-\begin{figure}\begin{center}
-\begin{tikzpicture}[scale=1.5]
-  \node (s1) at (0,0) {$zxyz=zxzyz$};
+  \node (s3a) at (-2,-2) {$x\circ y\doteq  x\circ y\circ x$};
+  \node (s3b) at (0,-2) {$z\circ x\circ y\circ z\doteq  z\circ y\circ x\circ z$};
+  \node (s3c) at (2,-2) {$x\circ y\doteq  y\circ x\circ y$};
+  \draw[arrow] (s2a) to (s3a) {};
+  \draw[arrow] (s2a) to (s3b) {};
+  \draw[arrow] (s2b) to (s3b) {};
+  \draw[arrow] (s2b) to (s3c) {};
 
-  \node (s2a) at (-1,-1) {$zxy=zxzy$};
-  \node (s2b) at (1,-1) {$yxz=yzxz$};
-  \draw (s1) to (s2a) {};
-  \draw (s1) to (s2b) {};
+  \node (s4a) at (-1,-3) {$z\circ x\circ y\doteq  z\circ y\circ x$};
+  \node (s4b) at (1,-3) {$x\circ y\circ z\doteq  y\circ x\circ z$};
+  \draw[arrow] (s3a) to (s4a) {};
+  \draw[arrow] (s3b) to (s4a) {};
+  \draw[arrow] (s3b) to (s4b) {};
+  \draw[arrow] (s3c) to (s4b) {};
 
-  \node (s3a) at (-2,-2) {$xy=xyx$};
-  \node (s3b) at (0,-2) {$zxyz=zyxz$};
-  \node (s3c) at (2,-2) {$xy=yxy$};
-  \draw (s2a) to (s3a) {};
-  \draw (s2a) to (s3b) {};
-  \draw (s2b) to (s3b) {};
-  \draw (s2b) to (s3c) {};
-
-  \node (s4a) at (-1.5,-3) {$zxy=zyx$};
-  \node (s4b) at (0,-3) {$xyx=x$};
-  \node (s4c) at (1.5,-3) {$xyz=yxz$};
-  \draw (s3a) to (s4a) {};
-  \draw (s3b) to (s4a) {};
-  \draw (s3b) to (s4b) {};
-  \draw (s3b) to (s4c) {};
-  \draw (s3c) to (s4c) {};
-
-  \node (s5a) at (-1.5,-4) {$xy=x$};
-  \node (s5b) at (0,-4) {$xy=yx$};
-  \node (s5c) at (1.5,-4) {$xy=y$};
-  \draw (s4a) to (s5a) {};
-  \draw (s4a) to (s5b) {};
-  \draw (s4b) to (s5a) {};
-  \draw (s4b) to (s5c) {};
-  \draw (s4c) to (s5b) {};
-  \draw (s4c) to (s5c) {};
-
-  \node (s6) at (0,-5) {$x=y$};
-  \draw (s5a) to (s6) {};
-  \draw (s5b) to (s6) {};
-  \draw (s5c) to (s6) {};
+  \node (s5) at (0,-4) {$x\circ y\doteq  y\circ x$};
+  \draw[arrow] (s4a) to (s5) {};
+  \draw[arrow] (s4b) to (s5) {};
 
@@
 \end{tikzpicture}
+\end{center}
+All of these theories are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, where $F$ is the formula given in the diagram above.%
+\footnote{All varieties of bands can be axiomatized in this way.}
+In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a: \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
 
-%\includegraphics[width=0.6\textwidth]{bands}
-\end{center}
-\caption{The Lattice of Varieties of Bands}\label{fig:bands}
-\end{figure}
+There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
+All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
+In our formalization, we make all of these morphisms implicit.%
+\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/bands.mmt}.}
+It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom, and these are provably equal due to proof irrelevance.
 
-%
 %\begin{figure}

### 2018-10-14 intro.tex
cleaned up intro and conc

@@
 These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
-This diagram is generated by all theories/morphisms defined, induced, or referenced in a user's development.
+This diagram is generated by all theories and morphisms occurring in a user's development.
 
@@
 For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
-This would result in the diagram on the right.
-Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
-Crucially, $m(-)$ preserves typing and provability.
+This would result in the diagram on the right.%
+\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
+Models are translated in the opposite direction.}
+%Crucially, $m(-)$ preserves typing and provability.
 
@@
 The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
-Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions; and if $c$ is used in a $T$-expression, the semantics of $c$ is $i(c)$ where $i$ is the uniquely determined implicit morphism $i:S\to T$.
-
-Despite their simplicity, the practical implications of implicit morphism are huge.
-For example, in the diagram above, we may choose to label $\cn{G2DG}$ implicit.
-Immediately, every abbreviation or theorem that we have formulated in the theory $\cn{Group}$ becomes available for use in $\cn{DivGroup}$ without any syntactic overhead.
-We can even label $\cn{DG2G}$ implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that $\cn{Group}$ and $\cn{DivGroup}$ are just different formalizations of the same concept.
-While these morphisms must be labeled manually, any inclusion morphism like the one from $\cn{Monoid}$ to $\cn{Group}$ is implicit automatically.
+Commutativity makes the following language extension well-defined: if $\ipc{i}{S}{T}$, then any identifier $c$ that is visible to $S$ may also be used in $T$-expressions, with the semantics being that $c$ abbreviates $i(c)$.
+For example, in the diagram above, we may choose to label \cn{DG2G} implicit.
+Immediately, every abbreviation or theorem that we have formulated in the theory \cn{DivGroup} becomes available for use in \cn{Group} without any syntactic overhead.
+We can even label \cn{G2DG} implicit as well if we prove the isomorphism property to ensure that $I$ remains commutative, thus capturing the mathematical intuition that \cn{Group} and \cn{DivGroup} are just different formalizations of the same concept.
+While these morphisms must be labeled manually, any inclusion morphism like the one from \cn{Monoid} to \cn{Group} can be made implicit automatically.
 
 \paragraph{Contribution}
-We present a formal system for developing structured theories with implicit morphisms.
+At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
+We recommend using implicit morphisms in all theory morphism--based formalisms.
+
+More concretely, we present a formal system for developing structured theories with implicit morphisms.
 Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
@@
 
-More precisely, our contribution is twofold.
-\medskip
+We describe several example applications of implicit morphisms in detail: the identification of isomorphic theories, definitional extensions of theories, building large hierarchies of theories with many rarely used intermediate theories, and transparently refactoring theory hierarchies.
 
-Firstly, we give a novel presentation of the syntax and semantics of \mmt.
-Contrary to the original presentation in \cite{RK:mmt:10}, which hard-coded a single structuring mechanism --- there called \textit{structures} --- our definitions allows for adding arbitrary independent structuring mechanisms.
-This required a from-scratch redesign of the original presentation in \cite{RK:mmt:10} (which is why this paper has no preliminaries section).
+\paragraph{Overview}
+In Sect.~\ref{sec:mmt}, we present the syntax and semantics of \mmt.
+Even though the \mmt language is not new, our presentation is an entirely novel contribution in itself: it is much simpler and more elegant than the original one in \cite{RK:mmt:10}.
+%Based on this extensible definition, we can introduce several additional structuring mechanisms.
+%Many of these have already been implemented in the \mmt tool but were not formally defined --- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
+%With our reformulation, this is now not only possible but very easy.
+%As an example, we define include declarations, the most important special case of implicit morphisms.
+%\medskip
+%
+%Secondly, we describe implicit morphisms as a novel feature of \mmt.
+Crucially, this increase in simplicity allows spelling out the syntax and semantics of implicit morphisms, which we do in Sect.~\ref{sec:impl}, within a few pages.
+In Sect.~\ref{sec:appl}, we present applications.
+Finally we review related and future work in Sect.~\ref{sec:conc}.
 
-Based on this extensible definition, we can introduce several additional structuring mechanisms.
-Many of these have already been implemented in the \mmt tool but were not formally defined --- a formal definition of their semantics as a part of the \mmt language had previously proved too complicated to spell out elegantly.
-With our reformulation, this is now not only possible but very easy.
-As an example, we define include declarations, the most important special case of implicit morphisms.
-\medskip
-
-Secondly, we describe implicit morphisms as a novel feature of \mmt.
-Due to our novel presentation of the syntax and semantics, only minor modifications are needed to define the syntax and semantics of implicit morphisms.
-
-A very common situation in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
-In these cases, it is often desirable to identify $S$ and $T$, e.g., when working with $S$, we want to be able to use $T$-syntax directly without having to apply an isomorphism.
-One of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
-We show that implicit morphisms elegantly allow exactly that kind of intuitive identification.
-As concrete examples, we consider manual isomorphisms, definitional extensions, and renamings.
-In particular, we recover the concept of realms \cite{realms} as a special case of implicit morphisms.
-
-In fact, implicit morphisms work so well that we have refactored the \mmt tool in such a way that implicit morphisms are now more primitive than inclusion morphisms.
-The semantics of inclusion morphisms is obtained by saying that inclusions are implicit morphisms that map all identifiers to themselves.
-Even the fundamental property that a theory may reference its own identifiers is then just a consequence of the fact that all identity morphisms are implicit.
-Therefore, surprisingly, adding the new feature of implicit morphisms to the \mmt kernel has made its design much simpler.
+%In fact, implicit morphisms work so well that we have refactored the \mmt tool in such a way that implicit morphisms are now more primitive than inclusion morphisms.
+%The semantics of inclusion morphisms is obtained by saying that inclusions are implicit morphisms that map all identifiers to themselves.
+%Even the fundamental property that a theory may reference its own identifiers is then just a consequence of the fact that all identity morphisms are implicit.
+%Therefore, surprisingly, adding the new feature of implicit morphisms to the \mmt kernel has made its design much simpler.

### 2018-10-14 mmt.tex
cleaned up intro and conc

@@
-\ednote{@Dennis: can you formally work through the example from the intro in here?
-That would be several example environments for (i) after introducing the syntax of theories, the theory Group and DivGroup (one in detail, one sketched), (ii) after introducing morphisms the two isomorphisms (one in detail, one sketched), (iii) after introducing includes, the refactoring of Group using an include, (iv) examples for how the refactored Group is flattened}
-
 \subsection{Overview}

### 2018-10-14 appl.tex

@@
 \end{compactenum}
-Now $t$ has the desired new structure.
-But, all old references to $s?n$ stay well-formed so that no global changes are needed, as in Fig.~\ref{fig:refac}.
-
 The situation before (left) and after (right) refactoring is given below:
@@
 \end{center}
+Afterwards, $t$ has the desired new structure.
+But all old references to $s?n$ stay well-formed so that no global changes are needed.

### 2018-10-14 example.tex

@@
 In our formalization, we make all of these morphisms implicit.%
-\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MitM/smglom/blob/devel/source/algebra/bands.mmt}.}
+\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
 It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom, and these are provably equal due to proof irrelevance.

### 2018-10-19 appl.tex
colons
finished for submission

@@
 A common need in developing large libraries (both in formal and informal developments) is to identify two theories $S$ and $T$ via a canonical choice of isomorphisms.
-In these cases, it is often desirable to identify $S$ and $T$, e.g., when working with $S$, we want to be able to use $T$-syntax directly without having to apply an isomorphism.
-One of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
+In these cases, it is often desirable to use $S$-syntax and $T$-syntax interchangeably.
+But one of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
 In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification.
 
-Note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
+In the sequel, we present several ways to obtain implicit isomorphisms conveniently.
+In general, note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
 Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.
@@
   \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
-  \item This is well-formed if there are no implicit morphism between $s$ and $t$ yet.
+  \item This is allowed if there are no implicit morphism between $s$ and $t$ yet.
   \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
@@
 Monoid2 =
-  M: type
-  connective : M ⟶ M ⟶ M	 # 1 ∘ 2 
+  M          : type
+  connective : M $\to$ M $\to$ M	 # 1 $\circ$ 2 
   neutral    : M
@@
 \paragraph{Definitional Extensions}
-We say that the named theory $t$ is a \textbf{definitional extension} of $S$ if $t=S$ or the body of $t$ contains
+We say that the named theory $T$ is a \textbf{definitional extension} of $S$ if $T=S$ or the body of $T$ contains
 \begin{compactitem}
@@
 
-If $t$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
-In particular, the isomorphism $S\to t$ maps $S$-constants to themselves and expands the definiens of all other constants.
+If $T$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
+In particular, the isomorphism $S\to T$ maps $S$-constants to themselves and expands the definiens of all other constants.
 
@@
 \begin{compactitem}
-  \item An include declaration $\icl{s}$ of a named theory $s$ inside theory $t$ may carry the attribute \keyword{definitional}.
-  \item In that case, we define $\id{s}\in\IMo{t}{s}$ (in addition to the implicit morphism $\id{t}\in\IMo{s}{t}$ which is induced by the inclusion).
+  \item An include declaration $\icl{S}$ of a named theory $S$ inside theory $T$ may carry the attribute \keyword{definitional}.
+  \item In that case, we define $\id{S}\in\IMo{T}{S}$ (in addition to the implicit morphism $\id{T}\in\IMo{S}{T}$ that is induced by the inclusion).
 \end{compactitem}
@@
 \begin{mmtcode}
-definitional InverseInvolution =
-  include Group
-  inverse_invol : ∀[x] (x⁻¹)⁻¹≐x  =  (proof omitted)
+InverseInvolution =
+  definitional include Group
+  inverse_invol : $\forall$[x] (x$^{-1}$)$^{-1}$$\doteq$ x  =  (proof omitted)
 \end{mmtcode}
@@
 A definitional extension is a special case of a conservative extension.
-More generally, all retractable extensions are conservative, i.e., all extensions $S\harr T$ such that there is a morphism $r:T\to S$ such that $r$ is the identity on $S$.
-
+More generally, all retractable extensions are conservative, i.e., all extensions $S\harr T$ such that there is a morphism $r:T\to S$ that is the identity on $S$.
 But we cannot make the retractions implicit morphisms in general because they are not necessarily isomorphisms.
@@
 \begin{mmtcode}
-implicit G2DG : Group -> DivisionGroup =
+implicit G2DG : Group -> DivGroup =
   U       := U
@@
 
-While the first one of these isomorphism declarations is straightforward, the second one requires checking that $m$ and $n$ are actually isomorphisms.
+While making one of these isomorphisms implicit is straightforward, doing both requires checking that $m$ and $n$ are actually isomorphisms.
 Otherwise, the uniqueness condition would be violated.
 Thus, we have to check $m;n=\id{s}$ and $n;m=\id{t}$.
-In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $c:E\in\flt{A}$.
+In general, the equality of two morphisms $f,g:A\to B$ is equivalent to $\vdash_B f(c)=g(c)$ for all $(c:E)\in\flt{A}$.
 Thus, if equality of expressions is decidable in the logic that \mmt is instantiated with, then \mmt can check this directly.
@@
 However, this does not work in practice.
-Already elementary examples require stronger, undecidable equality relations are used:
+Already elementary examples require stronger, undecidable equality relations:
 
@@
 Consider the isomorphism from Ex.~\ref{group:iso}.
-The result of mapping $x\circ y$ from $\cn{Group}$ to $\cn{DivGroup}$ and back is $x\circ(unit\circ y^{-1})^{-1}$.
+The result of mapping $x\circ y$ from $\cn{Group}$ to $\cn{DivGroup}$ and back is $x\circ(\cn{unit}\circ y^{-1})^{-1}$.
 Clearly, the group axioms imply that this is equal to $x\circ y$.
-But formally that requires working with the undecidable equality of first-order logic.
+But formally, that requires working with the undecidable equality of first-order logic.
 \end{example}
 
-Therefore, in our running example, we only make one of the two isomorphisms implicit.
+Therefore, in our running example, we can only make one of the two isomorphisms implicit at this point.
 
@@
 It allows systematically proving the equality of two morphisms and using that to make both isomorphisms implicit.
-This is novel work, but it requires significant prerequisites and is only peripherally related to implicit morphisms.
+This is novel work that requires significant prerequisites and is only peripherally related to implicit morphisms.
 Therefore, we only sketch the idea and leave the details to future work.
@@
 Dia   & \rep{(TDec \alt MDec\alt MEq)}  &\\
-MEq   & \keyword{equal} M=M:T\to T by \{\rep{Ass}\} &\\
+MEq   & \keyword{equal} M=M:T\to T \keyword{by} \{\rep{Ass}\} &\\
 \end{grammar}
@@
 
-We define the declaration $M=N:S\to T by \{\sigma\}$ to be well-formed iff
+We define the declaration $M=N:S\to T \keyword{by} \{\sigma\}$ to be well-formed iff
 \begin{compactitem} 
@@
 
-To make $m$ and $n$ from above implicit isomorphisms, we have to do three things: define $m$ and $n$, prove the equalities of $m;n=\id{s}$ and $n;m=\id{t}$, and make $m$ and $n$ implicit.
-Note that we cannot make both $m$ and $n$ implicit right away because that is only well-formed after proving the equalities.)
-Thus, we define a new attribute \keyword{implicit-later}, which states that a morphism should be considered implicit as soon as subsequent equality proves make it well-formed to do so.
-
 \begin{example}[Isomorphisms]
-We can now add declarations
+With the above extension in place, we can make both isomorphisms $m$ and $n$ from above implicit:
  \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
@@
 where the isomorphisms are as above and we omit all the equality proofs.
+
+Note that we cannot mark \cn{G2DG} as \keyword{implicit} because that is only well-formed after proving the equalities.
+So we use the new attribute \keyword{implicit-later}, which states that a morphism should be considered as implicit as soon as subsequent equality proofs make it well-formed to do so.
 \end{example}
@@
 For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $r$ to $s$.
-Thus, the change to should be straightforward as it does not change the semantics of $t$.
+This change to should be straightforward as it does not change the semantics of $t$.
 
@@
 It also requires updating every qualified reference from $r?n$ to $s?n$.
-Such references can occur anywhere where $t$ is used.
-That may include theories that the person who does the refactoring does not know or does not have access to.
 Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
+But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know or does not have access to.
 
@@
   \item We rename $s$ to $s'$.
-  \item We delete the declaration $n:E$ from $s'$.
+  \item We delete the declaration $D$ from $s'$.
   \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.

### 2018-10-19 conc.tex
finished for submission

@@
 \textbf{Type-level} conversions are functions between types such as the conversion from natural numbers to integers.
-\textbf{Theory-level} conversions morphisms between theories, like in this paper.
+\textbf{Theory-level} conversions are morphisms between theories, like in this paper, or similar constructs.
 The latter can be seen as a special case of the former: if every theory is seen as the type of its models (as in \cite{MRK:modelsof:18}), then reduction along an implicit morphism $S\to T$ induces a conversion from $T$-models to $S$-models.
@@
 This is common in systems that use type classes as an analogue to theories.
-For example, the \texttt{sublocale} declarations of the proof assistant Isabelle \cite{isabelle_locales} or the \texttt{deriving} declarations of the programming language Haskell can be seen as \textbf{anonymous} implicit morphisms.
-Making theory morphisms primitive objects, as in our case, yields a simpler and more expressive implicit-conversion-system at the price of having an additional primitive concept.
+For example, the \texttt{sublocale} declarations of the proof assistant Isabelle \cite{isabelle_locales} or the \texttt{deriving} declarations of the programming language Haskell can be seen as implicit morphisms even though no primitive concept of morphism objects is employed.
+Our implicit morphisms yield a simpler and more expressive theory-level conversion system at the price of having an additional primitive concept.
 % Löh paper on deriving via not published yet
@@
 In doing so, we will pay particular attention to some problems that implicit conversions can cause at large scales.
-Users can be confused when implicit conversions are applied that they are not aware of.
-Different users may also have different preferences for which conversions should be implicit.
-And most critically, different developments may be incompatible if they introduce different implicit morphisms between the same theories.
+Users can be confused when implicit conversions are applied that they are not aware of, and different users may also have different preferences for which conversions should be implicit.
+Moreover, critically, different developments may be incompatible if they introduce different implicit morphisms between the same theories.
 For those reasons, Scala, for example, only applies implicit conversions that are imported into the current namespace.

### 2018-10-19 example.tex
finished for submission

@@
 Instead, we could include \cn{Monoid} into the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
-This change is not possible in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents an insurmountable refactoring problem.
+This change is not easy in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents a very expensive refactoring problem.
 So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
@@
 
-As a more complex example, we have formalized a hierarchy of theories between \cn{Band} and \cn{Semilattice}:
+As a more complex case study, we have used our implementation of implicit morphisms in \mmt to formalize a hierarchy of theories between \cn{Band} and \cn{Semilattice}:% 
+\footnote{Our formalization can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
 \begin{center}
@@
 All of these theories are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, where $F$ is the formula given in the diagram above.%
-\footnote{All varieties of bands can be axiomatized in this way.}
+%\footnote{All varieties of bands can be axiomatized in this way.}
 In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a: \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
 
-There are various morphisms between them that describe the lattice structure of the corresponding varieties of bands.
-All of these map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
-In our formalization, we make all of these morphisms implicit.%
-\footnote{Our formalization of the lattice of bands using implicit morphisms can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
-It is straightforward to prove that the diagram commutes: any two morphisms are identical except for assignment for the axiom, and these are provably equal due to proof irrelevance.
+The diagram above gives all morphisms between them that describe the lattice structure of the corresponding varieties of bands, all of which map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
+In our formalization, we make all of these morphisms implicit.
+It is straightforward to prove that the diagram commutes: any two morphisms are identical except for the assignment to the axiom $a$, and these are equal due to proof irrelevance.
 

### 2018-10-19 implicit.tex
finished for submission

@@
 
-The motivation behind implicit morphisms is to generalize the benefits of include declarations even further.
-Sometimes an atomic morphism is conceptually similar to an inclusion: For example, we might want to declare $m:\cn{Division}\to\cn{Group}=\{\cn{divide} := \lambda x,y.x\circ y^{-1}, \;\ldots\}$ to show how division is ``included'' into the theory of groups.
-Other times we have to use structures because we need to duplicate declarations but we would still like to use them like includes.
-For example, we have to declare the theory of rings using two structures (for the commutative group and the monoid).
-But we still want a ring to be implicitly converted into, e.g., a commutative group.
-
-Our key idea is to use a commutative{\footnotemark} subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
-It contains all theories but only some of the morphisms -- the ones designated as \emph{implicit}.
-Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ -- if this morphism is $M$, we write $\ipc{M}{S}{T}$.
-\footnotetext{A diagram commutes if for any two morphisms $M,M'\in\Mo{S}{T}$, we have $M\equiv M'$.}
+Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
+It contains all theories but only some of the morphisms --- the ones designated as \emph{implicit}.
+Because the implicit-diagram commutes, there can be at most one implicit morphism from $S$ to $T$ --- if this morphism is $i$, we write $\ipc{i}{S}{T}$.
 %Practically, this hasis a multi-graph whose nodes are the theories and whose edges are the atomic morphisms.
@@
 %A diagram is commutative if any two paths between the same nodes are equal (with respect to $\equiv$), i.e., there is at most one morphism between any two nodes.
+The implicit-diagram generalizes the inclusion relation:
+All identity and inclusion morphisms are implicit, and we recover the inclusion relation $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
+And just like inclusion, the relation ``exists $i$ such that $\ipc{i}{S}{T}$'' is a preorder.
 
-The implicit-diagram generalizes the inclusion relation $\harr$.
-In particular, all identity and inclusion morphisms are implicit, and we recover $S\harr T$ as the special case $\ipc{\id{S}}{S}{T}$.
-Moreover, the relation ``exists $M$ such that $\ipc{M}{S}{T}$'' is also an order.
-
-Consequently, many of the advantages of inclusions carry over to implicit morphisms.
+Consequently, many of the advantages of inclusions carry over to implicit morphisms:
 \begin{compactitem}
 \item It is very easy to maintain the implicit-diagram, e.g., as a partial map that assigns to a pair of theories the implicit morphism between them (if any).
-\item We can generalize the visibility of identifiers: If $\ipc{M}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
+\item We can generalize the visibility of identifiers: If $\ipc{i}{S}{T}$, we can use all $S$-identifiers in $T$ as if $S$ were included into $T$.
 Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
-\item We can still use canonical identifiers $s?n$. Because there can be at most one implicit morphism $\ipc{M}{s}{T}$, using $s?n$ as an identifier in $T$ is unambiguous.
-Crucially, we can use $s?n$ without bothering what $M$ is.
+\item We can use canonical identifiers $S?n$ without worrying about ambiguity. Because there can be at most one implicit morphism $\ipc{i}{S}{T}$, using $S?n$ as an identifier in $T$ is unambiguous.
+%Crucially, we can use $S?n$ without bothering what $i$ is.
 \end{compactitem}
@@
 It remains to define which morphisms are implicit.
-For that purpose, we allow \mmt declarations to carry \textit{attributes}.
-Precisely, we add the following productions
+For that purpose, we allow \mmt declarations to carry \textit{attributes}:
+
+\begin{definition}[Attributes for Implicitness]
+We add the following productions
 \begin{grammar}
@@
 \end{grammar}
+\end{definition}
 
-The set of attributes is itself extensible, and the above grammar only list the one that we use to get started.
+The set of attributes is itself extensible, and the above grammar only lists the one that we use to get started.
 Additional attributes can be added when adding modularity principles.
@@
 We only have to make two minor changes to the semantics to accommodate implicit morphisms.
-The first change govern how we obtain implicit morphisms, the second one how we use them.
+The first change governs how we obtain implicit morphisms, the second one how we use them.
 
-\paragraph{Obtaining Implicit Morphisms}
-Based on the above attributes, we define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
+\begin{definition}[Obtaining Implicit Morphisms]
+We define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
 \begin{itemize}
@@
 
-To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
-Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms --
-in other words, a commutative diagram.
+Adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
+\end{definition}
 
@@
 \begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
-For include morphism, we add the following definition: if theory $t$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $t$ is implicit.
+For include morphisms, we add the following definition: if theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.
 
 Combined with composition morphisms, we see that all transitive includes between theories are implicit.
-That corresponds to the intuition that anything that is include should be available directly.
+That corresponds to the intuition that anything that is included is available by its original name.
 \end{example}
@@
 \begin{example}
-	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
+	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
+	For example, we can write $a/b$ in $\cn{Group}$, where $/$ refers to the identifier $\cn{DivGroup}?\cn{div}$.
 \end{example}
@@
 
-\paragraph{Using Implicit Morphisms}
 The intuition behind implicit morphisms is that all $S$-constants $c:E$ that can be mapped into the current theory via an implicit morphism $M:S\to T$ are directly available in $T$.
@@
 However, physically adding definitions can be inefficient.
-It is more elegant to modify the typing rules such that $\vdash_T c:M(c)=M(E)$.
+It is more elegant to modify the typing rules such that $\vdash_T c:M(c)=M(E)$ holds without any changes to $T$.
 
 To do that, we only have to make a small modification to the original rules of \mmt as presented in Sect.~\ref{sec:mmt}.
-To illustrate how simple the modification is, we repeat the original rule first for comparison:
-\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
-Now our modified rule is
-\[\rul{c:E[=e] \minn \flt{S}\tb \ipc{M}{S}{T}}{\vdash_T c=M(c):M(E)}\]
+To illustrate how simple the modification is, the following definition repeat the original rule first for comparison:
+
+\begin{definition}[Using Implicit Morphisms]\label{def:useimpl}
+We replace the rule
+\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E} \tb \mwith\tb \rul{c:E[=e] \minn \flt{S}\tb \ipc{M}{S}{T}}{\vdash_T c=M(c):M(E)}\]
+\end{definition}
 
@@
 This is a technical trick to subsume the original rule: if $c$ is already declared in $T$, we use $M=\id{T}$ and obtain $\vdash_T c=c:E$.
-We write $\ov{E}$ for the expression that arises from $E$ by recursively replacing every $c$ with its definiens.
 
 The following theorem is our central theoretical result.
-It shows that the above modification has the intended properties:
+It shows that the modification made in Def.~\ref{def:useimpl} has the intended properties:
 \begin{theorem}[Conservativity of Implicit Morphisms]
-Consider any combination of modular features from the examples above that includes at least implicit identity morphisms.
-
-Then for well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$
+For well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$:
 \begin{compactenum}
  \item Whenever the original system proves $\vdash_T e=e':E$, so does the modified one.
- \item Whenever the modified system proves $\vdash_T e=e':E$, then the original inference system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
+ \item Whenever the modified system proves $\vdash_T e=e':E$, then the original system proves $\vdash_T \ov{e}=\ov{e'}:\ov{E}$.
 \end{compactenum}
+Here $\ov{E}$ is the expression that arises from $E$ by recursively replacing every constant with its definiens.
 \end{theorem}
@@
 For the first claim, we proceed by induction on derivations.
-We only need to consider case where the original rule was applied.
+We only need to consider the case where the original rule was applied.
 So assume it yields $\vdash_T c[=e]:E$, i.e., $(c:E[=e]) \minn \flt{T}$.
@@
 For the second claim, we proceed by induction on derivations.
-We only need to consider the cases where the modified rule was applied.
+We only need to consider the case where the modified rule was applied.
 So assume it yields $\vdash_T c=M(c):M(E)$ for $(c:E[=e]) \minn \flt{S}$ and $\ipc{M}{S}{T}$.
@@
 \begin{compactitem}
- \item $M(c)=c$, i.e., $c$ does not have a definiens:
-   According to the definition of $M(-)$ this is only possible if $e$ is absent.
-   According to the definition of $\flt{T}$, this is only possible if $c=S?n$ for $S\harr T$ (including the special case $S=T$).
+ \item $M(c)=c$, i.e., $e$ is absent, and $\ov{c}=c$.
+   According to the definition of $\flt{S}$, this is only possible if $S\harr T$ (including the special case $S=T$).
    In that case, $\flt{S}\sq\flt{T}$ and $M$ is the include/identity morphism that maps all $S$-constants to themselves.
    Now applying the induction hypothesis to the well-formedness derivation of $E$ yields $\vdash_T c:\ov{E}$ as needed.
- \item $M(c)\neq c$, i.e., $c$ has a non-trivial definiens:
-  Definition expansion eliminates $c$ in favor of $e$.
-  Clearly $\ov{c}=\ov{M(c)}$, and we only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
+ \item $M(c)\neq c$, i.e., $e$ is present or $M$ assigns a non-trivial value to $c$.
+  Definition expansion eliminates $c$ in favor of $M(c)$, and thus $\ov{c}=\ov{M(c)}$.
+  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
 \end{compactitem}

### 2018-10-19 intro.tex
finished for submission

@@
 They can be used to structure theories and build large theories modularly from small components or to relate different theories to each other \cite{asl,devgraphs,littletheories}.
-Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps}, and knowledge representation \cite{RK:mmt:10}.
+Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps,isabelle_locales}, and knowledge representation \cite{RK:mmt:10}.
+Closely related concepts are used in both object-oriented (\emph{classes}) and (\emph{type classes}) functional programming languages.
 
@@
 
-These systems usually use a logic $L$ for the fine-granular formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
-This diagram is generated by all theories and morphisms occurring in a user's development.
+These systems usually use a logic $L$ for the low-level formalization of domain knowledge, and a diagram $D$ in the category of $L$-theories and $L$-morphisms for the high-level structure of large bodies of knowledge.
 
@@
 
-For example, a user might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
+For example, a document might reference an existing theory $\cn{Monoid}$, define a new theory $\cn{Group}$ that extends $\cn{Monoid}$, define a theory $\cn{DivGroup}$ (providing an alternative formulation of groups based on the division operation), and then define two theory morphisms $\cn{G2DG}:\cn{Group}\darr \cn{DivGroup}:\cn{DG2G}$ that witness an isomorphism between these theories.
 This would result in the diagram on the right.%
@@
 We recommend using implicit morphisms in all theory morphism--based formalisms.
-
 More concretely, we present a formal system for developing structured theories with implicit morphisms.
 Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
-\mmt is foundation-independent in the sense that it allows embedding a large variety of declarative languages (logics, type-theories, etc.).
-Therefore, all our results are foundation-independent --- users of a particular language $L$ can use \mmt directly as a structuring formalism for $L$ or can easily transfer our results to a dedicated implementation of $L$.
+\mmt is logic-independent, i.e., allows embedding a large variety of declarative languages (logics, type-theories, etc.).
+Therefore, all our results can be directly applied to any language $L$ represented in \mmt or easily transferred to dedicated implementations of $L$.
 
@@
 In Sect.~\ref{sec:appl}, we present applications.
-Finally we review related and future work in Sect.~\ref{sec:conc}.
+Finally we discuss related and future work in Sect.~\ref{sec:conc}.
 

### 2018-10-19 macros/listings.tex
typos
finished for submission

@@
-\usepackage{listings}\lstset{basicstyle=\sf\footnotesize,columns=fullflexible}
+\usepackage{listings}
 
-\lstnewenvironment{code}[1][mathescape=false]{\lstset{language=XML, #1}}{}
-
-\newcommand{\delima}{\mid\negmedspace \mid}
-\newcommand{\delimb}{\delima\negmedspace \delima}
-\newcommand{\GS}{\ensuremath{\textcolor{RedViolet}{\delimb\negmedspace\delimb}}}
-\newcommand{\RS}{\ensuremath{\textcolor{Green}{\delimb}}}
-\newcommand{\US}{\ensuremath{\textcolor{Turquoise}{\delima}}}
-
-\lstdefinestyle{mmt}{	
-	keywordstyle=\color{Emerald},
-	columns=flexible
-}
 \lstdefinelanguage{mmt}{
-	morekeywords={theory,view,include,namespace,rule,role,implicit,import},
+	morekeywords={theory,namespace,rule,role,import,implicit,definitional,renaming},
 	morecomment=[l][\color{Gray}]{//},
@@
 }
-\lstnewenvironment{mmtcode}[1][language=mmt]{\lstset{language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
-\lstnewenvironment{nmmtcode}[1][language=mmt]{\lstset{xleftmargin=1.2em,numbers=left,language=mmt,style=mmt,inputencoding=utf8,escapechar=\&,#1}}{}
-\lstdefinestyle{scala}{	
-	keywordstyle=\color{BurntOrange}
-}
-\lstdefinelanguage{scala}{
-	emphstyle=\underbar,
-	classoffset=0,keywordstyle=\bfseries\color{BurntOrange},
-	morekeywords={case,class,object,trait,extends,match,if,else,def,implicit,import,val,var,this,override,with,return,abstract},
-	classoffset=1,keywordstyle=\color{blue},
-	morekeywords={OMA,OMS,OMBIND,OMV,OMID,Context,VarDecl,Term,LocalName,GlobalName,MPath,Solver,Stack,Rule,
-	History,Inhabitable,Subtyping,Equality,Typing,Common,Judgment,FormationRule,CheckingRule,IntroductionRule,
-	EliminationRule,InferenceRule,TypingRule,ComputationRule,TypeBasedEqualityRule,TermBasedEqualityRule,
-	TermHeadBasedEqualityRule,Continue,CheckingCallback,Simplifiability,Simplify,RecurseOnly,SubtypingRule,
-	VarianceRule,TypeBasedSolutionRule,DerivedDeclaration,StructuralFeature,Elaboration,FormatBasedExtension,
-	Marker,ExtendedCheckingEnvironment,DeclaredModule,Universe,InhabitableRule,UniverseRule,ElementContainer,Declaration},
-	classoffset=2,keywordstyle=\color{PineGreen},
-	morekeywords={Option,Some,None,Boolean,String,List,Int,true,false,Nil},
-	classoffset=0,
-	morecomment=[l][\color{Gray}]{//},
-	morecomment=[s][\color{Gray}]{/*}{*/},
-	morecomment=[s][\color{OliveGreen}]{/**}{**/},
-	morestring=[b][\color{Purple}]",
-}
-\lstnewenvironment{scalacode}[1][language=scala]{\lstset{language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}
-\lstnewenvironment{nscalacode}[1][language=scala]{\lstset{xleftmargin=1.2em,numbers=left,language=scala,style=scala,inputencoding=utf8,escapechar=\&,#1}}{}
-
-%\newunicodechar{→}{\ensuremath{\rightarrow}}
-\lstset{
-	literate={→}{\ensuremath{\rightarrow\ }}1
-			{⟶}{\ensuremath{\rightarrow\ }}1
-			{Π}{\ensuremath{\Pi}}1
-			{⟹}{\ensuremath{\Longrightarrow\ }}1
-			{λ}{\ensuremath{\lambda}}1
-			{⟨}{\ensuremath{\langle}}1
-			{⟩}{\ensuremath{\rangle}}1
-			{❚}{\GS}1
-			{❙}{\RS}1
-			{❘}{\US}1
-			{Σ}{\ensuremath{\sum}}1
-			{…}{\ldots}1
-			{×}{\ensuremath{\times}}1
-			{π}{\ensuremath{\pi}}1
-			{⨁}{\ensuremath{\bigoplus}}1
-			{↪}{\ensuremath{\hookrightarrow}}1
-			{↩}{\ensuremath{\hookleftarrow}}1
-			{⊕}{\ensuremath{\oplus}}1
-			{∅}{\ensuremath{\emptyset}}1
-			{★}{\ensuremath{\star}}1
-			{ℕ}{\ensuremath{\mathbb N}}1
-			{∘}{\ensuremath{\circ}}1
-			{∀}{\ensuremath{\forall}}1
-			{≐}{\ensuremath{\doteq}}1
-			{⊦}{\ensuremath{\vdash}}1
-			{⁻}{\ensuremath{^-}}1
-			{¹}{\ensuremath{^1}}1
-}
+\lstnewenvironment{mmtcode}[1][language=mmt]{\lstset{language=mmt,keywordstyle=\bfseries,basicstyle=\ttfamily,mathescape,columns=fixed,#1}}{}

### 2018-10-19 mmt.tex
colons
finished for submission

@@
 \paragraph{Flat Modules}
-\mmt provides formal syntax for both structured modules and module expressions and defines their semantics via \textbf{flattening}, which defines for every module $M$ a flat module $\flt{M}$.
-
-\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expression, and the latter is optional.
-We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$ (see below for the definition of well-formed expressions).
+\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
+We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$.
+Every theory induces the set $\Exp{T}$ for the set of closed \textbf{expressions} using only the symbols $c\in\dom{T}$.
 Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
@@
 
-Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T e:e'$ (typing) and $\der_T e\equiv e'$ (equality) for every theory $T$ and $e,e'\in\Exp{T}$.
-%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
-The details can be found in \cite{rabe:howto:14}.
-
 \textbf{Flat morphisms} from a theory $S$ to a theory $T$ are lists of assignments $c:=e$ where $c\in\dom{S}$ and $e\in\Exp{T}$.
@@
 
-An \mmt \textbf{diagram} consists of a set of named structured module declarations.
+A \textbf{diagram} consists of a set of theory and morphism declarations.
 \begin{nomodexp}
-For a given diagram, we write $\Thy$ for the set of theory names.
-And we write $\Mo{S}{T}$ for the set of morphisms defined by
+For a given diagram, we write $\Thy$ for the set of declared \textbf{theories}.
+And we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
 \begin{compactitem}
@@
 \end{modexp}
-Then flattening assigns
-\begin{compactitem}
- \item to each $T\in\Thy$ the flat theory $\flt{T}$,
- \item to each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
-\end{compactitem}
+
+The theories and morphisms in a diagram may be structured (e.g., by using \emph{include} declarations), and \mmt defines their semantics via \textbf{flattening}, which defines for each $T\in\Thy$ the flat theory $\flt{T}$, and for each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
 
@@
 
-\mmt itself is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
-Our results in this paper are similarly logic-independent.
-The syntax and semantics of that base logic is provided by a set of rules for the judgments $\vdash_T e:E$ and $\vdash_T e=e'$ for $e,e',E\in\Exp{T}$.
-The declarations in theories and morphisms are subject to typing conditions using these judgments.
-The main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, e.g,., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
+\mmt is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
+Individual logics arise as fragments of \mmt: they single out the well-formed expressions by defining the \textbf{judgment} $\der_T e=e':E$ for typing and equality.
+(We treat the plain typing judgment $\der_T e:E$ as the special case $\der_T e=e:E$.)
+%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
+The declarations in theories and assignments in morphisms are subject to typing conditions, and the main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, i.e., if $\vdash_S e=e':E$, then $\vdash_T M(e)=M(e'):M(E)$.
 This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.
 
-The only aspect of the inference system for these judgments that is relevant to our purposes here is the typing rule for constants:
+We do not give all rules for these judgments.
+The only rule that is relevant to our purposes here is the one for constants:
 \[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c[=e]:E}\]
-Here, in order to avoid case distinctions for the case where a definition is present or not, we use a combined typing+equality judgment: $e_1=e_2:E$ represents the conjunction of $e_1:E$, $e_2:E$, and $e_1=e_2$.
+(Here, we merge the two cases where $c$ has a/no definiens $e$ into one rule for convenience.)
 Correspondingly, the definition of the homomorphic extension of a morphism with domain $S$ includes the following case for constants:
@@
 Here if $c$ has a definiens in $S$, we expand it before applying $M$.%
-\footnote{The \mmt tool accepts $c:e\in \flt{M}$ even if $(c:E=e')\in\flt{S}$ has a definition.
-In that case, \mmt checks $\vdash_T M(e')=e$ and puts $M(c)=e$.
+\footnote{The \mmt tool accepts $(c:=e)\in \flt{M}$ even if $(c:E=e')\in\flt{S}$.
+In that case, \mmt checks $\vdash_T M(e')=e:M(E)$ and puts $M(c)=e$.
 This is important for efficiency but not essential for our purposes here.}
 
-Note that these base cases introduce a mutual recursion between well-formedness and flattening:
+These base cases introduce a mutual recursion between well-formedness and flattening:
 the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
@@
 Vice versa, well-formedness is a precondition for defining the flattening --- the definition of flattening may become nonsensical if applied to ill-formed modules.
-
-It is desirable to define well-formedness independently of flattening.
-But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems first parse, check, and flatten a declaration entirely before moving on to the next declaration.
-Moreover, often the checking of well-formedness is as difficult or expensive as flattening anyway.
-In fact, inspecting practical systems with modular features (which often do not have a full rigorous formal specification) shows that our definition elegantly captures the essential commonalities between them.
+It might be desirable to define well-formedness independently of flattening.
+But our definition captures the typical behavior of practical systems, which first parse, check, and flatten one declaration entirely before moving on to the next one.
 
@@
 \begin{definition}[Theory]\label{def:theory}
-The grammar for theories and expressions is of the form
+The grammar for theories and expressions is
 \begin{grammar}
@@
 Dec      & n:E\opt{=E}           & \text{constant declaration}\\
-c        & t?n                   & \text{qualified constant identifiers} \\
+c        & T?n                   & \text{qualified constant identifiers} \\
 E        & c \alt \ldots         & \text{expressions built from constants} \\
@@
 In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
-We omit the remaining productions for expressions here, which allow forming complex expressions using application, binding, variables, literals, etc.
+We omit the productions and typing rules for the remaining expressions, which include application, binding, variables, literals, etc.
 \end{definition}
@@
 \begin{example}\label{ex:thgroup}
-For the purposes of our running examples, we assume a fixed base logic that provides a first-order logic a simple type system.
+For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
 \lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
@@
   U       : type
-  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
+  op      : U $\to$ U $\to$ U	 # 1 $\circ$ 2 
   unit    : U
-  inverse 	: U ⟶ U   # 1 ⁻¹ 
+  inverse : U $\to$ U   # 1 $^{-1}$ 
   // axioms omitted
@@
 \begin{mmtcode}
-DivisionGroup =
+DivGroup =
   U   : type
-  div 	: U ⟶ U ⟶ U  # 1 / 2 
-  unit 	: U
+  div : U $\to$ U $\to$ U  # 1 / 2 
+  unit: U
   // axioms omitted
@@
 
-At this point, this grammar only allows forming flat theories, i.e., $\Thy=\thy$.
-It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
-But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
-This makes it easy to incrementally add new declaration kinds and operators to the language.
-
+%
 %These theories are similar to OpenMath content dictionaries \cite{openmath}.
@@
 %Thus, symbols of the same name declared in different atomic theories are always distinguished.
-
+%
 We proceed accordingly for flat morphisms:
@@
 \begin{definition}[Morphism]\label{def:morphism}
-A \textbf{morphism} is of the form
+The grammar for \textbf{morphisms} is
 \begin{grammar}
 MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
-Ass    & c:=o                            & \text{assignment to symbol}\\
+Ass    & c:=E                            & \text{assignment to symbol}\\
 M      & m \alt \id{T} \alt M;M          & \text{morphism expressions}
 \end{grammar}
-such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
+A morphism must contain exactly one assignment $c:=e$ for each $(c:E)\in\flt{S}$, all of which must satisfy $\der_T e: M(E)$.
 \end{definition}
@@
 \begin{mmtcode}
-DG2G : DivisionGroup -> Group =
+DG2G : DivGroup -> Group =
   U     := U
-  div := [a,b] a ∘ (b⁻¹) 
+  div 	:= [a,b] a $\circ$ (b$^{-1}$) 
   unit 	:= unit
@@
 And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
-Additionally, the morphism must maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
+Additionally, the morphism maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
 \end{example}
@@
 \end{grammar}
-A diagram is well-formed if each atomic theory/morphism declaration has a unique name and is well-formed relative to the diagram preceding it.
+Each theory/morphism declaration must have a unique name and be well-formed relative to the diagram preceding it.
 \end{definition}
 
+At this point, the grammar only allows forming flat theories and morphisms.
+\begin{modexp}
+It is intentionally independent of the concrete choice of modularity principles, i.e., the kinds of module-structuring declarations (e.g., includes) and the module-expression-forming operators (e.g., union).
+But it already introduces all non-terminals for modularity: new declaration kinds for structured theories arise by adding productions for the non-terminal $Dec$ and new theory expressions are formed by adding productions for the non-terminal $T$.
+\end{modexp}
+Structuring features can be added to the language incrementally and individually.
+\begin{modexp}
 We give some examples how to add modularity principles to the grammar:
+\end{modexp}
 
@@
 \begin{grammar}
-  Dec & \icl{S} & \text{include a theory into a theory} \\
+  Dec & \icl{T} & \text{include a theory into a theory} \\
   Ass & \icl{M} & \text{include a morphism into a morphism}
@@
 Monoid =
-  U: type
-  op : U ⟶ U ⟶ U	 # 1 ∘ 2 
+  U 	: type
+  op 	: U $\to$ U $\to$ U	 # 1 $\circ$ 2 
   unit	: U
@@
   include Monoid
-  inverse : U ⟶ U   # 1 ⁻¹ 
+  inverse : U $\to$ U   # 1$^{-1}$ 
 \end{mmtcode}
@@
 
-For each example, we will define the semantics in the next section.
-
 \subsection{Semantics}
@@
 \paragraph{Flattening}
-Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
+The definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added independently of each other.
+\begin{modexp}
 In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
 This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $\flt{Y}$. 
-
 For theories $T$, at this point, this induction only has the base case of a reference to a declared theory $t$.
 Further cases arise when we add structuring declarations.
+\end{modexp}
 
+\begin{definition}[Flattening]
 For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
-\[\flt{t}=\flt{\Sigma}\]
-where
- \[\flt{\cdot}=\es\]
- \[\flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
-Here $\fltd{T}{D}$ is the flattening of declaration $d$ relative to $\flt{\Sigma}$.
-
+\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
+ \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
+Here $\fltd{T}{D}$ is the flattening of declaration $D$ relative to $\flt{\Sigma}$.
 At this point, we only have one case for declarations $D$, namely constant declarations.
@@
 
-Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
- \[\flt{\cdot}=\es\]
- \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
+Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, $\flt{m}$ is defined by induction on the assignments in $\sigma$:
+\[\flt{m}=\flt{\sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
 with the trivial base case
@@
 \begin{nomodexp}
-Moreover, for $T\in\Thy$ we define
-\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\flt{T}\}\]
-And for $M;N\in\Mo{R}{T}$, we define
-\[\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\flt{R}\}\]
-Note that in both cases, we only have to consider $R$-constants without definiens.
+Moreover, for $T\in\Thy$ we define $\flt{\id{T}}=\{c:=c \;|\; (c:E)\in\flt{T}\}$
+And for $M;N\in\Mo{R}{T}$, we define $\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; (c:E)\in\flt{R}\}$.
+Note that in both cases, we only have to consider constants without definiens.
 \end{nomodexp}
+\end{definition}
 
@@
 Therefore, includes can never lead to name clashes.
+Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
 
-Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
 The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
-Therefore, flattening and well-formedness must be distinguished:
+Therefore, flattening is not always defined:
   \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
-under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
+and $\sigma,\icl{M}$ is well-formed $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
+\end{example}
 
-The flat variant $\cn{Group}^\flat$ of our refactored theory $\cn{Group}$ from Ex.~\ref{syn:incl} is hence given by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in exactly the theory as presented earlier in \autoref{ex:thgroup} except for global identifiers from $\cn{Monoid}$ -- i.e.
-the prefixes $t$ in $t?n$.
+\begin{example}[Continuing Ex.~\ref{syn:incl}]
+$\cn{Group}^\flat$ is obtained by copying all included symbols (from $\cn{Monoid}$) over to $\cn{Group}$, resulting in the theory as given in Ex.~\ref{ex:thgroup} except that the identifiers are now, e.g., $\cn{Monoid}?U$ instead of $\cn{Group}?U$.
 \end{example}

### 2019-02-08 appl.tex
minor comments by reviewers

@@
  \item all assignments in its body are of the form $c:=c'$ for $T$-constants $c'$ without definiens
- \item every such $T$-constants $c'$ occurs in exactly one assignment.
+ \item every  $T$-constant $c'$ without definiens occurs in exactly one assignment.
 \end{compactitem}
@@
   \item A morphism declaration $r:s\to t=\{\ldots\}$ may carry the attribute \keyword{renaming}.
-  \item This is allowed if there are no implicit morphism between $s$ and $t$ yet.
+  \item This is allowed if there are no implicit morphisms between $s$ and $t$ yet.
   \item In that case, we define $r\in\IMo{s}{t}$ and $r^{-1}\in\Mo{t}{s}$.
@@
 \paragraph{Definitional Extensions}
-We say that the named theory $T$ is a \textbf{definitional extension} of $S$ if $T=S$ or the body of $T$ contains
+We say that the named theory $T$ is a \textbf{definitional extension} of $S$ if $T=S$ or the body of $T$ contains only
 \begin{compactitem}
- \item only constant declarations with definiens,
- \item only include declarations of theories that are definitional extensions of $S$.
+ \item constant declarations with definiens, and
+ \item include declarations of theories that are definitional extensions of $S$.
 \end{compactitem}
 
-If $T$ is a definitional extension of $S$, it is easy to prove that $t$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
-In particular, the isomorphism $S\to T$ maps $S$-constants to themselves and expands the definiens of all other constants.
+If $T$ is a definitional extension of $S$, it is easy to prove that $T$ and $S$ are isomorphic: both isomorphisms map all constants without definiens to themselves.
+In particular, the isomorphism $T\to S$ maps $S$-constants to themselves and expands the definiens of all other constants.
 
@@
 \begin{compactitem}
-  \item An include declaration $\icl{S}$ of a named theory $S$ inside theory $T$ may carry the attribute \keyword{definitional}.
+  \item An include declaration $\icl{S}$ of a named theory $S$ inside a theory $T$ may carry the attribute \keyword{definitional}.
   \item In that case, we define $\id{S}\in\IMo{T}{S}$ (in addition to the implicit morphism $\id{T}\in\IMo{S}{T}$ that is induced by the inclusion).
@@
 \paragraph{Canonical Isomorphisms}
-If we have isomorphisms $m:s\to t$ and $n:t\to s$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
+If we have isomorphisms $m:S\to T$ and $n:T\to S$, we simply spell them out in morphism declarations and add the keyword \keyword{implicit} to both.
 This requires no language extensions.
@@
 
-We define the declaration $M=N:S\to T \keyword{by} \{\sigma\}$ to be well-formed iff
+We define the declaration $\keyword{equal} M=N:S\to T \keyword{by} \{\sigma\}$ to be well-formed iff
 \begin{compactitem} 
@@
 A major drawback of using modular theories is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $r$ to $s$.
-This change to should be straightforward as it does not change the semantics of $t$.
+For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
+This change should be straightforward as it does not change the semantics of $t$.
 
 However, this is not a local change.
-It also requires updating every qualified reference from $r?n$ to $s?n$.
+It also requires updating every qualified reference from $s?n$ to $r?n$.
 Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.

### 2019-02-08 implicit.tex
minor comments by reviewers

@@
 
-Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit-diagram}.
+Our key idea is to use a commutative subdiagram of the \mmt diagram, which we call the \emph{implicit diagram}.
 It contains all theories but only some of the morphisms --- the ones designated as \emph{implicit}.
@@
 
-Adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
+Adding an $\keyword{implicit}$ attribute to a declaration is well-formed only if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories.
 \end{definition}
@@
 \begin{example}[Includes (continued from Ex.~\ref{sem:incl})]\label{impl:incl}
-For include morphisms, we add the following definition: if theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.
+For include morphisms, we add the following definition: if a theory $T$ contains the declaration $\icl{S}$, then the induced morphism from $S$ to $T$ is implicit.
 
 Combined with composition morphisms, we see that all transitive includes between theories are implicit.
-That corresponds to the intuition that anything that is included is available by its original name.
+That corresponds to the intuition that anything that is included is available by its original (i.e. qualified) name.
 \end{example}
@@
 	For example, we can write $a/b$ in $\cn{Group}$, where $/$ refers to the identifier $\cn{DivGroup}?\cn{div}$.
+
+	(Note that notations are carried over by implicit morphisms as well)
 \end{example}
@@
 \begin{theorem}[Conservativity of Implicit Morphisms]
-For well-formed diagrams and $S,T\in\Thy$ and $M\in\Mo{S}{T}$:
+For well-formed diagrams and $S,T\in\Thy$ and $M\in\IMo{S}{T}$:
 \begin{compactenum}
@@
   Definition expansion eliminates $c$ in favor of $M(c)$, and thus $\ov{c}=\ov{M(c)}$.
-  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$ That follows from the judgment preservation of morphisms.
+  We only have to show that $\vdash_T \ov{M(c)}:\ov{M(E)}$. That follows from the judgment preservation of morphisms.
 \end{compactitem}

### 2019-02-08 intro.tex
minor comments by reviewers

@@
 Areas in which tools based on theories and theory morphisms have been developed include specification \cite{obj,hets}, rewriting \cite{maude}, theorem proving \cite{imps,isabelle_locales}, and knowledge representation \cite{RK:mmt:10}.
-Closely related concepts are used in both object-oriented (\emph{classes}) and (\emph{type classes}) functional programming languages.
+Closely related concepts are used in both object-oriented (\emph{classes}) and functional (\emph{type classes}) programming languages.
 
@@
 This would result in the diagram on the right.%
-\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to an $T$-expression $m(E)$.
+\footnote{Note that we use the syntactic direction for the arrows, e.g., an arrow $m:S\to T$ states that any $S$-expression $E$ (e.g., a sort, term, formula, or proof) can be translated to a $T$-expression $m(E)$.
 Models are translated in the opposite direction.}
@@
 The key idea behind implicit morphisms is very simple:
-We maintain an additional diagram $I$, which is commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
+We maintain an additional diagram $I$, which is a commutative subdiagram of $D$ and whose morphisms we call \emph{implicit}.
 The condition of commutativity guarantees that $I$ has at most one morphism $i$ from theory $S$ to theory $T$, in which case we write $\ipc{i}{S}{T}$.
@@
 At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
-We recommend using implicit morphisms in all theory morphism--based formalisms.
+We recommend using implicit morphisms in all theory--morphism--based formalisms.
 More concretely, we present a formal system for developing structured theories with implicit morphisms.

### 2019-02-08 mmt.tex
minor comments by reviewers

@@
 \paragraph{Flat Modules}
-\textbf{Flat theories} are lists of declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
-We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$.
+\textbf{Flat theories} are lists of constant declarations $c:E[=e]$ where $E$ and $e$ are expressions, and the latter is optional.
+We write $\dom{T}$ for the set of constant identifiers $c$ in $T$.
 Every theory induces the set $\Exp{T}$ for the set of closed \textbf{expressions} using only the symbols $c\in\dom{T}$.
@@
 For a given diagram, we write $\Thy$ for the set of declared \textbf{theories}.
-And we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
+For theories $S,T\in\Thy$, we write $\Mo{S}{T}$ for the set of \textbf{morphisms} defined by
 \begin{compactitem}
- \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
+ \item for every declaration $m:S\to T=\{\cdot\}$, we have $m\in\Mo{S}{T}$,
  \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
- \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
+ \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$ (where $M;N$ corresponds to the usual function composition).
 \end{compactitem}
@@
 For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
-\lstinline|type| is the universe of types, \lstinline|A ⟶ B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
-We also use \mmt notations, which are attached to constant declarations as \lstinline|# <not>|; these are omitted from the formal grammar above because we only use them in the examples.
+\texttt{type} is the universe of types, \lstinline|A|$\rightarrow$\lstinline|B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
+We also use \mmt notations, which are attached to constant declarations as \lstinline|# <notation>|; these are omitted from the formal grammar above because we only use them in the examples.
 
@@
 \[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
- \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D}\]
+ \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D},\]
+ where $\cdot$ denotes the empty sequence.
+
 Here $\fltd{T}{D}$ is the flattening of declaration $D$ relative to $\flt{\Sigma}$.
@@
 \begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
-Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
+The $\keyword{include}$ operator adds new declarations $D$ to a theory, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
 We do this as follows
@@
 
-Note that the definition of $\fltd{\Sigma}{n:E=e}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
+Note that the definition of $\fltd{\Sigma}{(n:E=e)}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
 Therefore, includes can never lead to name clashes.
-Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
+Also note that, because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
 
-The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
+The situation is slightly more complicated for morphisms: if a morphism $\sigma$ out of $t$ includes two different morphisms out of $s$, these have to agree.
 Therefore, flattening is not always defined:
   \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
-and $\sigma,\icl{M}$ is well-formed $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
+if $\sigma,\icl{M}$ is well-formed and $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.
 \end{example}

### 2019-02-08 paper.tex
further cleanup, conclusion still missing
cleaned up intro and conc
finished for submission
minor comments by reviewers

@@
 \usepackage{graphicx,url,wrapfig,amsmath,amssymb}
+
+\usepackage{comment}
+% flip these to add module expressions; if so, some manual changes to intro and grammars are also necessary
+\excludecomment{modexp}
+\includecomment{nomodexp}
+% include unions as an example of module expressions
+\excludecomment{union}
 
@@
 \usepackage{macros/basics}
-%\usepackage[llncs]{theorems} %only for now so that the environments are defined; replace with nice package for this later
 \usepackage{local}
@@
 The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
+Concrete applications include elegantly identifying isomorphic theories and extending theories with definitions and theorems as well as efficiently building and maintaining large fine-granular import hierarchies of theories.
 \end{abstract}
@@
 
-\section{Theories and Theory Morphism}\label{sec:mmt}
+\section{Theories and Theory Morphisms}\label{sec:mmt}
 \input{mmt}
@@
 
-\section{Implicit Morphisms}
+\section{Implicit Morphisms}\label{sec:impl}
 \input{implicit}
 
-\section{Applications}
+\section{Applications}\label{sec:appl}
 \input{appl}
 
-\section{Extended Example}
-\input{example}
+\begin{modexp}
+\section{Advanced Module Expressions}\label{sec:complex}
+\input{complex}
+\end{modexp}
 
-%\section{Complex Theories and Morphisms}\label{sec:complex}
-%\input{complex}
-
-\section{Related Work}
-\input{related}
-
-\section{Conclusion}
+\section{Conclusion and Related Work}\label{sec:conc}
 \input{conc}

### 2019-02-11 response.txt

Updated response

Below we describe how we worked in all major comments.
We have worked in all minor comments as well.

---------------------------------------------
Comment 1 (Review 1):
  presentation is MMT-specific, unclear what to do with it
  no global vision of what to do with those ingredients and how the proposed concept can help

Comment 2 (Review 1):
  just a bigger example with a dozen theories and information about the advantages provided by implicit morphisms in this setting would be helpful to evaluate its importance

Comment 3 (Review 2) (see also Comment 5):
  the formal theory supporting implicit morphisms is very simple
  only major result of the paper, theorem 1, follows almost immediately, and the paper's contribution reduces to definitions

Response:

It helps to know the following.
Implicit morphism were first conceived by Rabe in 2010 and implemented as part of the Twelf system (which implements the dependent type theory LF).
The theory behind this implementation was never written up and therefore not published.

But the implementation already scaled well, and implicit morphisms were used in the LATIN logic atlas build by Rabe et al. in ~2009-2012.
The LATIN atlas already has around a 1000 theories and atomic morphisms, and about 50 of the latter are marked as implicit.
It also has a few hundred inclusions, each of which induces another implicit morphism.

In the following years, MMT was developed. As part of this development
- implicit morphisms were generalized from LF to the logic-independent level of MMT,
- their theory was worked out and written up,
- they were reimplemented from scratch as a part of MMT.
MMT is backwards-compatible with Twelf, and the LATIN atlas including its implicit morphisms can be used from within MMT.

The present paper was written as the climax of this development to coincide with the first MMT release in which implicit morphisms are officially supported.

The simplicity (Comment 3) is the result of improving syntax, semantics, and implementation for 8 years until we found the simplest possible formulation.
We think of our result as a language design result and of that simplicity as our main achievemnt.
But note that our response to Comment 5 will deepen the results of our paper.

The lack of big picture description (Comment 1) and larger case study (Comment 2) are the result of us focusing on the technical details and keeping the preliminaries and background minimal.
We believe now that was a mistake.

We believe we have remedied all three comments by adding the above text in a new paragraph in the introduction.
We have also increased the size of the example in Sect. 4.2 and 4.3.

---------------------------------------------
Comment 4 (Review 2):
  the hand-waving style led to mistakes or slight inconsistencies (see my comments below)

Response:

We have taken care of inconsistencies and work in all minor comments.
We expect that has naturally eliminated the hand-waving.

---------------------------------------------
Comment 5 (Review 2):
I would have liked to see more discussion on how their theory applies in other settings than MMT, in particular for a diagram of heterogeneous theories (e.g. over a Grothendieck institution).
My guess is that the transfer of results is not that immediate, for example you would need some mechanism for proofs.

Response:

MMT treats logics as theories and logic translations as theory morphisms (which in particular must translate any proof rules delared in the domain logic to proofs in the target logic).
Thus, morphisms between theories and logics are treated uniformly by MMT, and in particular logic morphisms can be implicit in the same way.
Indeed, several logic morphisms (e.g., PL -> FOL -> HOL etc.) are natural candidates for marking as implicit morphisms.

The description of logics and logic morphisms was originally omitted from the paper to make the theory simpler.
In light of Comment 3 and 5, we have added this description to Sect. 2 (which required a few small changes, which can be found by searching for "meta").
We have added a bigger example in Sect. 4.2 and describe in the new Sect. 4.3 how to build it as a diagram of heterogeneous theories.

---------------------------------------------
Noteworthy minor comments from Review 2:

--------
page 6, def. 4:
This seems faulty. Consider the example of Group, without inclusions.
Via flattening, you get
Group?U : type
Group?op : U -> U -> U
because with your definition (n : E)^\flat = t?n : E
I would expect to get
Group?op : Group?U -> Group?U -> Group?U
This must also be taken care of for imports.

Response:
The presentation in the paper is slightly confusing because "U" is concrete syntax, which is parsed in "Group?U" in abstract syntax.
We have added Rem. 1 to clarify that.

--------
page 10, renamings:
when using the attribute renaming, are the conditions of the
definition checked?

Response:
Yes.

--------
page 12:
instead of introducing "implicit-later", would it not be better to
have a mechanism for deriving that the morphism is implicit
(e.g. because the isomorphism proof exists)?

Response:
In general, not every isomorphism can automatically be implicit because there may be multiple different isomorphisms between the same theories.
However, in this particular case, the system could indeed infer that G2DG is the inverse of an implicit isomorphism and therefore should be implicit as well.
We rephrased the example accordingly.

---------------------------------------------
Comment 6 (Review 3):
  not convinced of the practical benefits of this particular implementation, and tend to consider that most of them can also be achieved (as per Section 4 of the paper), with minimal or no additional effort, using ordinary theory morphisms
  would gladly accept a counterargument

Response:
We agree that the effect can be achieved with minimal additional effort.
But we hold that implicit morphisms, which are ordinary theory morphisms with an additional boolean flag, are already the simplest possible way to do it, and we've spent years finding whittling down the ideas to the simplest possible solution.
(See also our response to Comment 1-3.)

This comment had 3 sub-comments, to which we respond below.
We believe that these responses constitute a strong counterargument.

------------
Comment 6.1: In regard to 'Identifying theories via implicit isomorphisms'

Response:
The reviewer is right with one exception: The main advantage of implicit morphisms m:A -> B is that m is later on inserted automatically without the user having to refer to m every time.
That is only possible if the user manually clarifies which morphisms should be implicit because there could be more than one isomorphism A -> B.

See also the response to Comment 6.2.

------------
Comment 6.2: In regard to 'Fine-granular and flexible theory hierarchies'

Response:
The well-formedness property of implicit morphisms is the same as for ordinary morphisms.
In particular, nothing special happens when the user makes the morphism CancellationMonoid -> Group implicit (except for checking that there is no implicit morphism of that type yet).

All the advantage come later when Group is used:

* A later theory that includes Group can directly refer to all identifiers from CancellationMonoid.
  MMT even lists these identifiers automatically in the auto-completion dialog, and the MMT parser automatically uses their notations.
  Thus, users can simply include Group as if it includes CancellationMonoid without having to know the details of the library strucutre.

* Any parameter of type CancellationMonoid can be substituted with an instance of Group.
  Similarly, any morphism into CancellationMonoid can be composed with a morphism out of Group. 
  In both cases, MMT automatically inserts the implicit moprhism.

Overall, this yields a big advantage when releasing a large library because users can work more intuitively without having to learn the library structure in detail (e.g., a user can use Group like a CancellationMonoid without worrying about the conversion).

We have added a paragraph at the end of Sect. 3.1 to emphasize these advantages.
Moreover, we have revised 4.2 to strengthen the motivation and use a larger example.

Regarding the question about inclusions:
Because inclusions are implicit by default, the commutativity requirement already ensures that an implicit morphism A -> B must be the identity on all theories that are included into both A and B.

------------
Comment 6.3: In regard to 'Transparent refactoring'

Response:
The reviewer is right.
A smarter way to do the refactoring example is to leave s as it is and introduce a second implicit morphism s' -> s.
We have changed the example accordingly.

The reviewer is also right that external refactoring mechanisms should also be used.
But when managing releases of big libraries, it is often desirable to ensure backwards compatibility without requiring users to run a refactoring propragation tool.
A typical application would be to deprecate the original theories but use implicit morphisms to ensure that they still work.
Then after a transition period the original theories are removed from the library and users are expected to switch to the new, cleaner and more orderly diagram.

We have added some sentences on that.

---------------------------------------------
Comment 7 (Review 3):
  discussion on the potential uses of implicit morphisms outside the scope of MMT
  connection with other generalisations of the set-theoretic notion of inclusion, e.g., factorisation or inclusion systems).

Response:
We have no results in this direction yet.
We thought of factorization and inclusion systems but did not mention them because we had no results about how they relate to implicit morphisms.
We'd be happy to receive ideas on this issue.
We added a sentence on that under future work.

### 2019-02-11 appl.tex

@@
 But one of the major short-comings of formal theories over traditional informal developments is that formal systems usually need to spell out these isomorphisms everywhere.
-In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification.
+In this section, we show that implicit morphisms elegantly allow exactly that kind of intuitive identification: we mark the canonical isomorphisms as implicit.
 
@@
 In general, note that because identity morphisms are implicit, our uniqueness requirement for implicit morphisms implies that two theories $S$ and $T$ must be isomorphic if there are implicit morphisms in both directions.
-Moreover, making a pair of isomorphisms implicit is well-formed if there are no other implicit morphisms between $S$ and $T$ yet.
+Moreover, making a pair of isomorphisms implicit is only allowed if there are no other implicit morphisms between $S$ and $T$ yet.
 
@@
 \begin{mmtcode}
-Monoid2 =
+Monoid2: $\Log$ =
   M          : type
@@
 	\begin{mmtcode}
-renaming MonoidRen : Monoid2 -> Monoid =
+renaming MonoidRen : Monoid2 -> Monoid =$^{\id{\Log}}$
   M          := U
@@
 \begin{mmtcode}
-InverseInvolution =
+InverseInvolution: $\Log$ =
   definitional include Group
@@
 \begin{mmtcode}
-implicit G2DG : Group -> DivGroup =
+implicit G2DG : Group -> DivGroup =$^{\id{\Log}}$
   U       := U
@@
 
-While making one of these isomorphisms implicit is straightforward, doing both requires checking that $m$ and $n$ are actually isomorphisms.
+While making one of these isomorphisms implicit is straightforward, doing it for both requires checking that $m$ and $n$ are actually isomorphisms.
 Otherwise, the uniqueness condition would be violated.
@@
 \[\keyword{implicit}:\cn{DG2G}:\cn{DivGroup}\to\cn{Group}=\text{(as above)}\]
- \[\keyword{implicit-later}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
+\[\keyword{implicit}:\cn{G2DG}:\cn{Group}\to\cn{DivGroup}=\text{(as above)}\]
 \[\keyword{equal}\cn{G2DG};\cn{DG2G}=\id{\cn{Group}}:\cn{Group}\to\cn{Group}=\text{(omitted)}\]
@@
 
-Note that we cannot mark \cn{G2DG} as \keyword{implicit} because that is only well-formed after proving the equalities.
-So we use the new attribute \keyword{implicit-later}, which states that a morphism should be considered as implicit as soon as subsequent equality proofs make it well-formed to do so.
+There is a subtle difficulty in marking \cn{G2DG} as \keyword{implicit}: the implicit-diagram only commutes after proving the equalities, which are only declared later.
+One option is to delay the commutativity check until the equalities have been checked.
+In that case, care must be taken to avoid using the implicitness of \cn{G2DG} while proving the equalities.
+But we obtain a more elegant solution from the observation that it is always sound and harmless to automatically make the inverse of an implicit isomorphism implicit as well.
+Thus, we can omit the attribute \keyword{implicit} on \cn{G2DG} altogether and use an implementation that infers that \cn{G2DG} is the inverse of an implicit isomorphism.
 \end{example}
 
-\subsection{Fine-Granular and Flexible Theory Hierarchies}
+\subsection{Fine-Granular and Flexible Theory Hierarchies}\label{sec:granular}
 \input{example}
+
+\subsection{Logic Morphisms}
+
+So far, we have assumed all theories use the same fixed meta-theory, e.g., $\Log$ in the running examples.
+But in practice, we usually develop theories heterogeneously by using the weakest possible meta-theory for each module.
+A strength of \mmt is that meta-theories are normal theories so that the same structuring formalism can be used for them, e.g., morphisms, includes, pushouts, and implicit morphisms are directly available for meta-theories.
+
+For example, consider the example from Figure~\ref{fig:magmas}.
+We could use first-order logic \cn{FOL} as the common meta-theory of all theories.
+But actually the much weaker logic \cn{Horn} (which uses restricted logical connectives that only allow creating Horn formulas) is sufficient for most of them.
+
+We do not want to declare a direct include $\cn{Horn}\to\cn{FOL}$ for the same reasons as discussed in  Section~\ref{sec:granular}.
+Instead, we want to give a morphism $\cn{EmbedHorn}:\cn{Horn}\to\cn{FOL}$, which maps all Horn formula constructors to their \cn{FOL} counterparts.
+We can now make \cn{EmbedHorn} implicit, thus capturing the fact that Horn logic is a fragment of \cn{FOL}.
+
+Moreover, assume we have built the diagram in Figure~\ref{fig:magmas} using the meta-theory \cn{Horn} where possible and \cn{FOL} where necessary.
+An example for the latter is \cn{TotalOrder}, which also uses $\vee$.
+Without implicit morphisms, we would have to write
+\[\cn{TotalOrder}:\cn{FOL}=\{\icl{\cn{EmbedHorn}(\cn{PartialOrder})},\ldots\}\]
+Here the \cn{Horn}-theory \cn{PartialOrder} must be explicitly translated to \cn{FOL} before including it, which is awkward for users both when writing and reading.
+But using implicit morphisms and the semantics of Ex.~\ref{impl:incl2}, we can simply write $\icl{\cn{PartialOrder}}$ --- the logic translation remains transparent to the user.
+
+In practice, many logic morphisms are naturally implicit, either because they are includes to begin with or because they represent a canonical logic embedding.
 
@@
 
-A major drawback of using modular theories is that it can preclude transparent refactoring.
-For example, consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
+A major drawback of using modular theories is that it can preclude transparent refactoring (to insert intermediate theories as mentioned in Section~\ref{sec:granular}).
+As an example, we consider a theory $t=\{\icl{r},\icl{s}\}$, and assume we want to move a constant declaration $D$ for the name $n$ from $s$ to $r$.
 This change should be straightforward as it does not change the semantics of $t$.
@@
 Even if the source files always use the unqualified reference $n$ (because the checker is smart enough to dynamically disambiguate them), this still requires a global rebuild to reach a consistent state again.
-But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know or does not have access to.
+But such references can occur anywhere where $t$ is used, and that may include files that the person who does the refactoring does not know about or does not have access to.
+One option in this case is to use an extra-logical refactoring tool that propagates such changes.
+But when managing releases of big libraries, it is often desirable to deprecate the original theories but still ensure backwards compatibility.
+Then after a transition period the original theories are removed from the library and users are expected to propagate the refactoring.
 
@@
 \begin{compactenum}
-  \item We rename $s$ to $s'$.
-  \item We delete the declaration $D$ from $s'$.
-  \item We create new theories $r'=\{\icl{r},\,D\}$ and $s=\{\icl{s'},D\}$.
+  \item We keep $r$ and $s$ as they are.
+  \item We create a new theory $r'=\{\icl{r},\,D\}$.
+  \item We create a new theory $s'$ that is like $s$ except for deleting the declaration $D$.
   \item We change $t$ to $t=\{\icl{r'},\icl{s'}\}$.
-  \item We add an implicit morphism $s\to t$ that maps $s?n$ to $r'?n$.
+  \item We add implicit morphisms $s'\to s$ (mapping $s'?x$ to $s?x$ for all $x$) and $s\to t$ (mapping $s?n$ to $r'?n$ and $s?x$ to $s'?x$ for all $x\neq n$).
 \end{compactenum}
-The situation before (left) and after (right) refactoring is given below:
+The situation before (left) and after (right) refactoring is given below.
+Note that the right diagram commutes.
 \begin{center}
@@
   \draw[right hook-latex] (r) to (rp) {};
-  \draw[right hook-latex] (sp) to (s) {};
+  \draw[arrow] (sp) to (s) {};
 
@@
 Afterwards, $t$ has the desired new structure.
-But all old references to $s?n$ stay well-formed so that no global changes are needed.
+But all old references to $r$ and $s$ stay well-formed so that no global changes are needed.
+Now $r$ and $s$ can be deprecated and eventually removed in favor of $r'$ and $s'$.

### 2019-02-11 example.tex

@@
 For example, Ex.~\ref{syn:incl} defines \cn{Group} with an include from \cn{Monoid}.
-Instead, we could include \cn{Monoid} into the intermediate theory \cn{CancellationMonoid} and include that into \cn{Group}.
-This change is not easy in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents a very expensive refactoring problem.
-So instead we could systematically build a hierarchy that uses every intermediate theory (as done in \cite{mathscheme}).
-But this yields a very deep and complex hierarchy that is hard to navigate for casual users.
-Moreover, it does not protect us from later on discovering yet another intermediate theory that should have been added.
+Instead, we could have used an intermediate theory and includes $\cn{Monoid}\harr \cn{CancellationMonoid}\harr\cn{Group}$.
 
-Implicit morphisms provide a simple solution to this problem because they behave effectively like inclusions but can be added later on.
-In the above example, we would
+It is very common to have increasingly strong theories $R,S,T$, where a design with two includes $R\harr S\harr T$ is not desirable:
 \begin{compactitem}
- \item define \cn{CancellationMonoid} with an include from \cn{Monoid},
- \item keep \cn{Group} as it is, i.e., also with an include \cn{Monoid},
- \item add an implicit morphism $\cn{CancellationMonoid}\to\cn{Group}$.
+ \item Often $R\harr T$ has been defined first and $S$ only later.
+   This is very common because people usually formalize the most important theories (e.g., \cn{Monoid} and \cn{Group}) first.
+   But inserting $S$ is not easy in retrospect --- changing the theory hierarchy (which is one of the most fundamental structures of a library) usually presents a very expensive refactoring problem.
+   And even if we systematically use includes for every known intermediate theory like $S$ (as done in \cite{mathscheme}), we might later discover a new intermediate theory that should have been added.
+ \item Often the most natural axioms to use in $T$ are the same independent of whether $T$ includes $R$ or $S$ (e.g., users might prefer the usual inverse-element axioms in \cn{Group} even if they have included \cn{CancellationMonoid}).
+ In that case, the axioms of $S$ become provable in $T$ if we use $R\harr S\harr T$.
+ This either causes $T$ to have redundant axioms or requires a more complex include mechanism that allows $T$ to include $S$ in a way that turns some of $S$-axioms into theorems.
 \end{compactitem}
-\medskip
 
-As a more complex case study, we have used our implementation of implicit morphisms in \mmt to formalize a hierarchy of theories between \cn{Band} and \cn{Semilattice}:% 
-\footnote{Our formalization can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
+Therefore, it is common to use a commuting triangle consisting of two includes $R\harr T$ and $R\harr S$ and one morphism $m:S\to T$.
+But this is awkward because the relation ``every $T$ is an $S$'' is now mediated by $m$ rather than being canonical.  
+Implicit morphisms provide a simple solution to this problem: we keep the triangle but make the morphism $m$ implicit.
+This captures exactly the canonical conversion from $S$ to $T$.
+
+\begin{figure}[htb]
 	\begin{center}
-\begin{tikzpicture}[xscale=2]
-  \node (s1) at (0,0) {$z\circ x\circ y\circ z\doteq  z\circ x\circ z\circ y\circ z$};
+		\includegraphics[width=.8\textwidth]{graph.png}
+		\caption{Magma hierarchy with includes (gray) and implicit morphisms (black)}\label{fig:magmas}
+	\end{center}
+\end{figure}
 
-  \node (s2a) at (-1,-1) {$z\circ x\circ y\doteq  z\circ x\circ z\circ y$};
-  \node (s2b) at (1,-1) {$y\circ x\circ z\doteq  y\circ z\circ x\circ z$};
-  \draw[arrow] (s1) to (s2a) {};
-  \draw[arrow] (s1) to (s2b) {};
+Already the elementary algebraic hierarchy provides countless examples of such situations.
+For a small fragment of the hierarchy of magmas, Figure~\ref{fig:magmas} shows one possible design using numerous implicit morphisms.
+In particular, it uses some of the examples and features from this paper, e.g., an implicit isomorphism to identify the order-theoretic and the algebraic development of semilattices.
 
-  \node (s3a) at (-2,-2) {$x\circ y\doteq  x\circ y\circ x$};
-  \node (s3b) at (0,-2) {$z\circ x\circ y\circ z\doteq  z\circ y\circ x\circ z$};
-  \node (s3c) at (2,-2) {$x\circ y\doteq  y\circ x\circ y$};
-  \draw[arrow] (s2a) to (s3a) {};
-  \draw[arrow] (s2a) to (s3b) {};
-  \draw[arrow] (s2b) to (s3b) {};
-  \draw[arrow] (s2b) to (s3c) {};
+It also uses multiple implicit morphisms to introduce the various intermediate theories between $\cn{Band}\harr\cn{SemiLattice}$. 
+All of these are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, e.g., \cn{LeftRegularBand} uses $F=\forall x,z. z\circ x\circ z \doteq z\circ x$.
+The implicit morphisms map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
+It is straightforward to prove that this part of the diagram commutes: any two morphisms are identical except for the assignment to the axiom $a$, and these are equal due to proof irrelevance.%
+\footnote{Our formalization of bands can be found at \url{https://gl.mathhub.info/MMT/examples/blob/devel/source/bands.mmt}.}
+%\footnote{All varieties of bands can be axiomatized in this way.}
 
-  \node (s4a) at (-1,-3) {$z\circ x\circ y\doteq  z\circ y\circ x$};
-  \node (s4b) at (1,-3) {$x\circ y\circ z\doteq  y\circ x\circ z$};
-  \draw[arrow] (s3a) to (s4a) {};
-  \draw[arrow] (s3b) to (s4a) {};
-  \draw[arrow] (s3b) to (s4b) {};
-  \draw[arrow] (s3c) to (s4b) {};
-
-  \node (s5) at (0,-4) {$x\circ y\doteq  y\circ x$};
-  \draw[arrow] (s4a) to (s5) {};
-  \draw[arrow] (s4b) to (s5) {};
-
-  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
-  %\node at (t1.north west) {$S$};
-  %\node at (t1) {$D$};
-  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
-  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
-  %\node at (t2.north west) {$T$};
-  %\node at (t2) {$C$};
-  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
-\end{tikzpicture}
-\end{center}
-All of these theories are of the form $t=\{\icl{\cn{Band}},\,a: F\}$, where $F$ is the formula given in the diagram above.%
-%\footnote{All varieties of bands can be axiomatized in this way.}
-In particular, $\cn{SemiLattice}=\{\icl{\cn{Band}},\,a: \forall [x]\forall [y] x\circ y\doteq y\circ x\}$.
-
-The diagram above gives all morphisms between them that describe the lattice structure of the corresponding varieties of bands, all of which map the constants from \cn{Band} to themselves and the axiom $a$ to a proof.
-In our formalization, we make all of these morphisms implicit.
-It is straightforward to prove that the diagram commutes: any two morphisms are identical except for the assignment to the axiom $a$, and these are equal due to proof irrelevance.
+%\begin{center}
+%\begin{tikzpicture}[xscale=2]
+%  \node (s1) at (0,0) {$z\circ x\circ y\circ z\doteq  z\circ x\circ z\circ y\circ z$};
+%  
+%  \node (s2a) at (-1,-1) {$z\circ x\circ y\doteq  z\circ x\circ z\circ y$};
+%  \node (s2b) at (1,-1) {$y\circ x\circ z\doteq  y\circ z\circ x\circ z$};
+%  \draw[arrow] (s1) to (s2a) {};
+%  \draw[arrow] (s1) to (s2b) {};
+%  
+%  \node (s3a) at (-2,-2) {$x\circ y\doteq  x\circ y\circ x$};
+%  \node (s3b) at (0,-2) {$z\circ x\circ y\circ z\doteq  z\circ y\circ x\circ z$};
+%  \node (s3c) at (2,-2) {$x\circ y\doteq  y\circ x\circ y$};
+%  \draw[arrow] (s2a) to (s3a) {};
+%  \draw[arrow] (s2a) to (s3b) {};
+%  \draw[arrow] (s2b) to (s3b) {};
+%  \draw[arrow] (s2b) to (s3c) {};
+%  
+%  \node (s4a) at (-1,-3) {$z\circ x\circ y\doteq  z\circ y\circ x$};
+%  \node (s4b) at (1,-3) {$x\circ y\circ z\doteq  y\circ x\circ z$};
+%  \draw[arrow] (s3a) to (s4a) {};
+%  \draw[arrow] (s3b) to (s4a) {};
+%  \draw[arrow] (s3b) to (s4b) {};
+%  \draw[arrow] (s3c) to (s4b) {};
+%
+%  \node (s5) at (0,-4) {$x\circ y\doteq  y\circ x$};
+%  \draw[arrow] (s4a) to (s5) {};
+%  \draw[arrow] (s4b) to (s5) {};
+% 
+%  %\node[thy,inner sep=.3cm,fill=gray] (t1) at (0,0) {};
+%  %\node at (t1.north west) {$S$};
+%  %\node at (t1) {$D$};
+%  %\node[thy,inner sep=.5cm,fill=lightgray] (s2) at (2,0) {};
+%  %\node[thy,inner sep=.3cm,fill=gray] (t2) at (2,0) {};
+%  %\node at (t2.north west) {$T$};
+%  %\node at (t2) {$C$};
+%  %\draw[view] (t1) to[out=10,in=170] node[above] {$\sigma$} (t2);
+%\end{tikzpicture}
+%\end{center}
 

### 2019-02-11 implicit.tex

@@
 Any $c\in\dom{S}$ is treated as a valid $T$-identifiers with definiens $M(c)$.
-\item We can use canonical identifiers $S?n$ without worrying about ambiguity. Because there can be at most one implicit morphism $\ipc{i}{S}{T}$, using $S?n$ as an identifier in $T$ is unambiguous.
+\item We can use canonical identifiers $S?n$ without worrying about ambiguity.
+Because there can be at most one implicit morphism $\ipc{i}{S}{T}$, using $S?n$ as an identifier in $T$ is unambiguous.
 %Crucially, we can use $S?n$ without bothering what $i$ is.
+\end{compactitem}
+
+In practice, the \mmt system searches for an implicit morphism $\ipc{i}{S}{T}$ whenever needed to make an expression well-formed, e.g.,
+\begin{compactitem}
+ \item An $S$-constant $c$ is used in $T$: treat $c$ as an abbreviation for $i(c)$.
+ \item A model $M$ of $T$ is used even though a model of $S$ is expected: reduce $M$ via $i$.
+ \item A morphism $M:R\to S$ is composed with a morphism $M':T\to U$: treat $M;M'$ like $M;i;M'$.
 \end{compactitem}
@@
  \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
+ \item if $t$ has meta-theory $L$, the identity map as an implicit morphism $L\to t$,
 \begin{nomodexp}
@@
 
+With implicit morphisms, we can also relax the semantics of many structuring features:
+
+\begin{example}[Includes and Meta-Theories (continued from Ex.~\ref{sem:incl})]\label{impl:incl2}
+Consider $t:L=\{\ldots,\icl{S},\ldots\}$ where $S$ has meta-theory $L'$.
+Instead of requiring $L'\harr L$ as in Ex.~\ref{sem:incl}, we require only $\ipc{i}{L'}{L}$.
+
+In that case, we treat $\icl{S}$ as an abbreviation for $\icl{i(S)}$, where $i(S)$ is the pushout of $S$ along $i$ (see \cite{rabe:howto} for details on \mmt pushouts).
+If $L'\harr L$, this reduces to the original semantics.
+\end{example}
+
 %\subsection{Commutativity}\label{sec:commute}

### 2019-02-11 intro.tex

@@
 At the highest level, our contribution is the observation that implicit morphisms form a sweet spot of a very simple language feature that has substantial practical uses.
-We recommend using implicit morphisms in all theory--morphism--based formalisms.
+We recommend using implicit morphisms in all theory morphism--based formalisms.
 More concretely, we present a formal system for developing structured theories with implicit morphisms.
@@
 
-We describe several example applications of implicit morphisms in detail: the identification of isomorphic theories, definitional extensions of theories, building large hierarchies of theories with many rarely used intermediate theories, and transparently refactoring theory hierarchies.
+We describe several example applications of implicit morphisms in detail: the identification of isomorphic theories, definitional extensions of theories, building large hierarchies of theories with many rarely used intermediate theories, seamlessly moving theories across logic morphisms, and transparently refactoring theory hierarchies.
+
+\paragraph{Previous Work}
+Implicit morphism were first conceived by Rabe in 2010 and implemented as part of the Twelf system \cite{twelf} (which implements the dependent type theory LF).
+The theory behind this implementation was never written up and not published.
+But the implementation already scaled well, and implicit morphisms were used in the LATIN logic atlas \cite{CHKMR:latinabs} built by Rabe and others in 2009--2012.
+The LATIN atlas already has around a 1000 theories and atomic morphisms, and about 50 of the latter are marked as implicit.
+It also has a few hundred inclusions, each of which induces another implicit morphism.
+
+Since then, MMT has been developed, and
+\begin{compactitem}
+ \item implicit morphisms were generalized from LF to the logic-independent level of MMT,
+ \item their theory was worked out,
+ \item they were reimplemented from scratch as a part of MMT.
+\end{compactitem}
+MMT is backwards-compatible with Twelf, and the LATIN atlas including its implicit morphisms can be used from within MMT.
+The present paper introduces these results in their final, most elegant form.
 

### 2019-02-11 mmt.tex

@@
 \mmt is independent of the base logic and provides a theory and theory morphism layer on top of an arbitrary declarative language.
-Individual logics arise as fragments of \mmt: they single out the well-formed expressions by defining the \textbf{judgment} $\der_T e=e':E$ for typing and equality.
+Individual logics $L$ arise as fragments of \mmt: they single out the well-formed expressions by defining the \textbf{judgment} $\der_T e=e':E$ for typing and equality.
 (We treat the plain typing judgment $\der_T e:E$ as the special case $\der_T e=e:E$.)
+The logic $L$ is itself represented as an \mmt theory, which we call the \textbf{meta-theory} of $T$.
+$L$ provides, in particular, the primitive operators and typing rules that are used to form the well-formed expressions of $T$. occurring in $T$.
+$L$ may and often has a meta-theory as well, usually a logical framework used to define the logic. 
+Unless mentioned otherwise, all results in this paper apply for a fixed arbitrary meta-theory, which we will occasionally omit from the notation.
+
 %This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
-The declarations in theories and assignments in morphisms are subject to typing conditions, and the main theorem about \mmt is that well-typed morphisms $M:S\to T$ preserve all judgments, i.e., if $\vdash_S e=e':E$, then $\vdash_T M(e)=M(e'):M(E)$.
+The declarations in theories and assignments in morphisms are subject to typing conditions, and the main theorem about \mmt is that under reasonable assumptions on $L$, the well-typed morphisms $M:S\to T$ preserve all judgments, i.e., if $\vdash_S e=e':E$, then $\vdash_T M(e)=M(e'):M(E)$.
 This includes the preservation of truth via the propositions-as-types principle if $E$ is a proposition and $e$ its proof.
@@
 \begin{grammar}
-TDec     & T=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
+TDec     & T\opt{:T}=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
 Dec      & n:E\opt{=E}           & \text{constant declaration}\\
@@
 
-In a theory declaration, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants only.
+In a theory declaration $T:L=\{\vec{D}\}$, the \textbf{meta-theory} $L$ (if present) must be a previously declared theory, each symbol \textbf{name} $n$ may be declared only once, and its \textbf{type} and \textbf{definiens} (if present) must be closed expressions over the previously introduced constants (including those of $L$).
 We omit the productions and typing rules for the remaining expressions, which include application, binding, variables, literals, etc.
@@
 \begin{example}\label{ex:thgroup}
-For the purposes of our running examples, we assume a fixed base logic that provides a simple type system.
+For the purposes of our running examples, we assume a fixed meta-theory $\Log$ that provides a simple type system:
 \texttt{type} is the universe of types, \lstinline|A|$\rightarrow$\lstinline|B| is the type of functions, and lambda abstraction is written \lstinline|[x:A] t x|.
@@
 \begin{mmtcode}
-Group =
+Group: $\Log$ =
   U       : type
@@
 \begin{mmtcode}
-DivGroup =
+DivGroup: $\Log$ =
   U   : type
@@
 \begin{grammar}
-MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
+MDec   & m:T\to T=^{\opt{M}}\{Ass,\ldots,Ass\} & \text{flat morphism declaration}\\
 Ass    & c:=E                            & \text{assignment to symbol}\\
@@
 \end{grammar}
-A morphism must contain exactly one assignment $c:=e$ for each $(c:E)\in\flt{S}$, all of which must satisfy $\der_T e: M(E)$.
+A morphism $m:S\to T=^M\{\vec{A}\}$ must contain exactly one assignment $c:=e$ for each $(c:E)\in\flt{S}$, all of which must satisfy $\der_T e: m(E)$.
+Moreover, if $S$ has a meta-theory $L$, then $M$ must be present and must be a morphism from $L$ to $T$.
 \end{definition}
@@
 \begin{mmtcode}
-DG2G : DivGroup -> Group =
+DG2G : DivGroup -> Group =$^{\id{\Log}}$
   U     := U
@@
 \end{mmtcode}
-It maps the universe and unit of a division group to the corresponding notions of a group.
+Here $\id{\Log}$ maps the meta-theory to itself.
+Then universe and unit of a division group are mapped to the corresponding notions of a group.
 And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
@@
 \begin{mmtcode}
-Monoid =
+Monoid: $\Log$ =
   U 	: type
@@
   unit	: U
-Group =
+Group: $\Log$ =
   include Monoid
@@
 \begin{definition}[Flattening]
-For the base case, of a theory $t=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
-\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb
+For the base case, of a theory $t:L=\{\Sigma\}$, we define $\flt{t}$ by induction on the declarations in $\Sigma$:
+\[\flt{t}=\flt{\Sigma} \tb\mwhere\tb \flt{\cdot}=\flt{L} \tb\mand\tb
  \flt{(\Sigma,D)}=\flt{\Sigma}\cup\fltd{\flt{\Sigma}}{D},\]
  where $\cdot$ denotes the empty sequence.
+ If $L$ is not present, we can assume it to be the empty theory.
 
@@
 
-Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, $\flt{m}$ is defined by induction on the assignments in $\sigma$:
-\[\flt{m}=\flt{\sigma} \tb\mwhere\tb \flt{\cdot}=\es \tb\mand\tb\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
+Correspondingly, for a declared morphism $m:S\to T=^{\opt{M}}\{\sigma\}$, $\flt{m}$ is defined by induction on the assignments in $\sigma$:
+\[\flt{m}=\flt{\sigma} \tb\mwhere\tb \flt{\cdot}=\flt{M} \tb\mand\tb\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
 with the trivial base case
  \[\fltd{\sigma}{(c:=e)}=\{c:=e\}\]
+If $M$ is not present, we can assume it to be empty morphism.
+
 \begin{nomodexp}
@@
 
-Note that the definition of $\fltd{\Sigma}{(n:E=e)}$ qualifies the constant name $n$ with the theory name $t$ to form the identifier $t?n$ in the flattening.
-Therefore, includes can never lead to name clashes.
-Also note that, because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
+There is some flexibility as to when the including and the included theory have different meta-theories $L$ resp. $L'$.
+We define that such an include is only allowed if $L'$ is also included into $L$.
+That way, an include declaration never changes the meta-theory of the including theory.
 
+Note that, because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
+
+Because we use qualified identifiers $t?n$, includes can never lead to name clashes.
 The situation is slightly more complicated for morphisms: if a morphism $\sigma$ out of $t$ includes two different morphisms out of $s$, these have to agree.
@@
 \end{example}
+
+\begin{remark}[Qualified Identifiers]
+According to the grammar, any occurrence of a constant $c$ in an expression is of the form $t?n$, i.e., qualified by its theory name.
+In the definition all declarations in $\flt{t}$ are qualified accordingly.
+This precludes any name clashes when constants of the same local name are declared in multiple theories that become part of $\flt{t}$, e.g., via the meta-theory or includes.
+
+The form $t?n$ is \emph{abstract} system-facing syntax.
+In \emph{concrete} user-facing syntax (as used in our examples), it is usually sufficient to write $n$ and let the \mmt parser infer which constant is meant.
+\end{remark}
 

### 2019-02-11 paper.tex

@@
 The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
-Concrete applications include elegantly identifying isomorphic theories and extending theories with definitions and theorems as well as efficiently building and maintaining large fine-granular import hierarchies of theories.
+Concrete applications include elegantly identifying isomorphic theories and extending theories with definitions and theorems as well as efficiently building and maintaining large, fine-granular, and heterogeneous hierarchies of theories.
 \end{abstract}

### 2019-02-11 implicit.tex

@@
 
-In that case, we treat $\icl{S}$ as an abbreviation for $\icl{i(S)}$, where $i(S)$ is the pushout of $S$ along $i$ (see \cite{rabe:howto} for details on \mmt pushouts).
+In that case, we treat $\icl{S}$ as an abbreviation for $\icl{i(S)}$, where $i(S)$ is the pushout of $S$ along $i$ (see \cite{rabe:howto:14} for details on \mmt pushouts).
 If $L'\harr L$, this reduces to the original semantics.

### 2019-02-11 intro.tex

@@
 The theory behind this implementation was never written up and not published.
-But the implementation already scaled well, and implicit morphisms were used in the LATIN logic atlas \cite{CHKMR:latinabs} built by Rabe and others in 2009--2012.
+But the implementation already scaled well, and implicit morphisms were used in the LATIN logic atlas \cite{CHKMR:latinabs:11} built by Rabe and others in 2009--2012.
 The LATIN atlas already has around a 1000 theories and atomic morphisms, and about 50 of the latter are marked as implicit.

### 2019-02-11 paper.tex

@@
 \bibliographystyle{alpha}
-\bibliography{../../../../frabe/Program_Data/Latex/bib/rabe,../../../../frabe/Program_Data/Latex/bib/systems,../../../../frabe/Program_Data/Latex/bib/institutions,../../../../frabe/Program_Data/Latex/bib/pub_rabe,../../../../frabe/Program_Data/Latex/bib/historical}
+\bibliography{macros/pub_rabe,macros/rabe,macros/systems}
 

### 2019-02-21 intro.tex

@@
 We recommend using implicit morphisms in all theory morphism--based formalisms.
-More concretely, we present a formal system for developing structured theories with implicit morphisms.
-Our starting point is the \mmt language \cite{RK:mmt:10}, which already provides a very general setting for defining and working with theories and morphisms.
-\mmt is logic-independent, i.e., allows embedding a large variety of declarative languages (logics, type-theories, etc.).
-Therefore, all our results can be directly applied to any language $L$ represented in \mmt or easily transferred to dedicated implementations of $L$.
+More concretely, we present a formal system for developing structured theories with implicit morphisms based on the \mmt language \cite{RK:mmt:10}.
+We choose \mmt because it provides a simple framework for working with theories and morphisms while remaining logic-independent: \mmt allows embedding a large variety of declarative languages $L$ (logics, type-theories, etc.) so that our results immediately yield an implementation of implicit morphisms for any such $L$.
+We also expect but do not describe here that other morphism-based systems can be easily extended to allow for implicit morphisms.
 

### 2019-02-21 paper.tex

@@
 The practical appeal of implicit morphisms is that they hit the sweet-spot of being extremely simple to understand and implement while significantly helping with structuring large collections of theories.
+
 Concrete applications include elegantly identifying isomorphic theories and extending theories with definitions and theorems as well as efficiently building and maintaining large, fine-granular, and heterogeneous hierarchies of theories.
+Our results are formulated and implemented in the \mmt language and system, and we expect they can be transfered to other morphism-based formalisms relatively easily.
 \end{abstract}
@@
 \bibliographystyle{alpha}
-\bibliography{macros/pub_rabe,macros/rabe,macros/systems}
+\bibliography{macros/pub_rabe,macros/rabe,macros/systems,macros/institutions}
 
